### 2019-02-12 coursework/academic-writing/strategy-priority.tex
import from old repository

\documentclass{scrartcl}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{patterns}
\usepackage{subcaption}
\usepackage{pifont}
\usepackage{graphicx}
\newcommand{\vampire}{{Vampire}}
\newcommand{\iprover}{iProver}
\newcommand{\E}{{\textsc E}}
\newcommand{\zthree}{{\textsc Z3}}
\newcommand{\tick}{\ding{51}}
\newcommand{\cross}{\ding{55}}

\title{Dynamic Strategy Priority: \\ Empower the strong and abandon the weak\footnote{Originally an extended abstract by the same authors that will be presented at AITP, March 2018. Re-purposed for COMP80142.}.}
\author{
 Michael Rawson \and 
 Giles Reger
}

\begin{document}
\maketitle
\begin{abstract}
\noindent
A key usability factor in interactive theorem proving is the time taken for automation tools to complete.
Automated theorem provers run lists of proof strategies sequentially, which can cause proofs to be found more slowly than necessary.
We show that it is possible to predict which strategies are likely to succeed while they are running using an artificial neural network.
We also implement a run-time strategy scheduler in the \vampire{} prover which improves average proof search time, and hence increases usability for interactive theorem proving.
\end{abstract}

\section{Introduction}
Modern automated theorem provers (e.g. \E{}~\cite{E}, \iprover{}~\cite{iProver}, \vampire{}~\cite{Vampire}, CVC4 \cite{CVC4}) for first-order logic rely on \emph{portfolio modes}~\cite{portfolio}, which utilise hundreds of distinct \emph{strategies}.
Of these strategies, only a few might solve a hard problem, often rapidly.
Typically, a portfolio of strategies has a pre-defined execution order: the prover will process strategies in this order, running each until a winning strategy is found or the prover runs out of time.

Portfolio modes are important as, in practice, there is no best strategy.
Furthermore, it is uncommon that two hard problems are efficiently solved by the same strategy.
However, portfolio execution is not without problems: deciding the optimal ordering and time allocation is hard in general~\cite{predict-success}, and produces overly-rigid, brittle engineering when applied to specific domains, such as those found in the TPTP problem set~\cite{TPTP}.
Moreover, for any particular problem, some lengthy strategies that are successful on other problems are doomed to failure from the outset --- see Figure \ref{fig:blocking} --- but are left to run unchecked by the prover, wasting time that could be spent on more productive strategies.

\begin{figure}
\begin{tikzpicture}[x=0.01\textwidth, y=.3in]
	\draw[thick, ->] (0, 6) -- (100, 6) node {};

	\draw (-3, 4.5) node {(a)};
	\draw[fill=gray] (0, 4) rectangle (30, 5) node [midway] {1 \cross};
	\draw[fill=gray] (30, 4) rectangle (50, 5) node [midway] {2 \cross};
	\draw[fill=gray] (50, 4) rectangle (80, 5) node [midway] {3 \cross};
	\draw[fill=green] (80, 4) rectangle (90, 5) node [midway] {4 \tick};
	\draw (90, 4) rectangle (100, 5) node [midway] {\ldots};

	\draw (-3, 2.5) node {(b)};
	\draw[fill=gray] (0, 2) rectangle (5, 3) node [midway] {1};
	\draw[fill=gray] (5, 2) rectangle (10, 3) node [midway] {2};
	\draw[fill=gray] (10, 2) rectangle (15, 3) node [midway] {3};
	\draw[fill=green] (15, 2) rectangle (20, 3) node [midway] {4};
	\draw (20, 2) rectangle (40, 3) node [midway] {\ldots};
	\draw[fill=gray] (40, 2) rectangle (45, 3) node [midway] {1};
	\draw[fill=gray] (45, 2) rectangle (50, 3) node [midway] {2};
	\draw[fill=gray] (50, 2) rectangle (55, 3) node [midway] {3};
	\draw[fill=green] (55, 2) rectangle (60, 3) node [midway] {4\tick};
	\draw (60, 2) rectangle (80, 3) node [midway] {\ldots};

	\draw (-3, 0.5) node {(c)};
	\draw[fill=gray] (0, 0) rectangle (5, 1) node [midway] {1};
	\draw[fill=gray] (5, 0) rectangle (10, 1) node [midway] {2};
	\draw[fill=gray] (10, 0) rectangle (15, 1) node [midway] {3};
	\draw[fill=green] (15, 0) rectangle (25, 1) node [midway] {4 \tick};
	\draw (25, 0) rectangle (40, 1) node [midway] {\ldots};
\end{tikzpicture}
\caption{Several strategy schedules and their effects. (a) Failing strategies might block a successful strategy. (b) Round-robin scheduling can help mitigate this problem. (c) This approach is improved if promising strategies run for longer.}
\label{fig:blocking}
\end{figure}

In this work we first demonstrate correlation between trends in dynamic properties of proof search, and the pending success or failure of a strategy.
We then utilise this to implement strategy scheduling, prioritising those strategies most likely to succeed.
This approach differs from previous work~\cite{E,static1,static2} which attempts to predict successful strategies \emph{a priori} from static features of the input problem; instead we tip running strategies for success based on dynamic, run-time features and use this information to make decisions at runtime.

\section{Data collection and processing}
In order to identify a correlation, some data is required.
Modifying Vampire to log execution data (such as memory usage, or more specific metrics such as the number of generated clauses) for different strategies obtained from its primary portfolio mode\footnote{CASC-mode, a portfolio designed for the CASC competition~\cite{CASC}.} is straightforward, but some data-collection decisions were made:
\begin{itemize}
	\item Only numerical data immediately available in the prover was collected, but there is scope here for both non-numeric and derivative data sources, which may provide greater insight into the proof state in future work.
	\item Data was collected at intervals of a fixed number of internal resolution steps. This may not necessarily correspond to fixed time intervals, as each step may be more or less expensive, depending on the strategy and the problem.
	\item All available data was collected, even if it emerged to be constant or unhelpful. This allowed an agnostic approach to learning in which the neural network training procedure selected relevant features.
\end{itemize}
In all, 376 features are recorded.
The execution traces produced are difficult to work with, however: some are short or non-existent, others are extremely lengthy.
Some feature values also have an extremely high variance.
To deal with these problems, a post-processing step is applied.
For each feature, the mean over the entire data set is then mapped to 0, and the dataset is scaled to unit variance.
Data that are too short are discarded (the strategy likely did not take very long in any case), and then a time average is taken over 10 ``buckets'' to produce a fixed trace size.
An example post-processed trace is shown in figure \ref{fig:trace}.

\begin{figure}
	\includegraphics[width=\textwidth]{plot}
	\caption{An example trace, displayed as a colourmap, after post-processing.}
	\label{fig:trace}
\end{figure}

\section{Predicting successful strategies}
Being able to predict which traces are going to succeed in time, and which will fail at first seems unlikely.
However, it is known that the ``slowly-growing search space'' maxim, which states that strategies which minimise the number of derived clauses over time are more likely to succeed, is an effective heuristic for finding good strategies in saturation-based theorem proving~\cite{predict-success}.
Since the data we use includes the number of derived clauses, among many other features, it appears more plausible that this approach might work at least as well as the slow-growth heuristic alone.
Engineering a prediction algorithm that attempts to partition traces into ``succeeding'' and ``failing'' classes is possible with the use of modern machine-learning techniques.
Conveniently, these methods do not usually produce a binary output, but instead some \(f(\mathbf{X}) \in \left[0, 1\right]\) which might be seen as the ``level of confidence'' in success of the trace, \(\mathbf{X}\).
This success score can be used to apply an ordering to executing strategies, allowing ``smart'' dynamic scheduling of strategies.

In particular, we evaluated:
\begin{enumerate}
	\item A simple neural network with one input for each datum in the trace and a single hidden layer.
	\item A convolutional network~\cite{cnn} which performed a 1-dimensional convolutional pass along the time axis for each feature before the hidden layer.
	\item A recurrent network~\cite{gru}, feeding the time series into a gated recurrent unit before processing.
\end{enumerate}
Results for this classification task are shown in figure \ref{fig:xvalidation}.
Both of the more-advanced classifiers performed better than the simple neural network.
However, the simple network was chosen for integration into Vampire for implementation simplicity, and for performance reasons --- it is ``good enough''.

\begin{figure}
	\centering
	\begin{tabular}{c c c}
		Method & Mean Accuracy & Standard Deviation\\
		\hline
		Simple neural network & 81.5\% & 2.0\%\\
		Convolutional network & 82.4\% & 3.1\%\\
		Recurrent network & 83.9\% & 1.9\%\\
	\end{tabular}
	\caption{\(k\)-fold cross-validation classification accuracy on a balanced dataset of succeeding and failing execution traces. \(k=5\).}
	\label{fig:xvalidation}
\end{figure}

\section{Intelligent scheduling for Vampire}
We show that this abstract predictor can be used in a concrete implementation for the Vampire prover.
In the modified prover, it is used to run several strategies from Vampire's portfolio in a modified scheduler: strategies self-report their own execution data to the supervisor process and halt for re-evaluation at regular intervals.
When a strategy halts, the scheduler then decides whether to re-schedule the strategy for some more time, or to swap it out for a different strategy.
The algorithm used is as follows, taking as input a set of strategies to run:
\begin{enumerate}
	\item Initialize an empty ``run pool'' of processes, with a maximum size equal to the desired number of workers (e.g. CPU cores available). Also initialize an empty priority queue of paused processes.
	\item Whenever the pool is under capacity, take the following steps:
	\begin{enumerate}
		\item If the best process in the paused queue has a priority greater than the static priority \(p_\textrm{static}\), wake it and move it into the running pool. In tests, a static priority of around 0.5 appeared to work best.
		\item Otherwise, take a new strategy from the input set and start a new process to run that strategy.
		\item If the input set is empty, take the process from the queue regardless of its priority.
	\end{enumerate}
	\item When a strategy pauses:
	\begin{enumerate}
		\item Remove the process from the pool.
		\item Re-evaluate the process priority using the neural network and the data it provided.
		\item Insert the process into the queue with the computed priority.
	\end{enumerate}
	\item When a strategy terminates, check if it succeeded. Otherwise, remove it from the pool.
	\item If the pool, the queue, and the set of input strategies are all depleted, all strategies have failed. Exit.
\end{enumerate}

\section{Benchmark results}
The modified prover (\(M\)) was compared against a recent release of Vampire (\(V\)) under identical conditions on the TPTP benchmark, containing around 20,000 relevant problems.
The two provers performed similarly in terms of the number of problems solved.
The baseline \(V\) proved 585 theorems that \(M\) did not, while \(M\) solved 111 theorems that \(V\) did not.
We expect that this discrepancy is a combination of any remaining bugs in the prover, ``unfortunate'' predictor performance in a small number of cases, and/or the overhead of computing predictions.

However, in terms of time taken to solve problems, \(M\) was significantly faster.
Restricting the dataset to problems that both provers could solve, \(V\) took 10.5 seconds on average, while \(M\) took 8.7.

\section{Conclusions and future work}
The aim of these experiments was to improve Vampire's overall performance, if not in the number of total theorems proved (which will only decrease in this experiment --- if the entire strategy schedule runs and fails, it doesn't matter which way it is ordered), but in the average time taken to prove problems.
This approach has been shown to produce a significant increase in speed without an excessive penalty in the number of problems solved.

There are several routes that could be explored in order to further improve performance.
As well as improving predictor performance by use of more sophisticated data curation, processing, and machine-learning techniques, it may also be possible to improve the na\"ive scheduling algorithm.
Further research might include designing scheduling algorithms which keep predictions as up-to-date as possible, maximise processor utilisation, minimise memory usage/swapping, reduce context-switching overhead, or even minimise the number of required calls to the prediction algorithm.

This form of optimisation for Vampire is relatively novel: historically the aim of the team has been to prove as many theorems as possible, rather than to improve the speed of moderately-hard problem solving.
These developments may be useful in improving Vampire's performance in the new SLH division in the CASC competition, as well as improving the overall usability of ITP via quicker ``hammer'' results, such as those reported via Sledgehammer~\cite{sledgehammer}.

\bibliographystyle{plain}
\bibliography{references}
\end{document}

### 2019-02-12 coursework/experiment-design/design.tex
import from old repository

\documentclass{article}
\usepackage[a4paper]{geometry}
\title{COMP80131 Assignment 1\\Dynamic Strategy Priority for Vampire}
\author{Michael Rawson}
\begin{document}
\maketitle

\section{Task}
Modern automated theorem provers (e.g. Vampire~\cite{Vampire}) use a fixed \emph{portfolio} of strategies~\cite{portfolio} to direct search for a proof, of which only a few might solve a hard problem quickly.
Portfolio provers execute their strategies in sequence, with a pre-defined order, and a time allocation for each strategy.
However, it is not known whether a strategy will succeed, and is is often the case that a failing strategy is left to churn for considerable time without interruption~\cite{predict-success}.
We investigate the feasibility of training a machine-learned classifier to predict the success or failure of a strategy from data drawn from an initial period of its execution.

\section{Rationale}
If it proves possible to classify running strategies, it will then allow the construction of a \emph{portfolio scheduler} which defers execution of strategies the classifier deems less likely to succeed until after more-promising strategies have run.
In this way, Vampire will find proofs faster on average, theoretically without losing any proofs it found previously.
The experiment will answer ``can one of a (fixed) list of machine-learning methods produce a statistically-significant classifier for this problem'' --- note that no fixed accuracy is required, as any working classifier will improve the current state of Vampire.
We hypothesise that at least one method will beat chance.
The expectation is that most methods will succeed in this, but that no classifier will achieve a high level of accuracy, as intuitively this is a ``hard problem'', similar to that of the undecidable \emph{Entscheidungsproblem}~\cite{turing}.

\section{Methodology}
The experiment will measure the performance of a fixed set of supervised classification methods.
Classifiers will be trained on a balanced set of Vampire strategy execution traces, tagged with the strategy's eventual success or failure.
Performance may then be evaluated by a variety of statistical methods, but all involve using the trained classifier with an ``unseen'' or ``test'' set (to avoid evaluating the classifier's tendency to overfit~\cite{overfitting}) and observing the number of correct vs. incorrect classifications.

\section{Data}
The training (and test) data will be obtained by running a modified version of Vampire on a set of problems~\cite{TPTP} designed for the CASC~\cite{CASC} benchmark.
This version will log execution data and the success or failure of strategies.
Since there are a large number of strategies and many problems, sufficient data may be obtained easily.

\section{Repetition}
Vampire is now open-source, and results from this experiment will be published.
Therefore, repetition by other researchers is possible and encouraged.
Data will be available, but may also be recreated without significant effort.

\section{Analysis}
Classifier performance will be evaluated by means of \(k\)-fold cross-validation~\cite{cross-validation}.
This will allow for both reliable estimates of classifier accuracy from all available data, and also calculation of statistical significance tests such as a \(p\)-value from cross-validation results.
If a classifier exceeds the null hypothesis with high statistical significance, the hypothesis may be accepted.

\bibliographystyle{plain}
\bibliography{design.bib}
\end{document}

### 2019-02-12 coursework/progress-report/report.tex
import from old repository

\documentclass[a4paper, 10pt]{article}
\usepackage[sorting=none]{biblatex}
\usepackage{graphicx}

\title{Research Progress Report}
\author{Michael Rawson\\3-year PhD, year 1\\ORCID: 0000--0001--7834--1567}
\date{June 2018}
\addbibresource{report-references.bib}
\begin{document}
\maketitle
\section{Background}
I research the application of machine learning techniques to automated theorem provers for first-order logic.

\subsection{Introduction}
Automated provers belong to the field of automated reasoning (see \cite{survey} for an overview), which broadly aims to automate various kinds of reasoning, such as proof (in various logics)~\cite{LEO, Satallax, Isabelle, E, iProver, Vampire, CVC4, Spass, Z3}, disproof by counter-example~\cite{Nitpick, Refute, MACE}, invariant generation~\cite{invariant}, or even abduction~\cite{science-abduction, abduction}.
Theorem provers have many possible applications, such as formal verification of both mathematics and software~\cite{survey}, assisting (or even replacing) manual efforts in interactive theorem proving~\cite{Isabelle}, and a variety of other topics in artificial intelligence, such as planning~\cite{planning}.

Machine learning aims to produce systems which can ``learn'' from some existing data without explicit programming.
I focus generally on \emph{supervised learning} (see \cite{supervised-learning} for an overview), in which the data ingested by a system is pre-labelled with the desired learning outcome, as opposed to \emph{unsupervised learning} in which a system learns structure of unlabelled data.

I aim to partially re-create the process of a human mathematician's training in an automated context by combining the two fields.
Mathematicians typically learn from studying many hundreds or thousands of proofs over the course of their training, gaining experience and hence improving their ability to create proofs of their own.
By training an automated prover that can learn from experience on previous proofs, I hope to improve the performance of the prover on similar, but non-identical problems.

\subsection{Previous Work}
The idea of ``intelligent'' provers has existed for some time~\cite{survey}, but it is only relatively recently that the idea began to be taken seriously, perhaps spurred on by recent successes in other problems previously thought intractable, such as computer Go~\cite{alphago}.
The sub-field therefore remains relatively unexplored.

Bridge produced typifying results and an overview of related work in his PhD thesis~\cite{bridge}, in which he achieves limited success in two main contributions:
\begin{enumerate}
	\item Binary classification of input conjectures into ``solved by E in 300 seconds'' or ``not solved'' --- intuitively, ``easy'' or ``hard''.
	\item Intelligently selecting one of a fixed set of E strategies based on the input conjecture in order to improve E performance.
\end{enumerate}
The methodology in this work is standard for this area: structured input formulae are transformed by some fixed function into a feature vector, followed by application of a classical machine-learning algorithm, in this case a support-vector machine~\cite{SVM}.
Work in this area has tended to fall into one of three sub-areas:

\subsubsection{Axiom Selection}
Axiom selection attempts to reduce the number of axioms from a large theory base that are given to a prover trying to show a conjecture --- automated provers tend to ``choke'' on large input problems, so reducing the input to only those axioms that are required for a proof will improve prover performance.
There is opportunity here to use a machine-learning method to select axioms~\cite{axiom-selection-survey}, in a similar way to learning-assisted information retrieval algorithms.
A number of recent works contribute to this area with a variety of techniques, including a na\"ive Bayesian approach~\cite{axiom-selection-bayes, axiom-selection-sledgehammer}, deep sequence models~\cite{axiom-selection-sequence} and an approach based on a combination of a graph embedding and a deep neural network~\cite{axiom-selection-graph}.

\subsubsection{Heuristic Selection}
Automated theorem provers tend to have collections of heuristics (and parameters for those heuristics, which increases the heuristic space to an intractable size)~\cite{MaLeS} which can affect proof search to such an extent that for a given prover/problem pair, only a few of the available heuristics might solve the problem.
Therefore, heuristic selection is an important area to improve on with the use of machine learning techniques.

Generalizing Bridge's heuristic selection, The MaLeS project~\cite{MaLeS} is the main work in this area, which provides a generic framework for tuning any given theorem prover, and has had notable success in benchmarks.
It uses a kernel-based method to select a heuristic schedule, after which the prover is run uninterrupted with the schedule.

\subsubsection{Direct Proof Guidance}
Axiom and heuristic selection affect proof search indirectly: after the initial prediction, the prover is left to run as normal.
There is some limited work on directly guiding proof search with a machine-learnt heuristic.
The MaLeCoP connection prover~\cite{MaLeCoP} and its successor, FEMaLeCoP~\cite{FEMaLeCoP} add a Bayesian heuristic into a connection prover system, with some success on the MPTP benchmark~\cite{MPTP} of mathematics problems.
In a resolution context, \textcite{deep-proof-search} added a deep neural network to select clauses in the E theorem prover, again with success in MPTP.

\section{Progress}
\subsection{Research}
Over the first half of this year, I worked on an apparently-novel approach to heuristic selection, in which heuristic strategies are scheduled by an intelligent scheduler during proof search, as opposed to previous work in which a strategy schedule is generated before proof search begins.
This allows the use of run-time prover data (memory use, size of clause sets, number of clauses processed and so on) to select promising strategies and run them first.
The scheduler could be trained by recording prover data on previous runs, then tagging it with success or failure of the run.
We implemented a prototype of this idea in the Vampire system and ran a number of experiments in order to determine the new system's performance and the effects of different scheduling parameters.

While this work was successful and interesting, in recent months I have come to think that more can be achieved in the area of direct proof guidance, and hence I've moved into conducting research in this direction.
However, I do not feel that this is a totally different research area: much of the methodology and literature is similar and the research itself is arguably two different ways of achieving the same goal.

Looking at existing work in the area of direct proof guidance, two problems emerge that I aim to solve in future work, and hence improve proof guidance:
\begin{enumerate}
	\item Almost all work (with the exception of \textcite{deep-proof-search}) does not use the structure of the current proof state itself, but instead uses artificial, ``lossy'' proxies for proof state. The current trend in machine learning is away from such an approach, preferring to represent raw data in a manner convenient for a deep neural network --- but representing proof state in this way is non-obvious.
	\item Prover implementations so far have suffered from latency introduced by running the learnt heuristic --- \textcite{deep-proof-search} in particular suffered from this and was forced to introduce a cut-off period after which proof search would no longer be guided.
\end{enumerate}

I believe the first problem, that of representing proof state in a neural network effectively, can be solved by a combination of graph embedding~\cite{axiom-selection-graph} and graph convolutional networks~\cite{GCN}.
Initial experimental results are promising.

It is my current research hypothesis that the second problem of latency can be solved by a new, distributed prover architecture, sketched in Figure \ref{fig:architecture}.
Heuristic latency becomes irrelevant if the heuristic can be evaluated in parallel.
At the time of writing I am working on a prototype system implementing this architecture, which is within a couple of weeks of becoming operational.

The prototype works by exploring a tableaux search space.
Search is stochastic, weighted by a heuristic value at each node which estimates ``how many steps from this node to a proof''.
This heuristic value for a given node is computed by dispatching the proof state at the node to an external heuristic cluster.
In order to parallelise this prover, exploration of a node pauses while the heuristic is computed.
Other areas of the tree can then be explored in parallel.

\begin{figure}
	\label{fig:architecture}
	\centering
	\includegraphics[width=0.8\textwidth]{architecture}
	\caption{A cartoon showing information flow in the proposed system.}
\end{figure}

\subsection{Publications}
The work on dynamic heuristic scheduling yielded a conference talk at AITP 2018~\cite{AITP}, and will also be presented and published at the PAAR 2018 workshop this summer.
I also presented initial findings for a currently-abandoned research direction at ARW 2018~\cite{ARW}.

\subsection{Thesis Sections}
Since I now have a reasonable grasp of previous work in this area, and some published material, the introductory and background thesis sections could be completed, along with a content section discussing the work on heuristic scheduling.

\section{Planning}
\subsection{Research Timetable}
Looking forward, I plan to finish the prototype of the distributed prover as soon as possible, at least by the start of the next year.
From there, a number of promising research directions appear, of which I can explore any number of the most promising:
\begin{itemize}
	\item Further machine-learning techniques and research for a better heuristic function.
	\item Engineering for efficiency.
	\item Engineering for arbitrary scaling.
	\item Different proof calculi: which works best for heuristic search?
	\item More powerful (or arbitrary) logics.
	\item Integration with existing provers and counter-model generators.
	\item Application to novel domains
\end{itemize}
Any of these directions may be backed up with a suitable benchmark.
I plan to explore these directions for the majority of the coming year, hopefully producing significant results that can be published in higher-impact venues than those this year, such as CADE or IJCAI.
This leaves the last year for exploring any remaining avenues and writing/compiling/submitting a thesis, with some considerable time left as a buffer which will be filled with further research if no issues arise.

\subsection{Planned Thesis Sections}
The approximate chapter layout of my thesis will be the following, with a brief side-step into indirect proof guidance with heuristic scheduling, followed by the main body of the work:
\begin{enumerate}
	\item Introduction
	\item Background and Previous Work
	\item Aside: Experiments with Heuristic Scheduling
	\begin{enumerate}
		\item Methodology and Engineering
		\item Experiment
		\item Results
		\item Conclusions
	\end{enumerate}
	\item Representing Proof State in Neural Networks
	\begin{enumerate}
		\item Methodology
		\item Experiments
		\item Results
		\item Conclusions
		\item Other Applications
	\end{enumerate}
	\item A ``Smart'' Theorem Prover
	\begin{enumerate}
		\item Engineering
		\item Variations (potentially more than one sub-section)
		\item Benchmarks
		\item Comparison and Conclusions
	\end{enumerate}
	\item Overall Conclusions
\end{enumerate}

\subsection{Risk Management}
The main risk to complete is the case in which my current design for distributed heuristic theorem proving is unworkable, and all attempts to fix it fail.
I consider this unlikely, but there are several ``escape routes'' in the event of this happening.
\begin{itemize}
	\item Further work on heuristic scheduling: there is plenty still left to research here.
	\item Integrate research on proof state representation into clause selection in Vampire: this is a prover that is known to work, so presumably less likely to fail.
	\item Scale down the scope of the project such that I can still show that heuristic proof search is possible, if slow. This still has applications in domains such as mathematics.
\end{itemize}

I therefore think it is improbable that I fail to submit a thesis of some form.

\section{Conclusion}
I have a good understanding of existing material in my target area (namely application of machine learning to automated theorem provers), some existing conference and publication acceptances, promising ongoing research, and future plans with a view towards completion of a thesis.
Therefore, I believe I have a good case for progression.

\printbibliography
\end{document}

### 2019-02-12 coursework/statistical-methods/report.tex
import from old repository

\documentclass{article}
\usepackage{geometry}
\title{COMP80131 Assignment 2\\Statistical Methods}
\author{Michael Rawson}
\begin{document}
\maketitle
\section{Corpulence}
I assume ``uncertainty'' to mean precisely the absolute possible error in the Corpulence Index.
Define
\[
C(m, h) = \frac{m}{h^3}
\]
Then the absolute error \(\Delta C\) is given by
\[
\Delta C(m, h) = \sqrt{\left(\frac{\partial C}{\partial m}\Delta m\right)^2 + \left(\frac{\partial C}{\partial h}\Delta h\right)^2}
\]
which simplifies to
\[
\Delta C(m, h) = \frac{\sqrt{h^2\Delta m^2 + 9m^2\Delta h^2}}{h^4}
\]
Since \(\Delta m, \Delta h\) are known, this may be used to compute an absolute error for a given mass and height.
For individual 1, they have a height of 1.6736m, and a mass of 62.51kg.
Their precise Corpulence Index without allowing for error is 13.335045361004468\ldots and the error in the measurement is 0.7183773671149155\ldots --- therefore, the Corpulence Index may be given as 13.3 \(\pm\) 0.7.

The second sentence is ambiguous.
I interpret it to mean ``compute the mean, variance and uncertainty in the mean for each of height, mass and CI''.
Uncertainty in the mean is given by \(\frac{\sigma}{\sqrt{N}}\), where \(\sigma^2\) is the variance.
\begin{center}
\begin{tabular}{l c c c}
& Mean & Variance & Uncertainty in the Mean \\
\hline
Height (m) & 1.775 & 0.003108 & 0.00544\\
Mass (kg) & 80.34 & 132.2 & 1.12\\
CI (kg\(m^{-3}\)) & 14.46 & 6.489 & 0.249\\
\end{tabular}
\end{center}

\section{Clinical Trial}
The total number of patients in each case is 49.
With no treatment, 39 patients recovered, compared to 41 with a placebo, and 43 of those treated.
These data suggest that the ``placebo effect'' has some effect in treating this illness, and that the treatment also has an effect, greater than that of the placebo --- however, these results may not be statistically significant.
Additionally, the data also suggests that the majority (nearly 80\%) will recover without intervention.

I compute the probability that recovering with a placebo and recovering with the treatment are equally likely --- if this is high, the data does not \emph{strongly} suggest that the treatment is more effective than the placebo.
To do this, I make some assumptions: I assume that testing a series of individuals for recovery (either from the placebo or the treatment set) forms a Bernoulli process, and hence that the number of recoveries in a sample of size \(N\) (e.g. 49) is binomially-distributed with a prior probability \(p\).
Hence we can apply Fisher's exact test (\(\chi^2\) could also be appropriate) to the following contingency table:

\begin{center}
\begin{tabular}{l c c}
& Placebo & Treatment\\
\hline
Recovered & 41 & 43\\
Did Not Recover & 8 & 6\\
\end{tabular}
\end{center}

The exact test returns \(p = 0.7738\), which does not imply at any significance level that the treatment has a different probability of success than the placebo.

From this analysis, I would recommend gathering more data if possible in order to try and establish a more-significant marginal advantage for the treatment.
It also appears medically suspect that there is a binary ``recovery'' variable, especially considering the demise of those untreated --- this is a very severe ``illness'', on par with the plague! Those who do not perish are unlikely to have recovered fully, so I would recommend recording more nuanced data about patients.

\section{Algorithms}
First, I compute both the mean and the standard deviation for all the algorithms presented.
The author's algorithm has the quickest average runtime over the 500 samples, 44.28s, with a standard deviation of 17.62.
However, the closest competitor (State-of-the-Art 1) is not much slower at 48.13s, with a standard deviation of 17.32, slightly less variable.
None of the other algorithms are as fast or as consistent as either of these.

To establish whether the author's algorithm performance is \emph{significantly} different (we have already shown it to be faster in this dataset) than the competitor (assuming that the runs are representative of algorithm ``performance'', and that the samples follow a normal distribution), I perform a two-sample \(t\)-test to establish a \(p\)-value for this hypothesis.
Using the formulae for standard error, \(t\), and degrees of freedom from lectures, the standard error is 1.105, \(t = -3.48\), and there are 997 degrees of freedom.
Converting this to a \(p\)-value, I obtain \(p = 0.0002\), a highly-significant result.

It should be noted that whether this is of practical significance depends greatly on the target domain.
\end{document}

### 2019-02-12 papers/AITP18/abstract.tex
import from old repository

\documentclass{easychair}
\usepackage{cite}
\newcommand{\Vampire}{{Vampire}}
\newcommand{\vampire}{\Vampire}
\newcommand{\Spass}{{\sc Spass}}
\newcommand{\iprover}{iProver}
\newcommand{\E}{{\sc E}}
\newcommand{\zthree}{{\sc Z3}}

\title{Dynamic strategy priority: \\ empower the strong and abandon the weak}
\author{
 Michael Rawson \and 
 Giles Reger
}
\institute{University of Manchester, Manchester, UK}
\authorrunning{Rawson and Reger}
\titlerunning{Dynamic Strategy Priority}

\begin{document}
\maketitle
Many modern automated theorem provers (e.g.\ CVC4\footnote{Whilst SMT solvers are less reliant on heuristic strategies than saturation-based techniques, they still typically employ various strategies, for example for quantifier instantiation.} \cite{CVC4}, E{}~\cite{E}, \iprover{}~\cite{iProver}, \vampire{}~\cite{Vampire}) rely on \emph{portfolio modes}~\cite{portfolio} utilising from tens to hundreds of distinct strategies, of which only a few might solve a hard problem quickly.
Typically, a portfolio of strategies has a pre-defined order that the prover will execute the strategies in, running each until a strategy succeeds. 
Portfolios are important as, in practice, there is no best strategy i.e.\ it is uncommon that two hard problems are efficiently solvable by the same strategy.
However, portfolio execution is not without problems: selecting the optimal ordering and time allocation is hard in general~\cite{predict-success}, and produces overly-rigid, brittle engineering when applied to specific domains, such as those found in TPTP~\cite{TPTP}.
Moreover, for any particular problem, some lengthy strategies that are successful on other problems are doomed to failure from the outset, but are left to run unchecked by the prover, wasting time that could be spent on more productive strategies.

In this work we first demonstrate correlation between trends in dynamic properties of proof search, and the success or failure of a strategy.
We then utilise this to implement strategy scheduling, prioritising those strategies most likely to succeed.
This approach differs from previous work~\cite{E,static1,static2} which attempts to predict successful strategies \emph{a priori} from static features of the input problem; instead we tip running strategies for success based on run-time features and use this information to make scheduling decisions. 
Initial experiments on Vampire produce a performant neural-network that achieves classification accuracy of 81\% (\(\pm 2\%\)).

\paragraph{Obtaining and filtering Vampire execution data.}
Modifying Vampire to log execution data for different strategies obtained from its primary portfolio mode\footnote{CASC-mode, a portfolio designed for the CASC competition~\cite{CASC}.} is straightforward, but there are choices to be made along the way.
First, which data \emph{sources} are interesting?
As a proof of concept, we focus mostly on the numerical data (e.g.\ the number of generated clauses) that is immediately available in the prover environment, but there is scope here for non-numeric and/or derived data sources that could provide greater insight into the proof state.
Data was logged at a fixed interval of resolution steps --- unfortunately, this does not necessarily correspond to a fixed amount of time. Addressing this is left as future work.
Overall, this methodology produces extremely voluminous, irregular data.
We chose to apply a rolling time average to normalise/reduce trace size, then allow the neural net to do its own feature selection.

\paragraph{Identifying and predicting successful strategies.}
Being able to predict which strategies are going to succeed, and which will fail (or exceed the time limit) at first seems unlikely.
However, it is known that the ``slowly-growing search space'' maxim, which states that strategies which minimise the number of derived clauses over time are more likely to succeed, is an effective heuristic for finding good strategies in saturation-based theorem proving~\cite{predict-success}.
Since the data we use includes the number of derived clauses, among many other features, it appears more plausible that a machine-learnt approach might work at least as well as the slow-growth heuristic alone.
We engineer a prediction algorithm that attempts to partition traces into ``succeeding'' and ``failing'' classes using a simple neural network. 
Conveniently, these methods do not usually produce a binary output, but instead some \(f(\mathbf{X}) \in \left[0, 1\right]\) which might be seen as the ``level of confidence'' in success of the trace \(\mathbf{X}\).
This success score can be used to apply an ordering to executing strategies, allowing ``smart'' dynamic scheduling of strategies.

\paragraph{Smart scheduling for Vampire}
We show that this abstract predictor can be used in a concrete implementation for the Vampire prover.
In the modified prover, it is used to run several strategies from Vampire's portfolio in a time-slicing scheduler: at each slice, the most promising strategies are run.
The eventual aim is to improve Vampire's overall performance, if not in the number of total theorems proved (this will likely not change in this experiment --- if the entire strategy schedule runs and fails, it doesn't matter which way it is ordered), but in the time taken to prove problems.
Current benchmark results indicate a 10\% average improvement in time, in exchange for losing some problems in the benchmark.

For this demonstration we focus on maximising the accuracy and efficiency of the predictor (a perfect predictor would schedule the best strategy first, every time!), rather than tweaking performance of the scheduler.
As well as improvements to our prediction techniques, further research might include designing scheduling algorithms which keep predictions as up-to-date as possible, maximise processor utilisation, minimise memory usage/swapping, reduce context-switching overhead, or even minimise calls to the predictor, and observing the effect on prover performance.

\bibliographystyle{plain}
\bibliography{abstract}
\end{document}

### 2019-02-12 papers/AITP19/abstract.tex
import from old repository

\documentclass{easychair}
\usepackage{cite}
\newcommand{\Vampire}{{Vampire}}
\newcommand{\vampire}{\Vampire}
\newcommand{\Spass}{{\sc Spass}}
\newcommand{\iprover}{iProver}
\newcommand{\E}{{\sc E}}
\newcommand{\zthree}{{\sc Z3}}

\title{Towards an Efficient Architecture for\\ Intelligent Theorem Provers}
\author{
 Michael Rawson \and 
 Giles Reger
}
\institute{University of Manchester, Manchester, UK}
\authorrunning{Rawson and Reger}
\titlerunning{Efficient Intelligent Provers}

\begin{document}
\maketitle
High-performance automated theorem provers for first-order logic (e.g. CVC4~\cite{CVC4}, E{}~\cite{E}, \iprover{}~\cite{iProver}, \vampire{}~\cite{Vampire}) include hand-coded heuristics to guide proof search, often exposed as individual prover options.
These heuristics perform well, but have a number of disadvantages including a lack of generality over problems (necessitating \emph{portfolio} modes~\cite{RawsonReger:PAAR2018,portfolio}), inability to learn from experience, and maintenance overhead.
There is therefore interest in employing machine-learning techniques to guide proof search in automatic theorem provers, with approaches such as FEMaLeCoP~\cite{femalecop}, ENIGMA~\cite{enigma}, or Deep Network Guided Proof Search~\cite{dngps} (DNGPS).

These systems experience a trade-off between the expressivity of their learning algorithms versus the impact of guidance on ``raw'' prover performance.
At extremes:

\begin{itemize}
	\item The heuristic is fast, but does not take into account the entire proof state (e.g. the MaLeCoP family), restricting the prover to learning from features.
	\item The heuristic takes into account the entire proof state (usually via neural networks), but is too slow to use all the time. The DNGPS system runs with the heuristic for a fixed amount of time, then reverts back to the old heuristics thereafter.
\end{itemize}

\noindent
Ideally, an intelligent system would guide search based on the structure of the current proof state, while also remaining performant enough to run continuously without significantly affecting prover performance.
We present a prover architecture which attempts to achieve this ideal, and show that it has several other desirable properties.

\paragraph{Desiderata}
In such a system we require the following:
\begin{enumerate}
	\item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of resources. Saturation-based provers such as E or Vampire can have very large proof states, for example.
	\item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms tend to operate more efficiently in batches. Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
	\item \emph{Subgoals must be independent.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau or connection provers), these should be independent of the rest of the search space. Otherwise, the learning system is trying to learn while blind to the context of the search.
\end{enumerate}

\paragraph{Calculus and Algorithm}
We implement a first-order tableau calculus without unification, with equality, on non-clausal formulae.
By using this very ``natural'' representation, the hope is that inherent proof structure will be more apparent to machine-learning algorithms, which do not have to invert the process of clausification.
The tableau space is explored in parallel by means of a UCT-maximising tree search (similar to that employed by MonteCoP~\cite{montecop}), with new goals placed on a global queue for evaluation in batches by means of arbitrary machine-learning methods, currently a GCN~\cite{GCN, gcn-premise}.

\paragraph{Advantages}
This prover architecture satisfies requirements 1--3, but also shows promise in other areas.
In terms of reasoning, the calculus used is relatively flexible, allowing for extension to reasoning with theories, induction, and full higher-order logic without modifying the whole prover.
In terms of efficiency, such a prover can also make full use of multi-core systems, allowing for linear exploration speedup with the number of available cores, eventually saturating the device or core used for running machine-learned algorithms.
The prover is also well-suited to a hybrid approach in which promising subgoals are dispatched to an existing first-order ATP.

\paragraph{Evaluation and Future Work}
Evaluation and implementation of an example prover system based on this architecture is ongoing, but initial results are promising, with the system appearing to ``learn to prove'' harder problems based on prior experience with easier problems.
Future work includes exploring calculus options, optimisation, further exploration of machine-learning methods, and using the prover as a ``pre-processor'' for an existing first-order ATP.

\vspace{-1em}

\bibliographystyle{plain}
\bibliography{abstract}
\end{document}

### 2019-02-12 papers/ARW18/poster.tex
import from old repository

\documentclass[24pt, a1paper, portrait]{tikzposter}
\tikzposterlatexaffectionproofoff
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{enumitem}
\usetikzlibrary{shapes, arrows}
\usepackage[sfdefault]{roboto}
\usepackage[T1]{fontenc}

\usetheme{Simple}
\definecolorstyle{manchester}
{
	\definecolor{highlight}{rgb}{0.414, 0.171, 0.566}
}
{
	\colorlet{titlebgcolor}{highlight}
	\colorlet{blocktitlefgcolor}{highlight}
}
\usecolorstyle{manchester}

\tikzstyle{decision} = [diamond, fill=highlight, text=white, align=center, node distance=15cm]
\tikzstyle{block} = [rectangle, fill=highlight, text=white, rounded corners, align=center, node distance=15cm, minimum height=4cm]
\tikzstyle{cloud} = [ellipse, fill=highlight, text=white, align=center, node distance=15cm, minimum height=4cm]
\tikzstyle{arrow} = [->, draw, draw=highlight, line width=0.3cm]

\renewcommand\labelitemi{\large{\bullet}}
%\renewcommand\labelitemi{$\vcenter{\hbox{\Large$\bullet$}}$ }

\title{A Proof Calculus for Learned Heuristics}
\author{Michael Rawson, Giles Reger}
\date{April 2018}
\institute{University of Manchester, UK}

\begin{document}
\maketitle

\block{Abstract}
{
	Machine learning techniques may prove capable of efficiently guiding search for proofs of human theorems.
	Many modern automated theorem provers (E, {\sc Spass}, iProver, Vampire, \ldots) make use of proof calculi such as saturation or instance generation which, while efficient, are unsuitable for current machine-learning technology.
	We propose the use of a type-theoretic calculus for a future theorem prover which attempts to leverage machine-learning techniques.
}

\begin{columns}
\column{0.5}
\block{The Prover Cycle}
{
	\begin{tikzfigure}[A generalisation of information flow in a modern prover. Note the impact that a strong (learning) heuristic may have.]
		\begin{tikzpicture}
			\node [decision] (choose) {choose next\\step};
			\node [cloud, right of=choose] (heuristic) {heuristic};
			\node [block, below of=choose] (perform) {execute step};
			\node [block, right of=perform] (process) {process\\consequences};

			\path [arrow] (choose) edge node [midway, right, align=center] {proof\\step} (perform);
			\path [arrow] (perform) edge node [midway, above] {new state} (process);
			\path [arrow] (process) edge node [sloped, above] {stable state} (choose);
			\path [arrow] (process) edge node [midway, right, align=center] {prover\\data} (heuristic);
			\path [arrow] (heuristic) edge node [midway, above] {guidance} (choose);
		\end{tikzpicture}
	\end{tikzfigure}
}

\block{Existing Calculi: Problems for ML}
{
	Existing calculi such as resolution do not adapt well to current machine-learning methods, for several reasons:

	\begin{itemize}[leftmargin=1.5cm]
		\item Artificial pre-processing (CNF, Skolemisation) reduces human structure in proof state
		\item Large, recursive input space: proof state
		\item Huge number of outputs to predict: possible next steps from an unbounded set of clauses
		\item A reliance on high performance for finding proofs
	\end{itemize}
	A better calculus would address these issues while retaining the reasonably-short proofs and high performance of existing calculi.
}

\column{0.5}
\block{Type Theory for First-Order Logic}
{
	Via the Curry-Howard correspondence (``propositions-as-types''), we can encode first-order logic in a system that resembles LF.
	For example, the theorem
	\[
		(\forall x, P(x)) \implies P(c) \wedge P(d)
	\]
	might be shown with the \(\lambda\)-term
	\[
		\lambda p : (\forall x, P(x)).\ \textrm{and}(p(c), p(d))
	\]
	Proof search is now term synthesis.
}

\block{Type Theory for Proof Search}
{
	By allowing ``holes'' in the system, search for an appropriate term can proceed iteratively: consider \emph{modus ponens}, translated into a type:
	\[
		P \to (P \to Q) \to Q
	\]

	This can obtained by repeatedly refining holes, \(?\), such that the term at any point still satisfies the target type.
	Note the convergence of the inferred type onto the target type as the proof proceeds, and the limited amount of choice at each step.

\begin{center}
	\begin{tabular}{l | l}
		\\
		Refined Term & Inferred Type \\
		\(?\) & \(?\) \\
		\(\lambda x:(P).?\) & \(P \to\ ?\)\\
		\(\lambda x:(P).\lambda f:(P \to Q).?\) & \(P \to (P \to Q) \to\ ?\)\\
		\(\lambda x:(P).\lambda f:(P \to Q). \left(?\ ?\right)\) & \(P \to (P \to Q) \to\ ?\)\\
		\(\lambda x:(P).\lambda f:(P \to Q). \left(f\ ?\right)\) & \(P \to (P \to Q) \to\ Q\)\\
		\(\lambda x:(P).\lambda f:(P \to Q). \left(f\ x\right)\) & \(P \to (P \to Q) \to\ Q\)\\
		\\
	\end{tabular}
\end{center}
	The input state is now more structured, and much smaller.
	Additionally, to fill any given hole, only a subset of all available variables will be appropriate for the target type --- this reduces the output space dramatically.
	Deductions and machine-learnt heuristics may also occur in parallel, as proof steps are not inter-dependent.
}

\end{columns}
\end{document}

### 2019-02-12 papers/PAAR18/strategy-priority.tex
import from old repository

\documentclass{llncs}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,patterns}
\usepackage{subcaption}
\usepackage{pifont}
\usepackage{graphicx}

\newcommand{\vampire}{{Vampire}}
\newcommand{\iprover}{iProver}
\newcommand{\E}{{\textsc E}}
\newcommand{\zthree}{{\textsc Z3}}
\newcommand{\tick}{\ding{51}}
\newcommand{\cross}{\ding{55}}
\newcommand{\todo}[1]{{\color{red} #1}}

\title{Dynamic Strategy Priority: \\ Empower the strong and abandon the weak.}
\titlerunning{Dynamic Strategy Priority}
\authorrunning{Rawson and Reger}
\author{
 Michael Rawson \and 
 Giles Reger
}
\institute{
University of Manchester, Manchester, UK
}

\begin{document}
\maketitle
\begin{abstract}
Automated theorem provers are often used in other tools as black-boxes for discharging proof obligations. One example where this has enjoyed a lot of success is within interactive theorem proving. 
%
A key usability factor in such settings is the time taken for the provers to complete.
Automated theorem provers typically run lists of proof strategies sequentially, which can cause proofs to be found more slowly than necessary if the ordering is suboptimal.
We show that it is possible to predict which strategies are likely to succeed while they are running using an artificial neural network.
We also implement a run-time strategy scheduler in the \vampire{} prover which utilises a trained neural network to improve average proof search time, and hence increases usability as a black-box prover.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%
Modern automated theorem provers (e.g. \E{}~\cite{E}, \iprover{}~\cite{iProver}, \vampire{}~\cite{Vampire}, CVC4~\cite{CVC4}) for first-order logic rely on \emph{portfolio modes}~\cite{portfolio}, which utilise tens to hundreds of distinct \emph{strategies}.
Of these strategies, only a few might solve a hard problem, often rapidly.
Typically, a portfolio of strategies has a pre-defined execution order: the prover will process strategies in this order, running each until a winning strategy is found or the prover runs out of time or strategies.

Portfolio modes are important as, in practice, there is no best strategy.
Furthermore, it is uncommon that two hard problems are efficiently solved by the same strategy.
However, portfolio execution is not without problems: deciding the optimal ordering and time allocation is hard in general~\cite{predict-success}, and produces overly-rigid, brittle engineering when applied to specific domains, such as those found in the TPTP problem set~\cite{TPTP}.
Moreover, for any particular problem, some lengthy strategies that are successful on other problems are doomed to failure from the outset --- as illustrated in Figure \ref{fig:blocking} --- but are left to run unchecked by the prover, wasting time that could be spent on more productive strategies.

\begin{figure}
\begin{tikzpicture}[x=0.01\textwidth, y=.3in]
	\draw[thick, ->] (0, 6) -- (100, 6) node {};

	\draw (-3, 4.5) node {(a)};
	\draw[fill=gray] (0, 4) rectangle (30, 5) node [midway] {1 \cross};
	\draw[fill=gray] (30, 4) rectangle (50, 5) node [midway] {2 \cross};
	\draw[fill=gray] (50, 4) rectangle (80, 5) node [midway] {3 \cross};
	\draw[fill=green] (80, 4) rectangle (90, 5) node [midway] {4 \tick};
	\draw (90, 4) rectangle (100, 5) node [midway] {\ldots};

	\draw (-3, 2.5) node {(b)};
	\draw[fill=gray] (0, 2) rectangle (5, 3) node [midway] {1};
	\draw[fill=gray] (5, 2) rectangle (10, 3) node [midway] {2};
	\draw[fill=gray] (10, 2) rectangle (15, 3) node [midway] {3};
	\draw[fill=gray] (15, 2) rectangle (20, 3) node [midway] {4};
	\draw (20, 2) rectangle (40, 3) node [midway] {\ldots};
	\draw[fill=gray] (40, 2) rectangle (45, 3) node [midway] {1};
	\draw[fill=gray] (45, 2) rectangle (50, 3) node [midway] {2};
	\draw[fill=gray] (50, 2) rectangle (55, 3) node [midway] {3};
	\draw[fill=green] (55, 2) rectangle (60, 3) node [midway] {4\tick};
	\draw (60, 2) rectangle (80, 3) node [midway] {\ldots};

	\draw (-3, 0.5) node {(c)};
	\draw[fill=gray] (0, 0) rectangle (5, 1) node [midway] {1};
	\draw[fill=gray] (5, 0) rectangle (10, 1) node [midway] {2};
	\draw[fill=gray] (10, 0) rectangle (15, 1) node [midway] {3};
	\draw[fill=green] (15, 0) rectangle (25, 1) node [midway] {4 \tick};
	\draw (25, 0) rectangle (40, 1) node [midway] {\ldots};
\end{tikzpicture}
\caption{Several strategy schedules and their effects. (a) Failing strategies might block a successful strategy. (b) Round-robin scheduling can help mitigate this problem. (c) This approach is improved if promising strategies run for longer.}
\label{fig:blocking}
\end{figure}

This work makes two contributions. 
We first demonstrate correlation between trends in dynamic properties of proof search, and the pending success or failure of a strategy (Sections~\ref{sec:data},~\ref{sec:predict}).
We then utilise this to implement strategy scheduling, prioritising those strategies most likely to succeed (Section~\ref{sec:intelligent}).
This approach differs from previous work~\cite{E,static1,static2} which attempts to predict successful strategies \emph{a priori} from static features of the input problem; instead we tip running strategies for success based on dynamic, run-time features and use this information to make decisions at runtime. Our experiments (Section~\ref{sec:experiments}) show that guiding scheduling in this way can significantly speed up portfolio-based approaches. Before introducing these contributions and results we first present necessary background in Sections~\ref{sec:vampire} and~\ref{sec:ml}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Strategies in Vampire}
\label{sec:vampire}
%
Vampire is a first-order superposition theorem prover. 
This section reviews its basic structure and components relevant to the rest of this paper. 
We will use the word \textit{strategy} to refer to a set of configuration parameter values that control proof search and \textit{proof attempt} to refer to a run of the prover using such a strategy.

%%%%%%%%%%%%%%%%%%
\subsection{Input and preprocessing}
%%%%%%%%%%%%%%%%%%

Vampire accepts problems in first-order logic with equality and a pre-defined set of first-order theories (e.g. arithmetic, arrays, datatypes). Vampire typically works with problems in the TPTP format \cite{TPTP} but also accepts problems in SMT-LIB~\cite{smt-lib}.

Problems are parsed and transformed into a set of input clauses used by the saturation algorithm.
This process involves clausification and a number of preprocessing steps.
These preprocessing steps can alter certain properties of the problem e.g. its size and distribution across the signature, sometimes significantly: for example, the E.T. system~\cite{ET} implements a pre-processor for E which selects small sets of axioms from a larger set, for large-theory reasoning.
This means that any prediction method that relies solely on the input problem (rather than dynamic characteristics such as post-preprocessing properties) will find it harder to predict the best parameters for proof search.

%Some preprocessing steps, such as naming and inequality splitting, can alter the signature. %This is important when it comes to cooperation between proof attempts. 
%Later we will see that cooperation is based on the idea that cooperating proof attempts are exploring different areas of the same search space, this would no longer be true if the original clauses differed. 
%Therefore, this work will assume that sets of strategies are \emph{preprocessing-compatible} i.e. after preprocessing the same set of input clauses are used by each strategy.

%%%%%%%%%%%%%%%%%%
\subsection{Saturation algorithms}
%%%%%%%%%%%%%%%%%%

Superposition provers such as Vampire use \emph{saturation algorithms with redundancy elimination}.
They work with a search space consisting of a set of clauses and use a collection of generating,
simplifying and deleting inferences to explore this space. 
Generating inferences, such as superposition, extend this search space by adding new clauses obtained by applying inferences to existing clauses.
Simplifying inferences, such as demodulation, replace a clause by a simpler one.
Deleting inferences, such as subsumption, delete a clause, typically when it becomes redundant (see \cite{resolutionHandbook}). Simplifying and deleting inferences must satisfy this condition to preserve completeness.

The goal is to \emph{saturate} the clause set with respect to the inference system. If the empty clause is derived then the input clauses are unsatisfiable. If no empty clause is derived and the search space is saturated then the input clauses are guaranteed to be satisfiable \emph{only if} a complete strategy is used. A strategy is complete if it is guaranteed that all inferences between non-deleted clauses in the search space will be applied. Vampire includes many incomplete strategies as they can be very efficient at finding unsatisfiability.

%Completeness relies on a fair \emph{clause selection} strategy
%In Vampire the application of inferences make heavy use of \emph{term indexing}

All saturation algorithms implemented in Vampire belong to the family of \emph{given clause algorithms}, which achieve completeness via a fair \emph{clause selection} process that prevents the indefinite skipping of old clauses. These algorithms typically divide clauses into three sets, \emph{unprocessed}, \emph{passive} and \emph{active}, and follow a simple \emph{saturation loop}:
\begin{enumerate}
	\item Add non-redundant \emph{unprocessed} clauses to \emph{passive}. Redundancy is checked by attempting to \emph{forward simplify} the new clause using processed clauses.
	\item Remove processed (passive and active) clauses made redundant by newly processed clauses, i.e.\ \emph{backward simplify} existing clauses using these clauses.
	\item Select a given clause from \emph{passive}, move it to \emph{active} 
	and perform all generating inferences between the given clause and all other active clauses,
	adding generated clauses to \emph{unprocessed}.
\end{enumerate}
Later we will show how iterations of this saturation loop from different proof attempts can be interleaved. Vampire implements three saturation  algorithms:
\begin{enumerate}
	\item \emph{Otter} uses both passive and active clauses for simplifications.
	\item \emph{Limited Resource Strategy (LRS)} \cite{LRS} extends Otter with a heuristic
	   that discards clauses that are unlikely to be used with the current resources, i.e.\ time and memory.
		 This strategy is incomplete but also generally the most effective at proving unsatisfiability.
	\item \emph{\sc Discount} uses only active clauses for simplifications.
\end{enumerate}
A recent development in Vampire is AVATAR \cite{avatar,playing} which integrates with the saturation loop to perform clause splitting. The general idea is to use a SAT solver to select a subproblem by naming each clause sub-component by a propositional variable, running a SAT solver on these abstracted clauses, and using the subsequent propositional model to select components to include in proof search. 
At the end of each iteration of the loop we check whether the underlying subproblem has changed. AVATAR can occasionally make loops run a little longer, but no more than other steps such as backward subsumption. Otherwise, the notion of saturation loop remains the same when using AVATAR.

There are also other proof strategies that fit into the above loop format 
and can be interleaved with superposition based proof attempts.
For example, instance generation \cite{instgen} saturates the set of clauses with respect to the instance generation rule and finite model finding \cite{Reger2016sat} iteratively checks larger model sizes (these loops tend to be much longer than those from other algorithms).

%%%%%%%%%%%%%%%%%%
\subsection{Strategies in Vampire}
%%%%%%%%%%%%%%%%%%

%A strategy is a set of configuration parameter values that controls proof search.
Vampire includes more than 50 parameters. By only varying parameters and values used by Vampire at the last CASC competition, we obtain over 500 million strategies. These parameters control
\begin{itemize}
	\item Preprocessing steps (24 different parameters)
	\item The saturation algorithm and related behaviour e.g. clause selection
	\item Inferences used (16 different kinds with variations)
	%\item Parameters to control splitting via AVATAR e.g. when it occurs
\end{itemize}
Even restricting these parameters to a single saturation algorithm and straightforward preprocessing steps, the number of possible strategies is vast. For this reason, Vampire implements a portfolio \emph{CASC mode} \cite{Vampire} that categorises problems based on syntactic features and attempts a sequence of approximately 50 strategies over a five minute period (this number can vary significantly). These strategies are the result of extensive benchmarking and have been shown, experimentally, to work well on unseen problems i.e. those not used for training.

The syntactic features used by the current portfolio mode are coarse-grained and include the rough size of the problem (e.g. bucketed into tiny, small, medium, large, huge), the presence of features such as equality or certain theories, and whether the problem is effectively propositional or horn. Not all combinations of these are considered. Portfolio mode is created by considering a set of training data over a list of strategies and attempting to cover as much of it as possible by splitting the set of problems into smaller groups based on these features until all solutions fit into a given time limit. This process is greedy and places the strategy that solves the most problems during training first and then the next best strategy after removing the problems solved by the first solver and so on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine Learning and Theorem Proving} \label{sec:ml}

In this section we review the necessary background in machine learning in the context of theorem proving, in particular neural networks.
Other technologies exist which are capable of similar performance, but neural networks produced the best combination of convenience and performance for our purposes.

\subsection{Classification with neural networks}
\emph{Binary classification problems} can be heuristically solved by means of \emph{supervised learning} methods.
A binary classification problem is, informally, the statement ``given input \(X\), does \(X\) belong to class \(A\) or class \(B\)'', where \(X\) is a member of the target domain \(\mathcal{D}\), and \(A, B\) are disjoint subsets of \(\mathcal{D}\) such that \(\mathcal{D} = A \cup B\).
Supervised learning methods attempt to solve binary classification problems by starting with a neutral, untrained heuristic for classifying \(X\) into \(A\) or \(B\), then changing their internal state to improve the heuristic by looking at examples of \(A\) and \(B\) (``training'').
Typically, supervised learning methods do not work on \(X\) directly, but from a set of ``features'' \emph{derived} from \(X\).
Engineering features, or choosing which of a set of supplied features to use, is a field of study in itself.

\begin{figure}[t]
	\begin{tikzpicture}[node distance=\layersep]
		\def\layersep{3cm}
		\tikzstyle{neuron} = [circle, draw, fill=black!25, minimum size=20pt]
		% input layer
		\foreach \name / \y in {1, ..., 3}
			\node[neuron, pin=left:Input \name] (input-\name) at (0, -\y cm) {};

		% hidden layer
		\foreach \name / \y in {1, ..., 5}
			\node[neuron, yshift=1cm] (hidden-\name) at (\layersep, -\y cm) {};

		% final layer
			\node[neuron, right of=hidden-3, pin={[pin edge={->}]right:Classification}] (final){};

		% input-hidden connections
		\foreach \source in {1, ..., 3}
			\foreach \target in {1, ..., 5}
				\path (input-\source) edge (hidden-\target);

		% hidden-final connections
		\foreach \source in {1, ..., 5}
			\path (hidden-\source) edge (final);

		% annotations
		\node[text centered, above of=input-1, node distance=2cm] {Input Layer};
		\node[text centered, above of=hidden-1, node distance=1cm] {Hidden Layer};
		\node[text centered, above of=final, node distance=3cm] {Output Layer};
	\end{tikzpicture}
	\centering
	\caption{A neural network with 3 inputs, a single densely-connected hidden layer, and a single output neuron. This is a type of network that might be used for simple binary classification tasks. A version of this architecture with more input and hidden neurons is used for this work.}
	\label{fig:neural-architecture}
\end{figure}
\begin{figure}
	\begin{tikzpicture}[xscale=0.75, yscale=4]
		\draw[->] (-6, 0) -- (6, 0) node[right] {$x$};
		\draw[->] (0, 0) -- (0, 1) node[above] {$\sigma(x)$};
		\draw[blue] (0, 0) plot[domain=-6:6] (\x, {1 / (1 + exp(- \x))});
	\end{tikzpicture}
	\centering
	\caption{The sigmoid function \(\sigma(x) = \frac{1}{1 + e^{-x}}\). This function ``squeezes'' the output to the range \([0, 1]\).}
	\label{fig:sigmoid}
\end{figure}

A neural network consists of a collection of artificial ``neurons'', wired together with connections into a \emph{neural network architecture} (see Figure \ref{fig:neural-architecture} for an example).
Each neuron \(n\) computes a real-valued function \(f_n\), which may be tuned with real-valued parameters: every neuron has a \emph{bias} \(b_n\), and \emph{weight vector} \(\textbf{w}_{n}\).

\(f_n\) is defined on an input vector \(\textbf{x}\) as
\[
	f_n(\textbf{x}) = A\left(\textbf{w}_n\cdot\textbf{x} + b_n\right)
\]

i.e. a weighted sum of all connected input neurons' activation functions, plus the bias of the neuron, all fed through an \emph{activation function} \(A(x)\).
The choice of activation function is dependent on both the position of the neuron in the architecture, and on the target domain: a common choice is the sigmoid function \(\sigma(x)\), as shown in Figure \ref{fig:sigmoid}.

Neural network parameters may be \emph{trained} by a variety of different techniques (randomly, neuro-evolution methods, backpropagation methods, \emph{etc.}).
We used a backpropagation method, which, given a training input \(X\) and correct output \(y\), tunes the network parameters by propagating an error signal backwards through the network to obtain updated parameters for each neuron, hence reducing the error signal for that example~\cite{backpropagation}.
Repeating this process for multiple training examples can produce very good results, but frequently only with the correct combination of learning algorithm, starting weights, network architecture, and training examples.

\subsection{Related work}
Machine learning has potential for significant effect in this area: despite the computational difficulties experienced by theorem provers, frequently proofs have human (or otherwise regular) structure, which may be exploited by the use of learning methods.
The authors are aware of work in the fields of premise selection~\cite{deepmath, mash, malarea, deep-graph-embedding}, static strategy selection~\cite{bridge, males, e-males}, and more recently, direct proof guidance~\cite{malecop, femalecop, deep-proof-search}.
However, we are not aware of any previous work in the area of strategy selection at runtime for conventional theorem provers. The closest area of work is that of static strategy selection where work, with the exception of Bridge et al. \cite{bridge} focusses on static properties of the input problem rather than dynamic properties of the proof search space. Bridge considered various dynamic features of the search space after 100 steps of the saturation algorithm in the default mode (of the E theorem prover \cite{E}).

So far, machine learning techniques have typically used classical machine-learning techniques such as na\"ive Bayesian methods, SVMs~\cite{svm}, and decision trees~\cite{decision-tree}.
However, some newer work has begun to utilise recently-developed deep-neural-network methods~\cite{deepmath, deep-graph-embedding, deep-proof-search}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Feature collection} \label{sec:data}
%
This section discusses which features are extracted from \vampire{} for prediction and how they are extracted. This information is firstly required to identify a correlation and will then be used to train a predictor.

Modifying Vampire to log execution data (such as memory usage, or more specific metrics such as the number of generated clauses) for different strategies obtained from its primary portfolio mode\footnote{CASC-mode, a portfolio designed for the CASC competition~\cite{CASC}.} is straightforward, but some data-collection decisions were made:
\begin{itemize}
	\item Only numerical data immediately available in the prover was collected, but there is scope here for both non-numeric and derivative data sources, which may provide greater insight into the proof state in future work. Suppose that quantities \(A\) and \(B\) are measured directly, but are more often discussed in terms of categories \(C\) that all \(A, B\) pairs fall into. Data could then include \(C\) as well, or instead of, \(A\) and \(B\), embedding \(C\) into the neural network via a one-hot encoding.
	\item Data was collected at intervals of a fixed number of internal resolution steps (in the experiments presented in this paper, this value was 10). This may not necessarily correspond to fixed time intervals, as each step may be more or less expensive, depending on the strategy and the problem.
	\item All available data was collected, even if it emerged to be constant or unhelpful. This allowed an agnostic approach to learning in which the neural network training procedure selected relevant features.
\end{itemize}
In all, 376 features are recorded, including the number of active clauses, passive clauses, and subsumptions, for instance.
The execution traces produced are difficult to work with, however: some are short or non-existent, others are extremely lengthy. The mean execution trace length was 536 (standard deviation: 1208) with the longest trace being 9971 recorded steps long.
Some feature values also have an extremely high variance.
To deal with these problems, a post-processing step is applied.
For each feature, the mean over the entire data set is then mapped to 0, and the dataset is scaled to unit variance.
Data that are too short (fewer than 10 steps) are discarded (the strategy likely did not take very long in any case).
The remaining data is then sliced into 10 evenly-sized ``buckets'', then an average of each bucket is taken to produce 10 values for every trace, and hence a fixed data size.

However, even now these data are not representative of the classification problem desired: these traces show \emph{completed} runs of Vampire, whereas the classifier will be used to predict the success or otherwise of runs of Vampire that are still in progress.
Hence, we take ``snapshots'' at various stages (in these experiments, at every quarter) of the trace, discarding the rest of the data, then post-process the remaining trace as described above.
Conveniently, this also provides 4 times the original number of training examples.
An example post-processed trace is shown in Figure \ref{fig:trace}.

\begin{figure}
	\includegraphics[width=\textwidth]{plot}
	\caption{An example trace, displayed as a colourmap, after post-processing.}
	\label{fig:trace}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Predicting successful strategies} \label{sec:predict}
%
Being able to predict which proof search attempts will succeed in time, and which will fail, based solely on information in the execution trace may seem unlikely.
However, it is known that the ``slowly-growing search space'' maxim, which states that strategies which minimise the number of derived clauses over time are more likely to succeed, is an effective heuristic for finding good strategies in saturation-based theorem proving~\cite{predict-success}.
Since the data we use includes the number of derived clauses, among many other features, it appears more plausible that this approach might work at least as well as the slow-growth heuristic alone.
Engineering a prediction algorithm that attempts to partition traces into ``succeeding'' and ``failing'' classes is possible with the use of modern machine-learning techniques.
Conveniently, these methods do not usually produce a binary output, but instead some \(f(\mathbf{X}) \in \left[0, 1\right]\) which might be seen as the ``level of confidence'' in success of the trace, \(\mathbf{X}\).
This success score can be used to apply an ordering to executing strategies, allowing ``smart'' dynamic scheduling of strategies.

In particular, we evaluated (using the \emph{Keras}~\cite{keras} backend-agnostic neural network suite and the \emph{scikit-learn}~\cite{scikit-learn} utilities)
\begin{enumerate}
	\item A simple neural network with one input for each datum in the trace and a single hidden layer.
	\item A convolutional network~\cite{cnn} which performs a 1-dimensional convolutional pass along the time axis for each feature before the hidden layer.
	\item A recurrent network~\cite{gru}, feeding the time series into a gated recurrent unit before processing.
\end{enumerate}
Results for this classification task are shown in Figure \ref{fig:xvalidation}. The consistent level of accuracy achieved is encouraging. 
Both of the more-advanced classifiers performed better than the simple neural network.
However, the simple network was chosen for integration into Vampire for implementation simplicity, and for performance reasons --- it is ``good enough''.

\begin{figure}
	\centering
	\begin{tabular}{c c c}
		Method & Mean Accuracy & Standard Deviation\\
		\hline
		Simple neural network & 81.5\% & 2.0\%\\
		Convolutional network & 82.4\% & 3.1\%\\
		Recurrent network & 83.9\% & 1.9\%\\
	\end{tabular}
	\caption{\(k\)-fold cross-validation classification accuracy on a balanced dataset of around 10,000 (total) succeeding and failing execution traces. \(k=5\).}
	\label{fig:xvalidation}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Intelligent scheduling for Vampire} \label{sec:intelligent}
%
We show that this abstract predictor can be used in a concrete implementation for the Vampire prover.
In the modified prover, it is used to run several strategies from Vampire's portfolio in a modified scheduler: strategies self-report their own execution data to the supervisor process and halt for re-evaluation at regular intervals.
When a strategy halts, the scheduler then decides whether to re-schedule the strategy for some more time, or to swap it out for a different strategy.
The algorithm used is as follows, taking as input a set of strategies to run: \
\begin{enumerate}
	\item Initialize an empty ``run pool'' of processes, with a maximum size equal to the desired number of workers (e.g. CPU cores available). Also initialize an empty priority queue of paused processes.
	\item Whenever the pool is under capacity, take the following steps:
	\begin{enumerate}
		\item If the best process in the paused queue has a priority greater than the static priority \(p_\textrm{static}\), wake it and move it into the running pool. In tests, a static priority of around 0.5 appeared to work best.
		\item Otherwise, take a new strategy from the input set and start a new process to run that strategy.
		\item If the input set is empty, take the process from the queue regardless of its priority.
	\end{enumerate}
	\item When a strategy pauses:
	\begin{enumerate}
		\item Remove the process from the pool.
		\item Re-evaluate the process priority using the neural network and the data it provided.
		\item Insert the process into the queue with the computed priority.
	\end{enumerate}
	\item When a strategy terminates, check if it succeeded. Otherwise, remove it from the pool.
	\item If the pool, the queue, and the set of input strategies are all depleted, all strategies have failed. Exit.
\end{enumerate}

To embed the neural network into Vampire, we took the trained network weights from our Python-based experiments, and generated a C source file with these weights included as a large array.
The neural network's architecture was then re-implemented manually in C++, using the network weights compiled into the new program.
This approach had several advantages: while perhaps not as efficient as an architecture such as TensorFlow~\cite{tensorflow} (which may use graphics hardware to accelerate computations), our approach is \emph{reasonably} efficient, but is also low-latency, does not incur any additional dependencies, and does not add significantly to program start-up time.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental results} \label{sec:experiments}

We evaluate whether our prediction results carry through to improved performance in Vampire.

\paragraph{Experimental setup} We take all relevant problems for Vampire from TPTP 6.4.0 (17,281 problems in total) and run three variations of Vampire 4.2.1's CASC portfolio mode:
\begin{itemize}
	\item \emph{baseline} was the standard portfolio mode
	\item \emph{no-prediction} uses the dynamic scheduling architecture without prediction. This effectively produces a round-robin scheduling of strategies.
	\item \emph{prediction} uses the trained neural network to predict whether a strategy will be successful as previously described.
\end{itemize}	
The default 3GB memory limit and 300s time limit from the standard portfolio mode were kept but in addition each variation was run in a \emph{multicore} setup using 4 cores to distribute strategies over.Experiments were run on the StarExec cluster \cite{starexec} where each node contains an Intel Xeon 2.4GHz processor.

\begin{table}[t]
\caption{Results for the three variations.\label{tab:numbers}}
\centering
\begin{tabular}{l|ccccc}
				& Solved & New & Unique & Average time (wall) & Average time (cpu) \\ \hline
\emph{baseline}		& 12,899 &- & 420 & 3.03s $\pm$ 9.91s &10.53s $\pm$ 33.60s\\
\emph{no-prediction}	& 11,827 & 56& 33 & 2.86s $\pm$ 8.33s&10.38s $\pm$ 32.24s\\
\emph{prediction} 	& 12,425 & 111& 88 &2.59s $\pm$ 9.55s & 8.74s $\pm$ 31.03s \\
\end{tabular}
\end{table}

\paragraph{Results} Figure~\ref{fig:results} illustrates the overall results by plotting the number of problems solved against time taken. Table~\ref{tab:numbers} gives some raw numbers. In this table \emph{New} refers to the number of problems solved by this variation that were not solved by \emph{baseline}, \emph{Unique} refers to the number of problems solved by this variation and not by the other two, and the average solution times are given for wall-clock and cpu-time with standard deviation.

Table~\ref{tab:speedup} gives further statistics comparing the two scheduling variations with the baseline. For the purposes of this table we only compare the results on problems that both the given variation and baseline solve and where the difference between solution times is greater than 1 second. This second part is important as it allows for small variations in solution times due to natural non-determinism\footnote{This allowance is often not made but it can heavily skew results. Without it all variations look almost identical as the cases where variations behave in the same way dominate. This does not indicate that the variations are not an improvement, but there are many cases where there is no difference. The most likely explanation for this is that solutions are quick and no scheduling is required e.g. if the problem is solved during preprocessing.}. Speedup is a multiplication factor (a speedup of 2 means a proof was found in half the time), similarly for slowdown. Better means that the given variation was faster, whereas worse means that the variation was slower.

\begin{figure}[t]
\centering
\includegraphics[width=0.6\textwidth]{results.jpeg}
\caption{Plot of problems solved against time for the three variations.\label{fig:results}}
\end{figure}

\begin{table}[t]
\caption{Amount of speedup or slowdown compared to the baseline solver.\label{tab:speedup}}
\centering
\begin{tabular}{l|cc|cc}
		& \multicolumn{2}{c|}{no-prediction} & \multicolumn{2}{c}{prediction} \\ \cline{2-5}
		& wall & cpu & wall & cpu \\ \hline
Average Speedup &1.54 & 6.90 & 1.94 & 6.24 \\ \hline
Number of times better & 451 & 788 & 650 & 1,297 \\
Number of times worse & 1,895 & 2,518 & 642 & 940\\ \hline
Average Speedup (when better) & 7.02 & 28.19 &  3.47& 10.46 \\
Average Slowdown (when worse) & 29.02 & 78.82 & 18.83 & 44.52\\
\end{tabular}
\end{table}

\paragraph{Discussion} An immediate observation is that the overall number of problems solved is slightly worse for the variations performing scheduling. An immediate explanation for this could be that the strategies in the portfolio mode might still be quite fragile i.e. their performance might be degrading with the small overhead of context-switching and additional memory contention. Further experiments could explore this by using more generic strategies with longer individual time limits (many strategies in CASC mode run for less than 1 second). However, the overall number of problems solved is still high enough to make the result useful. The new problems solved by the scheduling variations could also be attributed to this non-determinism, suggesting again that more effort should be spent to provide a broader set of complementary strategies to choose from. We note that one explanation for this second effect could have been if the strategy schedule was longer than the time given and prediction moved a strategy into the allowed time. This is a behaviour we might expect for short time limits but for this experiment all strategies were run.

The average solution times are improved with the two variations and the variation using prediction achieves the best solution time on average. When looking at the more detailed results of Table~\ref{tab:speedup} we see that in the majority of cases the prediction variant was faster than the baseline, which was not the case when no prediction was applied. Furthermore, the impact of getting it wrong was larger without prediction (i.e. the slowdown is bigger).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and future work} \label{sec:conclusion}
%
The aim of these experiments was to improve Vampire's overall performance, if not in the number of total theorems proved, but in the average time taken to prove problems.
This approach has been shown to produce a significant increase in speed without an excessive penalty in the number of problems solved.

There are several routes that could be explored in order to further improve performance.
As well as improving predictor performance by use of more sophisticated data curation, processing, and machine-learning techniques, it may also be possible to improve the na\"ive scheduling algorithm.
Further research might include designing scheduling algorithms which keep predictions as up-to-date as possible, maximise processor utilisation, minimise memory usage/swapping, reduce context-switching overhead, or even minimise the number of required calls to the prediction algorithm.

This form of optimisation for Vampire is relatively novel: historically the aim of the team has been to prove as many theorems as possible, rather than to improve the speed of moderately-hard problem solving.
As immediate impact, these developments may be useful in improving Vampire's performance in the new SLH division in the CASC competition, as well as improving the overall usability of ITP via quicker ``hammer'' results, such as those reported via Sledgehammer~\cite{sledgehammer}.

\bibliographystyle{plain}
\bibliography{references}
\end{document}

### 2019-02-12 papers/CADE19/paper.tex
paper outline for Giles

\documentclass{llncs}
\title{Older or heavier? Decaying gracefully with age/weight shapes.}
\titlerunning{Dynamic Strategy Priority}
\authorrunning{Reger \and Rawson}
\author{Michael Rawson \and Giles Reger}
\institute{University of Manchester, Manchester, UK}

\begin{document}
\maketitle
\begin{abstract}
Vampire~\cite{vampire} is an automatic theorem prover for first-order logic.
During proof search in given-clause algorithms, Vampire repeatedly selects clauses from either an \emph{age} or a \emph{weight} queue in a fixed, but configurable \emph{age/weight ratio} (AWR).
We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
Based on this observation we develop several new modes for Vampire which vary AWR according to a ``shape'' during proof search.
The modes solve a number of new problems in the TPTP~\cite{tptp} benchmark.
\end{abstract}

\section{Introduction}
TODO: standard Vampire fluff here.\\
TODO: explain a generic given-clause algorithm, define clause age and weight.\\
TODO: remark on portfolio modes - more choice is (sometimes) a good thing.

\section{Optimising Age/Weight Ratios}
TODO: talk about optimising for fixed age/weight.\\
TODO: talk about

\section{Variable AWR for Vampire}
TODO: describe "decay" and "converge" modes for Vampire.\\
TODO: discuss the effect of the "frequency" flag, plus alternatives.

\section{Experimental Evaluation}
TODO: analyse (two sets of) results, discuss drop in performance in exchange for uniques.

\section{Future Work}
TODO: more shapes, better ways of doing frequency decay, integrate into existing strategy schedules.

\section{Conclusions}
TODO: this is a thing, you can do it, you should do it (?), we can probably do better.

\bibliographystyle{plain}
\bibliography{references}
\end{document}

### 2019-02-14 papers/CADE19/paper.tex
talk about implementation in Vampire

@@
 \documentclass{llncs}
-\title{Older or heavier? Decaying gracefully with age/weight shapes.}
+\usepackage{graphicx}
+\usepackage[caption=false]{subfig}
+\usepackage[dvipsnames]{xcolor}
+\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
+\title{Old or Heavy?\\Decaying Gracefully with Age/Weight Shapes.}
 \titlerunning{Dynamic Strategy Priority}
@@
 \begin{abstract}
-Vampire~\cite{vampire} is an automatic theorem prover for first-order logic.
+Vampire is an automatic theorem prover for first-order logic.
 During proof search in given-clause algorithms, Vampire repeatedly selects clauses from either an \emph{age} or a \emph{weight} queue in a fixed, but configurable \emph{age/weight ratio} (AWR).
 We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
-Based on this observation we develop several new modes for Vampire which vary AWR according to a ``shape'' during proof search.
-The modes solve a number of new problems in the TPTP~\cite{tptp} benchmark.
+Based on these observations we develop several new modes for Vampire which vary AWR according to a ``shape'' during proof search.
+The modes solve a number of new problems in the TPTP benchmark.
 \end{abstract}
@@
 \section{Introduction}
-TODO: standard Vampire fluff here.\\
-TODO: explain a generic given-clause algorithm, define clause age and weight.\\
-TODO: remark on portfolio modes - more choice is (sometimes) a good thing.
+\label{sec:introduction}
+\todo{standard Vampire~\cite{vampire} fluff here}.\\
+\todo{explain a generic given-clause algorithm, define clause age and weight --- and how the ``balancing'' algorithm works internally}.\\
+\todo{remark on portfolio modes --- more choice is (sometimes) a good thing}.
 
 \section{Optimising Age/Weight Ratios}
-TODO: talk about optimising for fixed age/weight.\\
-TODO: talk about
+\todo{talk about optimising for fixed age/weight}.\\
+\todo{talk about experiments with changing age/weight}.
 
 \section{Variable AWR for Vampire}
-TODO: describe "decay" and "converge" modes for Vampire.\\
-TODO: discuss the effect of the "frequency" flag, plus alternatives.
+If changing AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
+In general we would like to describe any possible sequence that the AWR could follow during proof search.
+However, some details constrain the design space:
+\begin{enumerate}
+	\item Changing the AWR too frequently or sharply has little effect, due to the ``balancing'' algorithm --- see Section \ref{sec:introduction}.
+	\item A general (configurable) \emph{shape} is more likely to be widely applicable than a specific series of data points.
+	\item The shape must extend naturally to an indefinitely-long proof search.
+\end{enumerate}
+
+In this work we selected two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
+Since a simple linear shape does not extend well to indefinite proof search (what should happen after either 1:1 or the target AWR is reached?), an exponential decay function is used instead, as shown in Figure \ref{fig:decay-and-converge}.
+
+These shapes are further parameterised by an integral \emph{shape frequency} setting, which controls the rate of decay or convergence: every \(n\) steps, the difference between the current and the target AWR is halved, rounding where necessary.
+In future, this might allow the use of repeating patterns such as a sinusoid, hence \emph{frequency}.
+
+Our approach here was restricted by the balancing algorithm used internally, as AWR steps must be discrete and do not take effect immediately.
+An alternative approach might use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducably) taken with the use of a seeded PRNG.
+This would permit use of continuous age/weight functions, but would also introduce the possibility of ``getting unlucky'' in which a clause preferred by the age/weight probability is not selected due to an improbable-but-possible series of PRNG samples.
+
+\begin{figure}
+	\subfloat{\includegraphics[width=0.5\textwidth]{shape-decay}}
+	\subfloat{\includegraphics[width=0.5\textwidth]{shape-converge}}
+	\caption{The new \emph{decay} and \emph{converge} AWR shapes as implemented in Vampire. Different curves exhibit the effect of the AWR shape frequency setting.}
+	\label{fig:decay-and-converge}
+\end{figure}
+
+Together, these new settings allow a number of new option combinations, which can be used in conjunction with Vampire's portfolio \emph{CASC mode} pending integration into the strategy schedules.
 
 \section{Experimental Evaluation}
-TODO: analyse (two sets of) results, discuss drop in performance in exchange for uniques.
+In practice, decaying from an age-biased initial AWR to 1:1 (or converging backwards) was not found to be an improvement.
+However, age/weight shapes starting with a weight-based AWR were found to be useful in some cases.
+An experimental evaluation of this approach on the TPTP~\cite{tptp} benchmark follows.
+
+\todo{analyse (two sets of) results, discuss drop in performance in exchange for uniques}.
 
 \section{Future Work}
-TODO: more shapes, better ways of doing frequency decay, integrate into existing strategy schedules.
+\todo{more shapes, better ways of doing frequency decay, integrate into existing strategy schedules}.
 
 \section{Conclusions}
-TODO: this is a thing, you can do it, you should do it (?), we can probably do better.
+\todo{this is a thing, you can do it, you should do it (?), we can probably do better}.
 

### 2019-02-20 papers/CADE19/paper.tex
added explanation about AWR experiments

@@
 \section{Optimising Age/Weight Ratios}
-\todo{talk about optimising for fixed age/weight}.\\
-\todo{talk about experiments with changing age/weight}.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
+	\caption{
+The number of active clauses reported by Vampire after successful 1-second runs on a TPTP problem.
+A base-2 logarithm is used to produce a continuous scale for the \(x\)-axis.
+For sufficiently negative \(x\), proof search is biased towards age to such an extent that a proof is found via breadth-first search (and hence the number of active clauses remains constant), whereas after \(x\) becomes sufficiently positive, proof search does not terminate for this problem in the given time.
+In between these limits, the function is ``noisy'' and discontinuous.
+These properties are typical for the observed set of problems.
+}
+	\label{fig:example-optimal-awr}
+\end{figure}
+
+\subsection{Fixed AWR}
+Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
+For example, Figure \ref{fig:example-optimal-awr} shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of active clauses means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
+It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
+It also avoids reproducibility issues involved with using system timing approaches.
+}.
+On this problem, a good AWR value is over 400\% better by this metric than the worst AWR value.
+
+This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, and problems which were not solved by the default mode in under a second for any AWR value.
+The whole set yielded similar AWR data for 8,123 problems.
+These data show that while there is no ``best'' AWR for the whole set (Figure \ref{fig:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Figure \ref{fig:optimal-awr-improvement}).
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{no-best-awr}
+	\caption{The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
+No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.}
+	\label{fig:no-best-awr}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{optimal-awr-improvement}
+	\caption{The distribution of the relative performance gain choosing the best AWR value creates, relative to the worst AWR value.}
+	\label{fig:optimal-awr-improvement}
+\end{figure}
+
+\subsection{Variable AWR}
+Although choosing a good AWR value is important, this is covered in part by the use of strategy scheduling~\cite{CADE18} in which many AWR values are tried in sequence (along with other prover options).
+Additionally, a constant AWR fixed for the entire proof search is unlikely to be optimal for any given problem.
+This can be shown by running Vampire with a randomised sequence of age/weight ratios given by a random walk repeatedly, then finding the best after a large number of repetitions.
+Applying this method with 10,000 repetitions to the problem seen earlier (\texttt{PRO017+2}) yields the example AWR trend shown in Figure \ref{fig:random-walk}, which reduces the best number of activations from 330 with a fixed AWR, to 287 with a varying AWR.
+
+In experiments, several AWR ``shapes'' were commonly seen, including:
+\begin{itemize}
+	\item Near-constant, as seen currently.
+	\item A decreasing trend from a higher AWR to a lower AWR.
+	\item An increasing trend from a lower AWR to a higher AWR.
+\end{itemize}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\textwidth]{random-walk}
+	\caption{The AWR series that produced the lowest number of activations on a particular problem. This is a search strategy that a single fixed AWR cannot reproduce.}
+	\label{fig:random-walk}
+\end{figure}
 
@@
 
+\subsection{Decaying AWR}
+\subsection{Converging AWR}
 \todo{analyse (two sets of) results, discuss drop in performance in exchange for uniques}.

### 2019-02-21 papers/CADE19/paper.tex
added results section

@@
 \section{Optimising Age/Weight Ratios}
+\subsection{Fixed AWR}
+Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
+For example, Figure \ref{fig:example-optimal-awr} shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of active clauses means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
+It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
+It also avoids reproducibility issues involved with using system timing approaches.
+}.
+On this problem, a good AWR value is over 400\% better by this metric than the worst AWR value.
+
 \begin{figure}
@@
 \end{figure}
-
-\subsection{Fixed AWR}
-Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
-For example, Figure \ref{fig:example-optimal-awr} shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of active clauses means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
-It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
-It also avoids reproducibility issues involved with using system timing approaches.
-}.
-On this problem, a good AWR value is over 400\% better by this metric than the worst AWR value.
 
@@
 In this work we selected two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
-Since a simple linear shape does not extend well to indefinite proof search (what should happen after either 1:1 or the target AWR is reached?), an exponential decay function is used instead, as shown in Figure \ref{fig:decay-and-converge}.
+Since a simple linear shape does not extend well to indefinite proof search (it is unclear what should happen after either 1:1 or the target AWR is reached), an exponential decay function is used instead, as shown in Figure \ref{fig:decay-and-converge}.
 
@@
 In practice, decaying from an age-biased initial AWR to 1:1 (or converging backwards) was not found to be an improvement.
-However, age/weight shapes starting with a weight-based AWR were found to be useful in some cases.
+However, age/weight shapes decaying from or converging to a weight-based AWR were found to be useful in some cases.
 An experimental evaluation of this approach on the TPTP~\cite{tptp} benchmark follows.
 
-\subsection{Decaying AWR}
-\subsection{Converging AWR}
-\todo{analyse (two sets of) results, discuss drop in performance in exchange for uniques}.
+Vampire first ran to establish baseline performance in its competition \emph{CASC mode} on all problems in TPTP, with a wallclock time limit of 300 seconds.
+For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.
+The baseline solved 13,057 problems in total.
+No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
+
+\subsection{Raw Performance}
+The following shows the performance in terms of solved problems of all the configurations tested.
+These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
+
+\begin{center}
+\begin{tabular}{l l l}
+	Configuration & Frequency & Proved\\
+	\hline
+	baseline & -- & 13057\\
+	\hline
+	converge & 1 & 13039\\
+	converge & 5 & 13029\\
+	converge & 10 & 13028\\
+	converge & 50 & 13015\\
+	converge & 100 & 12976\\
+	converge & 500 & 12895\\
+	converge & 1000 & 12837\\
+	converge & 5000 & 12775\\
+	converge & 10000 & 12751\\
+	\hline
+	decay & 1 & 12698\\
+	decay & 5 & 12702\\
+	decay & 10 & 12698\\
+	decay & 50 & 12712\\
+	decay & 100 & 12726\\
+	decay & 500 & 12795\\
+	decay & 1000 & 12860\\
+	decay & 5000 & 12982\\
+	decay & 10000 & 13002\\
+\end{tabular}
+\end{center}
+
+\subsection{Uniques}
+The following shows the number of problems the new configurations solved that were not solved by the baseline prover.
+This shows the number of new problems that might be able to be solved if the new configurations were added to the competition mode.
+In total, 134 problems were solved by the new configurations that were not solved by the baseline.
+
+\begin{center}
+\begin{tabular}{l l l}
+	Configuration & Frequency & Unique\\
+	\hline
+	converge & 1 & 24\\
+	converge & 5 & 27\\
+	converge & 10 & 35\\
+	converge & 50 & 45\\
+	converge & 100 & 51\\
+	converge & 500 & 63\\
+	converge & 1000 & 52\\
+	converge & 5000 & 53\\
+	converge & 10000 & 53\\
+	\hline
+	decay & 1 & 48\\
+	decay & 5 & 51\\
+	decay & 10 & 48\\
+	decay & 50 & 49\\
+	decay & 100 & 46\\
+	decay & 500 & 29\\
+	decay & 1000 & 29\\
+	decay & 5000 & 16\\
+	decay & 10000 & 7\\
+\end{tabular}
+\end{center}
+
+\subsection{New Problems}
+Some problems were solved which were marked as ``Unknown'' status in the TPTP headers.
+Vampire baseline (and several configurations) were able to solve \texttt{GEO120+1}, but only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.
 

### 2019-02-21 papers/CADE19/paper.tex
convert no-best-awr figure to tabular format, add figure PDFs

@@
 
-This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, and problems which were not solved by the default mode in under a second for any AWR value.
-The whole set yielded similar AWR data for 8,123 problems.
+This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, problems which were not solved by the default mode in under a second for any AWR value, and problems solved in pre-processing.
+The whole set yielded similar AWR data for 7,947 problems.
 These data show that while there is no ``best'' AWR for the whole set (Figure \ref{fig:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Figure \ref{fig:optimal-awr-improvement}).
@@
 	\centering
-	\includegraphics[width=0.8\textwidth]{no-best-awr}
+	\begin{tabular}{l l l l l}
+AWR & Minimum & Maximum & Mean & Standard Deviation\\
+\hline
+8:1 & 1.0 & 150.67 & 1.37 & 1.98\\
+5:1 & 1.0 & 122.22 & 1.33 & 1.64\\
+4:1 & 1.0 & 101.44 & 1.32 & 1.42\\
+3:1 & 1.0 & 105.00 & 1.29 & 1.41\\
+2:1 & 1.0 & 112.67 & 1.27 & 1.45\\
+3:2 & 1.0 & 119.89 & 1.27 & 1.51\\
+5:4 & 1.0 & 125.00 & 1.26 & 1.55\\
+1 & 1.0 & 133.56 & 1.26 & 1.63\\
+2:3 & 1.0 & 147.67 & 1.28 & 1.79\\
+2 & 1.0 & 162.67 & 1.31 & 1.97\\
+3 & 1.0 & 190.56 & 1.37 & 2.30\\
+4 & 1.0 & 17.41 & 1.38 & 0.67\\
+5 & 1.0 & 133.67 & 1.44 & 1.70\\
+6 & 1.0 & 106.44 & 1.47 & 1.46\\
+7 & 1.0 & 104.89 & 1.49 & 1.44\\
+8 & 1.0 & 101.33 & 1.53 & 1.45\\
+10 & 1.0 & 101.78 & 1.60 & 1.53\\
+12 & 1.0 & 101.67 & 1.65 & 1.62\\
+14 & 1.0 & 103.00 & 1.70 & 1.75\\
+16 & 1.0 & 101.33 & 1.74 & 1.76\\
+20 & 1.0 & 100.89 & 1.82 & 1.94\\
+24 & 1.0 & 101.33 & 1.89 & 2.08\\
+28 & 1.0 & 99.22 & 1.96 & 2.24\\
+32 & 1.0 & 100.00 & 1.99 & 2.36\\
+40 & 1.0 & 98.78 & 2.09 & 2.64\\
+50 & 1.0 & 98.78 & 2.17 & 2.88\\
+64 & 1.0 & 101.22 & 2.28 & 3.19\\
+128 & 1.0 & 107.44 & 2.57 & 4.16\\
+1024 & 1.0 & 222.01 & 2.83 & 7.55
+	\end{tabular}
 	\caption{The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
@@
 	\includegraphics[width=0.8\textwidth]{random-walk}
-	\caption{The AWR series that produced the lowest number of activations on a particular problem. This is a search strategy that a single fixed AWR cannot reproduce.}
+	\caption{The AWR series that produced the lowest number of activations on a particular problem, smoothed in order to show the actual effect on proof search.
+This is a search strategy that a single fixed AWR cannot reproduce.}
 	\label{fig:random-walk}

### 2019-02-27 papers/CADE19/paper.tex
intro

@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage{amssymb}
+\usepackage{amsmath}
 \newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
+
 \title{Old or Heavy?\\Decaying Gracefully with Age/Weight Shapes.}
@@
 \institute{University of Manchester, Manchester, UK}
+
+%
+%   Logical notation
+%
+\newcommand{\imply}{\rightarrow}
+%\renewcommand{\implies}{\imply}
+\newcommand{\liff}{\Leftrightarrow}
+\newcommand{\lniff}{\not\Leftrightarrow}
+\newcommand{\iffl}{\leftrightarrow}
+\newcommand{\xor}{\otimes}
+\newcommand{\orl}{\vee}
+\newcommand{\bigorl}{\bigvee}
+\newcommand{\andl}{\wedge}
+\newcommand{\bigandl}{\bigwedge}
+\newcommand{\notl}{\neg}
+\renewcommand{\models}{\vDash}
+\newcommand{\nmodels}{\nvDash}
+\newcommand{\subst}[2]{#1 \mapsto #2}       % for substitutions
+\newcommand{\Subst}[1]{\{#1\}}      % for substitutions
+\newcommand{\emptysubst}{\varepsilon}
+\newcommand{\true}{\mathsf{true}}      % boolean value true
+\newcommand{\false}{\mathsf{false}}     % boolean value false
+\newcommand{\eql}{\simeq}           % equality
+\newcommand{\neql}{\not\simeq}           % equality
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
 \label{sec:introduction}
-\todo{standard Vampire~\cite{vampire} fluff here}.\\
-\todo{explain a generic given-clause algorithm, define clause age and weight --- and how the ``balancing'' algorithm works internally}.\\
+
+At the core of all modern saturation-based theorem provers is \emph{clause selection}.
+It has been known for a long time that the way that clauses are selected in the given clause algorithm is important and early work in 80s already introduce the concept of selecting clauses as a ratio between the age and weight of a clause with the intuition that this balances between breadth and depth first search of the clause space.
+Whilst complex
+
+\cite{DBLP:conf/cade/SchulzM16}
+
+We explore this
+
+After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes the following contributions:
+\begin{itemize}
+	\item An experimental 
+	\item 
+	\item An implementation of this approach in the Vampire~\cite{vampire} theorem prover and an experimental analysis demonstrating its impact. This new method for clause selection led to X
+\end{itemize}
+Conclusions are drawn and future work discussed in Section~\ref{sec:conclusion}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Background} \label{sec:background}
+
+This section introduces the relevant background for the rest of the paper.
+
+%%%%%%
+\paragraph{First-Order Logic.} We consider the standard first-order predicate logic with
+equality. Terms are of the form $f(t_1,\ldots,t_n), c$ or $x$ where
+$f$ is a \emph{function symbol} of arity $n \geq 1$, $t_1,\ldots, t_n$ are terms, $c$ is a zero arity function symbol (i.e. a constant) and $x$ is a variable. 
+%
+The \emph{weight} of a term $t$ is defined as $w(t) =1$ if $t$ is a variable or a constant
+and as $w(t) = 1+\sum_{i=1,\ldots,n} w(t_i)$ if $t$ is of the form
+$f(t_1,\ldots,t_n)$. In other words, the weight or a term is the
+number of symbols in it.
+%
+Atoms are of the form $p(t_1,\ldots,t_n), q$ or $t_1 \eql t_2$ where $p$ is a \emph{predicate symbol} of arity $n$, $t_1, \ldots, t_n$ are terms, $q$ is a zero arity predicate symbol and $\eql$ is the \emph{equality symbol}.
+%
+The weight function naturally extends to atoms: $w(p(t_1,\ldots,t_n)) = 1+\sum_{i=1,\ldots,n} w(t_i)$, $w(t_1 \eql t_2) = w(t_1)+w(t_2)$,
+and $w(q)=1$.
+%
+A literal is either an atom $A$, in which case we call it \emph{positive}, or a negation $\notl A$,
+in which case we call it \emph{negative}.
+%We write negated equalities as $t_1 \neql t_2$. 
+%
+The weight of a literal is the weight of the corresponding atom. 
+% 
+%We write $t[s]_p$ and $L[s]_p$ to denote that a term $s$ occurs in a term $t$ (in a literal $L$) at a position $p$.
+%
+A \emph{clause} is a disjunction of literals $L_1 \orl \ldots \orl L_n$ for $n \geq 0$.
+We disregard the order of literals and treat a clause as a multiset. 
+%
+The weight of a clause is the sum of the weight of its literals.
+
+%%%%%%
+\paragraph{The Given Clause Algorithm.}
+
+Saturation-based theorem provers \emph{saturate} a set of clauses $S$ with respect to an inference system $\mathbb{I}$ e.g.
+they compute a set of clauses $S'$ by applying rules in $\mathbb{I}$ to clauses in $S$ until no new clauses are generated.
+%
+If the empty clause is generated then $S$ is unsatisfiable.
+%
+Calculi such as resolution and superposition have conditions that ensure \emph{completeness}, which means that a saturated set $S$ is satisfiable if it does not contain the empty clause as an element.
+%
+As first-order logic is only semi-decidable, it is not necessarily the case that $S$ has a finite saturation, and even if it does it may be unachievable in practice using the available resources. 
+%
+Therefore, much effort in saturation-based first-order theorem proving involves 
+controlling proof search to make finding the empty clause more likely (within reasonable resource bounds). 
+%
+One important notion is that of \emph{redundancy}, being able to remove clauses from the search space that are not required. 
+%
+Another important notion, the subject of this paper, are literal selections that place restrictions on the inferences that can be performed. 
+%
+Both notions come with additional requirements for completeness.
+
+\todo{explain a generic given-clause algorithm; define clause age}
+
+%%%%%%
+\paragraph{Age-Weight Based Clause Selection.}
+
+\todo{ how the ``balancing'' algorithm works internally}
+
+%%%%%%
+\paragraph{Portfolio Solvers.}
+
 \todo{remark on portfolio modes --- more choice is (sometimes) a good thing}.
 
+%%%%%%
+\paragraph{Related Work.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Optimising Age/Weight Ratios}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Variable AWR for Vampire}
+\label{sec:variable}
+
 If changing AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
@@
 
-\section{Future Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusions and Future Work}
+\label{sec:conclusions}
+
+\todo{this is a thing, you can do it, you should do it (?), we can probably do better}.
+
 \todo{more shapes, better ways of doing frequency decay, integrate into existing strategy schedules}.
-
-\section{Conclusions}
-\todo{this is a thing, you can do it, you should do it (?), we can probably do better}.
 

### 2019-02-27 papers/CADE19/paper.tex
restructure, add comments, add content

@@
 
-%
-%   Logical notation
-%
-\newcommand{\imply}{\rightarrow}
-%\renewcommand{\implies}{\imply}
-\newcommand{\liff}{\Leftrightarrow}
-\newcommand{\lniff}{\not\Leftrightarrow}
-\newcommand{\iffl}{\leftrightarrow}
-\newcommand{\xor}{\otimes}
-\newcommand{\orl}{\vee}
-\newcommand{\bigorl}{\bigvee}
-\newcommand{\andl}{\wedge}
-\newcommand{\bigandl}{\bigwedge}
-\newcommand{\notl}{\neg}
-\renewcommand{\models}{\vDash}
-\newcommand{\nmodels}{\nvDash}
-\newcommand{\subst}[2]{#1 \mapsto #2}       % for substitutions
-\newcommand{\Subst}[1]{\{#1\}}      % for substitutions
-\newcommand{\emptysubst}{\varepsilon}
-\newcommand{\true}{\mathsf{true}}      % boolean value true
-\newcommand{\false}{\mathsf{false}}     % boolean value false
-\newcommand{\eql}{\simeq}           % equality
-\newcommand{\neql}{\not\simeq}           % equality
+\input{def}
+\newcommand{\Balance}{\ensuremath{\mathit{balance}}}
+
+\newcommand{\sandm}{Schulz and M{\"{o}}hrmann}
 
@@
 \begin{abstract}
-Vampire is an automatic theorem prover for first-order logic.
-During proof search in given-clause algorithms, Vampire repeatedly selects clauses from either an \emph{age} or a \emph{weight} queue in a fixed, but configurable \emph{age/weight ratio} (AWR).
+Modern saturation-based theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process.
+This clause selection has a large impact on the performance of proof search and has been the subject of much folklore.
+The standard approach is to alternate between selecting the \emph{oldest} clause and the \emph{lightest} clause with a fixed, but configurable \emph{age/weight ratio} (AWR).
 We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
-Based on these observations we develop several new modes for Vampire which vary AWR according to a ``shape'' during proof search.
+Based on these observations we develop several new modes for the Vampireprover  which vary AWR according to a ``shape'' during proof search.
 The modes solve a number of new problems in the TPTP benchmark.
@@
 
-At the core of all modern saturation-based theorem provers is \emph{clause selection}.
-It has been known for a long time that the way that clauses are selected in the given clause algorithm is important and early work in 80s already introduce the concept of selecting clauses as a ratio between the age and weight of a clause with the intuition that this balances between breadth and depth first search of the clause space.
-Whilst complex
+Currently, the most successful theorem provers (such as Vampire~\cite{}, E~\cite{}, or SPASS~\cite{}) for first-order logic are saturation-based, utilising the well-known \emph{given clause algorithm}.
+Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
+Clearly, the order in which clauses are selected is key to the performance of the algorithm.
+Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{DBLP:conf/cade/SchulzM16} systematically studied these.
+Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or shapes) that prove to be pragmatically useful.
 
-\cite{DBLP:conf/cade/SchulzM16}
+We restrict our attention to clause selection strategies that alternate between selecting clauses based on age (e.g.
+in a first-in-first-out manner) and weight (e.g.
+those with the fewest symbols first).
+It was confirmed by \sandm{} that alternating these two heuristics outperforms either by itself.
+We refer to the ratio with which we use either heuristic as the \emph{age/weight ratio} (AWR) as this is the terminology employed by the Vampire theorem prover, which we use as a vehicle for our study.
 
-We explore this
-
-After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes the following contributions:
-\begin{itemize}
-	\item An experimental 
-	\item 
-	\item An implementation of this approach in the Vampire~\cite{vampire} theorem prover and an experimental analysis demonstrating its impact. This new method for clause selection led to X
-\end{itemize}
-Conclusions are drawn and future work discussed in Section~\ref{sec:conclusion}.
+After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes two main the contributions.
+Firstly, in Section~\ref{sec:awr:study} we experimentally confirm the folklore that (i) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (ii) there is no `best' age/weight ratio, indeed, there exist a large range of pragmatically useful ratios.
+Secondly, we demonstrate (Section~\ref{sec:varying:study}) that varying the age-weight ratio over time can achieve better performance than a fixed ratio and use this to motivate the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
+Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement, proving many new problems unsolvable by any current configuration of Vampire as well as other competitive solvers.
 
@@
 %%%%%%
-\paragraph{First-Order Logic.} We consider the standard first-order predicate logic with
-equality. Terms are of the form $f(t_1,\ldots,t_n), c$ or $x$ where
+\paragraph{First-Order Logic and Weight.} Our setting is the standard first-order predicate logic with equality.
+A formal definition of this logic is not required for this paper but an important notion is that of the \emph{weight} of a clause.
+In first-order logic,  terms are built from function symbols and variables, literals are built from terms, and clauses are disjunctions of literals.
+The weight of a term is the number of function and variable symbols occurring in it and this notion is lifted to clauses.
+
+\iffalse
+Terms are of the form $f(t_1,\ldots,t_n), c$ or $x$ where
 $f$ is a \emph{function symbol} of arity $n \geq 1$, $t_1,\ldots, t_n$ are terms, $c$ is a zero arity function symbol (i.e. a constant) and $x$ is a variable. 
@@
 The weight of a clause is the sum of the weight of its literals. 
+\fi
 
 %%%%%%
-\paragraph{The Given Clause Algorithm.}
+\paragraph{Saturation-Based Proof Search.}
 
@@
 %
-Another important notion, the subject of this paper, are literal selections that place restrictions on the inferences that can be performed. 
+Another important notion are literal selections that place restrictions on the inferences that can be performed. 
 %
 Both notions come with additional requirements for completeness.
-
-\todo{explain a generic given-clause algorithm; define clause age}
+Vampire often gives up these requirements for pragmatic reasons (incomplete strategies have been found to be more efficient than complete ones in certain cases) and in such cases the satisfiability of $S$ upon saturation is \emph{unknown}.
 
 %%%%%%
-\paragraph{Age-Weight Based Clause Selection.}
+\paragraph{The Given Clause Algorithm and AWR Clause Selection.}
+To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, whereas the \Passive\ clauses are those that have not.
+Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop.
+%
+Realisations of the given clause algorithm generally differ in how they organise simplifications.
+There are two main approaches (both implemented by Vampire): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.
 
-\todo{ how the ``balancing'' algorithm works internally}
+The algorithm is centred around the \emph{clause selection} process. As previously mentioned, there are two main heuristics for this:
+\begin{itemize}
+	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause, simulating a \emph{breadth-first} search of the clause space.
+	\item \emph{By Weight (or symbol-counting)} clause selection prefers the youngest clause (with the fewest symbols).
+The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are (in some sense) closer to this.
+Furthermore, lighter clauses are friendlier (less explosive) to proof search in various ways e.g.
+they are more general in terms of subsumption and tend to have fewer children.
+\end{itemize}
+It has been shown (see \sandm{}) that alternating these heuristics is beneficial and we consider this setting here.
+In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm.
+The balance is initialised to 0 and used as follows: a positive balance means that a clause should be selected by age, whereas a negative balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
+
+Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.
+
+\begin{figure}[t]
+\begin{tabbing}
+~~~~~~\=~~~~\=~~~~\=~~~\=~~~~\=~~~~\=~~~~\=~~~~\=~~~~\=\kill
+\>\INPUT: $\Init$: set of clauses\semicol, $a:w$ age-weight ratio \\
+\>\VAR\ $\Active$, $\Passive$, $\Unprocessed$: set of clauses\semicol \\
+\>\VAR\ $\Given$, $\New$: clause\semicol\\
+\>$\Active\ \assign\ \emptyset$\semicol \\
+\>$\Unprocessed\ \assign\ \Init$\semicol \\
+\Mark
+\>$\Balance\ \assign\ 0$\semicol\\
+\>\LOOP\\ 
+\>\>\WHILE\ $\Unprocessed \neq \emptyset$ \\
+\>\>\>$\New \assign \Pop(\Unprocessed)$\semicol \\ 
+\>\>\>\IF\ $\New = \emptyclause$ \THEN\ \RETURN\ \textit{unsatisfiable}\semicol \\
+\>\>\>\IF\ $\Retained(\New)$ \THEN \` (* retention test *) \\ %\inc
+\>\>\>\>simplify $\New$ by clauses in $\Active \cup \Passive$ \semicol 
+                        \` (* forward simplification *) \\
+\>\>\>\>\IF\ $\New = \emptyclause$ \THEN\ \RETURN\ \textit{unsatisfiable}\semicol \\
+\>\>\>\>\IF\ $\Retained(\New)$ \THEN
+                        \` (* another retention test *) \\
+\>\>\>\>\>simplify $\Active$ and  $\Passive$ using $\New$ \semicol
+                        \` (* backward simplification *) \\
+\>\>\>\>\>move the simplified clauses to $\Unprocessed$\semicol \\
+\>\>\>\>\>add $\New$ to $\Passive$\\ %\dec 
+\>\>\IF\ $\Passive = \emptyset$ \THEN\
+          \RETURN\ \textit{satisfiable} or \textit{unknown}\\
+\Mark
+\>\>\IF\ $\Balance > 0$ \THEN\\
+\Mark
+\>\>\>\>$\Given\ \assign\ $ lightest clause in \Passive\semicol\\
+\Mark
+\>\>\>\>$\Balance \assign\ \Balance - w$\semicol\\
+\Mark
+\>\>\ELSE\\
+\Mark
+\>\>\>\>$\Given\ \assign\ $ oldest clause in \Passive\semicol\\
+\Mark
+\>\>\>\>$\Balance \assign\ \Balance + a$\semicol\\
+\>\>move $\Given$ from $\Passive$ to $\Active$\semicol \\
+\>\>$\Unprocessed \assign \Infer(\Given,\Active)$\semicol \` (* generating inferences *)\\
+\end{tabbing}
+
+                \caption{The Discount Saturation Algorithm with AWR clause selection
+	        \label{fig:discount}}
+
+\end{figure}
 
@@
 
-\todo{remark on portfolio modes --- more choice is (sometimes) a good thing}.
+Vampire is a portfolio solver and is typically run in a mode that attempts multiple different \emph{strategies} in quick succession e.g.
+in a 30 second run it may attempt 10 or more different strategies.
+These strategies employ many different options including different saturation algorithms (including Otter and Discount), preprocessing options, literal selection strategies, inference rules, and clause selection heuristics.
+The portfolio mode is a significant improvement on any single strategy.
+
+We note here that Vampire's portfolio mode includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of nongoal clauses, thus preferring clauses in or derived from the goal in clause selection.
+Use of this heuristic is orthogonal to the age/weight ratio and we do not investigate its impact any further here.
 
 %%%%%%
-\paragraph{Related Work.}
+\paragraph{Related Work.} \todo{Mention clause selection strategies of Otter, Prover9, E, and SPASS}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Optimising Age/Weight Ratios}
-\subsection{Fixed AWR}
-Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
-For example, Figure \ref{fig:example-optimal-awr} shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of active clauses means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
+\newpage
+\section{Optimising Age/Weight Ratios} \label{sec:awr:study}
+
+We set out to experimentally confirm two assumptions from folklore:
+\begin{enumerate}
+	\item The choice of age/weight ratio often has a significant effect on the performance of proof search
+	\item There is no generally best age/weight ratio, even for a syntactically well-defined subset of problems
+\end{enumerate}
+These were supported by the work of \sandm{} but here we focus on these points and explore them more deeply.
+
+%Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
+As an initial illustrative example of how varying the AWR effects the number of clauses required to be processed before a proof is found consider Figure \ref{fig:example-optimal-awr}.
+This shows \todo{FINISH}
+
+  shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of active clauses means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
 It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
@@
 
-\begin{figure}
+\begin{figure}[t]
 	\centering
@@
 In between these limits, the function is ``noisy'' and discontinuous.
-These properties are typical for the observed set of problems.
+These properties are typical for the observed set of problems. \todo{Move definition out of caption and use more generally.}
 }
@@
 
-\begin{figure}
+\noindent
+ \todo{Get rid of Min column in Fig 3. Can we also capture the solved/unsolved balance? Perhaps just say what \% of solved problems were solved using that ratio.}
+\todo{This discussion should be expanded. Results need more explanation and we should summarise e.g. yes we confirmed assumptions.}\\
+\todo{Can we extract some general information on correlation (or lack thereof) between best AWRs and syntactic features?}
+
+\begin{figure}[t]
 	\centering
@@
 
-\subsection{Variable AWR}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Variable AWR for Vampire}
+
+In this section we motivate and define a variable clause selection approach.
+
+%%%%%%%
+\subsection{The Optimal AWR Over Time} \label{sec:varying:study}
+
 Although choosing a good AWR value is important, this is covered in part by the use of strategy scheduling~\cite{CADE18} in which many AWR values are tried in sequence (along with other prover options).
-Additionally, a constant AWR fixed for the entire proof search is unlikely to be optimal for any given problem.
+Additionally, given that varying the AWR can have such a large impact, it seems likely that a constant AWR fixed for the entire proof search is unlikely to be optimal for any given problem.
 This can be shown by running Vampire with a randomised sequence of age/weight ratios given by a random walk repeatedly, then finding the best after a large number of repetitions.
 Applying this method with 10,000 repetitions to the problem seen earlier (\texttt{PRO017+2}) yields the example AWR trend shown in Figure \ref{fig:random-walk}, which reduces the best number of activations from 330 with a fixed AWR, to 287 with a varying AWR.
+
+\todo{A little more explanation of the Fig... eg what is AWR trend?}
 
@@
 \end{itemize}
+\todo{Need to be more specific about `commonly seen'}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Variable AWR for Vampire}
-\label{sec:variable}
+%%%%%%%
+\subsection{Varying AWR in Vampire} \label{sec:varying:implementation}
 
@@
 
+\todo{What are these figures showing?}
+
 These shapes are further parameterised by an integral \emph{shape frequency} setting, which controls the rate of decay or convergence: every \(n\) steps, the difference between the current and the target AWR is halved, rounding where necessary.
 In future, this might allow the use of repeating patterns such as a sinusoid, hence \emph{frequency}.
+
+\todo{Given that we don't, perhaps we should change the parameter to \emph{rate} or something else?}
 
@@
 
-\section{Experimental Evaluation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Experimental Evaluation} \label{sec:experiments}
+
+\todo{Change marketing here. Success is adding new options to portfolio mode. Only mention CASC once and refer to portfolio mode elsewhere.}
+
 In practice, decaying from an age-biased initial AWR to 1:1 (or converging backwards) was not found to be an improvement.

### 2019-02-27 papers/CADE19/paper.tex
some feedback implemented, meeting

@@
 \usepackage{amsmath}
-\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
+\usepackage{todonotes}
+%\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
 
@@
 \newcommand{\Balance}{\ensuremath{\mathit{balance}}}
-
 \newcommand{\sandm}{Schulz and M{\"{o}}hrmann}
@@
 We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
-Based on these observations we develop several new modes for the Vampireprover  which vary AWR according to a ``shape'' during proof search.
+Based on these observations we develop several new modes for the Vampire prover  which vary AWR according to a ``shape'' during proof search.
 The modes solve a number of new problems in the TPTP benchmark.
@@
 
-Currently, the most successful theorem provers (such as Vampire~\cite{}, E~\cite{}, or SPASS~\cite{}) for first-order logic are saturation-based, utilising the well-known \emph{given clause algorithm}.
+Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{}, or SPASS~\cite{}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
 Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
 Clearly, the order in which clauses are selected is key to the performance of the algorithm.
-Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{DBLP:conf/cade/SchulzM16} systematically studied these.
+Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
 Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or shapes) that prove to be pragmatically useful.
@@
 
+\subsection{Logarithmic AWR}
+Optimising AWR values is more easily achieved if they have a continuous scale.
+AWR values are mathematically \(\mathbb{Q}^{+}\), the positive rational numbers, but in practice are more easily seen logarithmically: 1:16 is the same distance in terms of the effect on proof search from 1:8 as it is from 1:32.
+Therefore, we introduce the \emph{logarithmic} AWR \(L\), defined in terms of age \(A\) and weight \(W\) as
+\[
+	L = \log_2{\left(\frac{A}{W}\right)}
+\]
+With this definition, AWR can now be seen continuously.
+As \(L\) tends to positive infinity, Vampire selects only by weight, whereas if \(L\) tends to negative infinity Vampire selects only by age.
+\(L = 0\) represents the middle ground of a 1:1 age/weight ratio.
+
+\subsection{Experiments}
 %Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
@@
 The number of active clauses reported by Vampire after successful 1-second runs on a TPTP problem.
-A base-2 logarithm is used to produce a continuous scale for the \(x\)-axis.
 For sufficiently negative \(x\), proof search is biased towards age to such an extent that a proof is found via breadth-first search (and hence the number of active clauses remains constant), whereas after \(x\) becomes sufficiently positive, proof search does not terminate for this problem in the given time.
 In between these limits, the function is ``noisy'' and discontinuous.
-These properties are typical for the observed set of problems. \todo{Move definition out of caption and use more generally.}
+These properties are typical for the observed set of problems.
 }
@@
 \noindent
- \todo{Get rid of Min column in Fig 3. Can we also capture the solved/unsolved balance? Perhaps just say what \% of solved problems were solved using that ratio.}
+ \todo{Can we also capture the solved/unsolved balance? Perhaps just say what \% of solved problems were solved using that ratio.}
 \todo{This discussion should be expanded. Results need more explanation and we should summarise e.g. yes we confirmed assumptions.}\\
@@
 	\begin{tabular}{l l l l l}
-AWR & Minimum & Maximum & Mean & Standard Deviation\\
+AWR & Maximum & Mean & Standard Deviation\\
 \hline
-8:1 & 1.0 & 150.67 & 1.37 & 1.98\\
-5:1 & 1.0 & 122.22 & 1.33 & 1.64\\
-4:1 & 1.0 & 101.44 & 1.32 & 1.42\\
-3:1 & 1.0 & 105.00 & 1.29 & 1.41\\
-2:1 & 1.0 & 112.67 & 1.27 & 1.45\\
-3:2 & 1.0 & 119.89 & 1.27 & 1.51\\
-5:4 & 1.0 & 125.00 & 1.26 & 1.55\\
-1 & 1.0 & 133.56 & 1.26 & 1.63\\
-2:3 & 1.0 & 147.67 & 1.28 & 1.79\\
-2 & 1.0 & 162.67 & 1.31 & 1.97\\
-3 & 1.0 & 190.56 & 1.37 & 2.30\\
-4 & 1.0 & 17.41 & 1.38 & 0.67\\
-5 & 1.0 & 133.67 & 1.44 & 1.70\\
-6 & 1.0 & 106.44 & 1.47 & 1.46\\
-7 & 1.0 & 104.89 & 1.49 & 1.44\\
-8 & 1.0 & 101.33 & 1.53 & 1.45\\
-10 & 1.0 & 101.78 & 1.60 & 1.53\\
-12 & 1.0 & 101.67 & 1.65 & 1.62\\
-14 & 1.0 & 103.00 & 1.70 & 1.75\\
-16 & 1.0 & 101.33 & 1.74 & 1.76\\
-20 & 1.0 & 100.89 & 1.82 & 1.94\\
-24 & 1.0 & 101.33 & 1.89 & 2.08\\
-28 & 1.0 & 99.22 & 1.96 & 2.24\\
-32 & 1.0 & 100.00 & 1.99 & 2.36\\
-40 & 1.0 & 98.78 & 2.09 & 2.64\\
-50 & 1.0 & 98.78 & 2.17 & 2.88\\
-64 & 1.0 & 101.22 & 2.28 & 3.19\\
-128 & 1.0 & 107.44 & 2.57 & 4.16\\
-1024 & 1.0 & 222.01 & 2.83 & 7.55
+8:1 & 150.67 & 1.37 & 1.98\\
+5:1 & 122.22 & 1.33 & 1.64\\
+4:1 & 101.44 & 1.32 & 1.42\\
+3:1 & 105.00 & 1.29 & 1.41\\
+2:1 & 112.67 & 1.27 & 1.45\\
+3:2 & 119.89 & 1.27 & 1.51\\
+5:4 & 125.00 & 1.26 & 1.55\\
+1 & 133.56 & 1.26 & 1.63\\
+2:3 & 147.67 & 1.28 & 1.79\\
+2 & 162.67 & 1.31 & 1.97\\
+3 & 190.56 & 1.37 & 2.30\\
+4 & 17.41 & 1.38 & 0.67\\
+5 & 133.67 & 1.44 & 1.70\\
+6 & 106.44 & 1.47 & 1.46\\
+7 & 104.89 & 1.49 & 1.44\\
+8 & 101.33 & 1.53 & 1.45\\
+10 & 101.78 & 1.60 & 1.53\\
+12 & 101.67 & 1.65 & 1.62\\
+14 & 103.00 & 1.70 & 1.75\\
+16 & 101.33 & 1.74 & 1.76\\
+20 & 100.89 & 1.82 & 1.94\\
+24 & 101.33 & 1.89 & 2.08\\
+28 & 99.22 & 1.96 & 2.24\\
+32 & 100.00 & 1.99 & 2.36\\
+40 & 98.78 & 2.09 & 2.64\\
+50 & 98.78 & 2.17 & 2.88\\
+64 & 101.22 & 2.28 & 3.19\\
+128 & 107.44 & 2.57 & 4.16\\
+1024 & 222.01 & 2.83 & 7.55
 	\end{tabular}
@@
 Our approach here was restricted by the balancing algorithm used internally, as AWR steps must be discrete and do not take effect immediately.
-An alternative approach might use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducably) taken with the use of a seeded PRNG.
+An alternative approach might use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducibly) taken with the use of a seeded PRNG.
 This would permit use of continuous age/weight functions, but would also introduce the possibility of ``getting unlucky'' in which a clause preferred by the age/weight probability is not selected due to an improbable-but-possible series of PRNG samples.
@@
 
-\subsection{Raw Performance}
+\subsection{Benchmark Performance}
 The following shows the performance in terms of solved problems of all the configurations tested.
+``Proved'' refers to the total number of problems a configuration solved, whereas ``Uniques'' is the number of problems a configuration solved which were not solved by the baseline.
 These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
-
-\begin{center}
-\begin{tabular}{l l l}
-	Configuration & Frequency & Proved\\
-	\hline
-	baseline & -- & 13057\\
-	\hline
-	converge & 1 & 13039\\
-	converge & 5 & 13029\\
-	converge & 10 & 13028\\
-	converge & 50 & 13015\\
-	converge & 100 & 12976\\
-	converge & 500 & 12895\\
-	converge & 1000 & 12837\\
-	converge & 5000 & 12775\\
-	converge & 10000 & 12751\\
-	\hline
-	decay & 1 & 12698\\
-	decay & 5 & 12702\\
-	decay & 10 & 12698\\
-	decay & 50 & 12712\\
-	decay & 100 & 12726\\
-	decay & 500 & 12795\\
-	decay & 1000 & 12860\\
-	decay & 5000 & 12982\\
-	decay & 10000 & 13002\\
-\end{tabular}
-\end{center}
-
-\subsection{Uniques}
-The following shows the number of problems the new configurations solved that were not solved by the baseline prover.
-This shows the number of new problems that might be able to be solved if the new configurations were added to the competition mode.
 In total, 134 problems were solved by the new configurations that were not solved by the baseline.
@@
 \begin{center}
-\begin{tabular}{l l l}
-	Configuration & Frequency & Unique\\
+\begin{tabular}{l l l l}
+	Configuration & Frequency & Proved & Uniques\\
 	\hline
-	converge & 1 & 24\\
-	converge & 5 & 27\\
-	converge & 10 & 35\\
-	converge & 50 & 45\\
-	converge & 100 & 51\\
-	converge & 500 & 63\\
-	converge & 1000 & 52\\
-	converge & 5000 & 53\\
-	converge & 10000 & 53\\
-	\hline
-	decay & 1 & 48\\
-	decay & 5 & 51\\
-	decay & 10 & 48\\
-	decay & 50 & 49\\
-	decay & 100 & 46\\
-	decay & 500 & 29\\
-	decay & 1000 & 29\\
-	decay & 5000 & 16\\
-	decay & 10000 & 7\\
+	baseline & -- & 13057 & 0\\
+	converge & 1 & 13039 & 24\\
+	converge & 5 & 13029 & 27\\
+	converge & 10 & 13028 & 35\\
+	converge & 50 & 13015 & 45\\
+	converge & 100 & 12976 & 51\\
+	converge & 500 & 12895 & 63\\
+	converge & 1000 & 12837 & 52\\
+	converge & 5000 & 12775 & 53\\
+	converge & 10000 & 12751 & 53\\
+	decay & 1 & 12698 & 48\\
+	decay & 5 & 12702 & 51\\
+	decay & 10 & 12698 & 48\\
+	decay & 50 & 12712 & 49\\
+	decay & 100 & 12726 & 46\\
+	decay & 500 & 12795 & 29\\
+	decay & 1000 & 12860 & 29\\
+	decay & 5000 & 12982 & 16\\
+	decay & 10000 & 13002 & 7\\
 \end{tabular}
@@
 
-\todo{this is a thing, you can do it, you should do it (?), we can probably do better}.
-
-\todo{more shapes, better ways of doing frequency decay, integrate into existing strategy schedules}.
+\todo[inline]{this is a thing, you can do it, you should do it (?), we can probably do better}
+\todo[inline]{more shapes, better ways of doing frequency decay, integrate into existing strategy schedules}
 

### 2019-02-27 papers/CADE19/paper.tex
fix results section

@@
 \begin{enumerate}
-	\item The choice of age/weight ratio often has a significant effect on the performance of proof search
-	\item There is no generally best age/weight ratio, even for a syntactically well-defined subset of problems
+	\item The choice of age/weight ratio often has a significant effect on the performance of proof search.
+	\item There is in general no single best age/weight ratio for a given set of problems.
 \end{enumerate}
@@
 The number of active clauses reported by Vampire after successful 1-second runs on a TPTP problem.
-For sufficiently negative \(x\), proof search is biased towards age to such an extent that a proof is found via breadth-first search (and hence the number of active clauses remains constant), whereas after \(x\) becomes sufficiently positive, proof search does not terminate for this problem in the given time.
-In between these limits, the function is ``noisy'' and discontinuous.
+In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but does follow an approximate trend and settles toward the global optimum.
 These properties are typical for the observed set of problems.
@@
 	\caption{The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
-No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.}
+No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
+A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
+\sandm~found that 1:5 had a similar property, but this might be explained by differences in test environment.}
 	\label{fig:no-best-awr}
@@
 
-In experiments, several AWR ``shapes'' were commonly seen, including:
-\begin{itemize}
-	\item Near-constant, as seen currently.
-	\item A decreasing trend from a higher AWR to a lower AWR.
-	\item An increasing trend from a lower AWR to a higher AWR.
-\end{itemize}
-\todo{Need to be more specific about `commonly seen'}
+Unsurprisingly, in ad-hoc experiments on other problems, the best shape is rarely constant.
+This suggests that implementing other shapes, such as an increasing or decreasing trend, might lead to quicker proofs in the Vampire theorem prover.
 
@@
 Together, these new settings allow a number of new option combinations, which can be used in conjunction with Vampire's portfolio \emph{CASC mode} pending integration into the strategy schedules.
+This version of the prover is currently in a separate branch in the main Vampire repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Experimental Evaluation} \label{sec:experiments}
+\section{Experimental Evaluation}
+\label{sec:experiments}
 
@@
 \subsection{Benchmark Performance}
-The following shows the performance in terms of solved problems of all the configurations tested.
-``Proved'' refers to the total number of problems a configuration solved, whereas ``Uniques'' is the number of problems a configuration solved which were not solved by the baseline.
+Figure \ref{fig:results} shows the performance in terms of solved problems of all the configurations tested.
 These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
@@
 
-\begin{center}
-\begin{tabular}{l l l l}
-	Configuration & Frequency & Proved & Uniques\\
+\begin{figure}
+\centering
+\begin{tabular}{l l l l l l}
+	Configuration & Frequency & Proved & Fresh & Uniques & \(u\)-score\\
 	\hline
-	baseline & -- & 13057 & 0\\
-	converge & 1 & 13039 & 24\\
-	converge & 5 & 13029 & 27\\
-	converge & 10 & 13028 & 35\\
-	converge & 50 & 13015 & 45\\
-	converge & 100 & 12976 & 51\\
-	converge & 500 & 12895 & 63\\
-	converge & 1000 & 12837 & 52\\
-	converge & 5000 & 12775 & 53\\
-	converge & 10000 & 12751 & 53\\
-	decay & 1 & 12698 & 48\\
-	decay & 5 & 12702 & 51\\
-	decay & 10 & 12698 & 48\\
-	decay & 50 & 12712 & 49\\
-	decay & 100 & 12726 & 46\\
-	decay & 500 & 12795 & 29\\
-	decay & 1000 & 12860 & 29\\
-	decay & 5000 & 12982 & 16\\
-	decay & 10000 & 13002 & 7\\
+	baseline & -- & 13057 & 0 & 1 & 714.2\\
+	converge & 1 & 13039 & 24 & 3 & 714.3\\
+	converge & 5 & 13029 & 27 & 1 & 709.5\\
+	converge & 10 & 13028 & 35 & 5 & 714.3\\
+	converge & 50 & 13015 & 45 & 5 & 712.8\\
+	converge & 100 & 12976 & 51 & 1 & 705.9\\
+	converge & 500 & 12895 & 63 & 4 & 698.3\\
+	converge & 1000 & 12837 & 52 & 0 & 688.6\\
+	converge & 5000 & 12775 & 53 & 1 & 682.4\\
+	converge & 10000 & 12751 & 53 & 0 & 678.7\\
+	decay & 1 & 12698 & 48 & 1 & 673.6\\
+	decay & 5 & 12702 & 51 & 1 & 674.9\\
+	decay & 10 & 12698 & 48 & 1 & 674.2\\
+	decay & 50 & 12712 & 49 & 2 & 679.1\\
+	decay & 100 & 12726 & 46 & 1 & 678.8\\
+	decay & 500 & 12795 & 29 & 1 & 685.5\\
+	decay & 1000 & 12860 & 29 & 2 & 692.6\\
+	decay & 5000 & 12982 & 16 & 2 & 707.1\\
+	decay & 10000 & 13002 & 7 & 0 & 706.3\\
 \end{tabular}
-\end{center}
+\caption{
+Results for the tested configurations.
+\emph{Proved} refers to the total number of problems a configuration solved.
+\emph{Fresh} is the number of problems a configuration solved which were not solved by the baseline.
+\emph{Uniques} is the number of problems a configuration solved which were not solved by any other configuration.
+\emph{\(u\)-score} is a refined unique score which correlates to a configuration's utility in solving new problems, as used in Hoder \emph{et al.}~\cite{u-score}.
+}
+\label{fig:results}
+\end{figure}
 

### 2019-02-28 papers/CADE19/paper.tex
refs

@@
 
-Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{}, or SPASS~\cite{}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
+Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, or SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
 Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.

### 2019-02-28 papers/CADE19/paper.tex
changes

@@
 \begin{itemize}
-	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause, simulating a \emph{breadth-first} search of the clause space.
+	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause (produced earlier in proof search), simulating a \emph{breadth-first} search of the clause space.
+In Vampire the age of a clause is the number of inferences performed to produce it (e.g.
+input clauses have age 0).
 	\item \emph{By Weight (or symbol-counting)} clause selection prefers the youngest clause (with the fewest symbols).
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
+\newpage~\newpage
 \section{Optimising Age/Weight Ratios} \label{sec:awr:study}
@@
 
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
+	\caption{
+The number of active clauses reported by Vampire after successful 1-second runs on a TPTP problem.
+In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but follows an approximate trend and settles toward the global optimum.
+This is typical for the observed set of problems.
+}
+	\label{fig:example-optimal-awr}
+\end{figure}
+
 \subsection{Experiments}
@@
 
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
-	\caption{
-The number of active clauses reported by Vampire after successful 1-second runs on a TPTP problem.
-In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but does follow an approximate trend and settles toward the global optimum.
-These properties are typical for the observed set of problems.
-}
-	\label{fig:example-optimal-awr}
-\end{figure}
-
 This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, problems which were not solved by the default mode in under a second for any AWR value, and problems solved in pre-processing.
 The whole set yielded similar AWR data for 7,947 problems.
-These data show that while there is no ``best'' AWR for the whole set (Figure \ref{fig:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Figure \ref{fig:optimal-awr-improvement}).
+These data show that while there is no ``best'' AWR for the whole set (Table \ref{tab:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Figure \ref{fig:optimal-awr-improvement}).
 
@@
 
-\begin{figure}[t]
+\begin{table}[t]
+	\caption{The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
+No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
+A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
+\sandm~found that 1:5 had a similar property, but this might be explained by differences in test environment.}
 	\centering
-	\begin{tabular}{l l l l l}
-AWR & Maximum & Mean & Standard Deviation\\
+	\begin{tabular}{l | c | c  c }
+AWR & Maximum Drop & Mean Drop & (Standard Deviation)\\
 \hline
@@
 	\end{tabular}
-	\caption{The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
-No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
-A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
-\sandm~found that 1:5 had a similar property, but this might be explained by differences in test environment.}
-	\label{fig:no-best-awr}
-\end{figure}
+
+	\label{tab:no-best-awr}
+\end{table}
 
@@
 
+\clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-02-28 papers/CADE19/paper.tex
fix some todos, add combined converge/decay results

@@
 \usepackage{todonotes}
-%\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background} \label{sec:background}
+\section{Background}
+\label{sec:background}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage~\newpage
 \section{Optimising Age/Weight Ratios} \label{sec:awr:study}
@@
 	\caption{
-The number of active clauses reported by Vampire after successful 1-second runs on a TPTP problem.
+The number of activations reported by Vampire after successful 1-second runs on a TPTP problem.
 In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but follows an approximate trend and settles toward the global optimum.
@@
 \subsection{Experiments}
-%Changing Vampire's fixed AWR has a significant effect on the number of clauses required to be processed before a proof is found.
 As an initial illustrative example of how varying the AWR effects the number of clauses required to be processed before a proof is found consider Figure \ref{fig:example-optimal-awr}.
@@
 
-  shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of active clauses means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
+  shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of activations means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
 It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
@@
 
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
+	\caption{
+The number of activations reported by Vampire after successful 1-second runs on a TPTP problem.
+In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but does follow an approximate trend and settles toward the global optimum.
+These properties are typical for the observed set of problems.
+}
+	\label{fig:example-optimal-awr}
+\end{figure}
+
 This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, problems which were not solved by the default mode in under a second for any AWR value, and problems solved in pre-processing.
 The whole set yielded similar AWR data for 7,947 problems.
-These data show that while there is no ``best'' AWR for the whole set (Table \ref{tab:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Figure \ref{fig:optimal-awr-improvement}).
+These data show that while there is no ``best'' AWR for the whole set (Table \ref{tab:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Table \ref{tab:awr-improvement}).
 
@@
 \todo{This discussion should be expanded. Results need more explanation and we should summarise e.g. yes we confirmed assumptions.}\\
-\todo{Can we extract some general information on correlation (or lack thereof) between best AWRs and syntactic features?}
 
@@
 
-\begin{figure}
+\begin{table}
+	\caption{Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
+We compare the base settings of 1:1 (Vampire's default), 1:5 (the best-behaved from Table \ref{tab:no-best-awr}), and the worst setting for the problem.}
 	\centering
-	\includegraphics[width=0.8\textwidth]{optimal-awr-improvement}
-	\caption{The distribution of the relative performance gain choosing the best AWR value creates, relative to the worst AWR value.}
-	\label{fig:optimal-awr-improvement}
-\end{figure}
+	\begin{tabular}{c | c | c c}
+		& Maximum Gain & Mean Gain & (Standard Deviation)\\
+		\hline
+		1:1 & 133.56 & 1.26 & 1.63 \\
+		1:5 & 133.67 & 1.44 & 1.70 \\
+		(worst) & 222.01 & 3.95 & 7.60 \\
+	\end{tabular}
+	\label{tab:awr-improvement}
+\end{table}
 
-\clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\todo{A little more explanation of the Fig... eg what is AWR trend?}
-
 Unsurprisingly, in ad-hoc experiments on other problems, the best shape is rarely constant.
@@
 
-\todo{Given that we don't, perhaps we should change the parameter to \emph{rate} or something else?}
-
 Our approach here was restricted by the balancing algorithm used internally, as AWR steps must be discrete and do not take effect immediately.
@@
 
-Together, these new settings allow a number of new option combinations, which can be used in conjunction with Vampire's portfolio \emph{CASC mode} pending integration into the strategy schedules.
-This version of the prover is currently in a separate branch in the main Vampire repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.
+Together, these new settings allow a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.
+This version of the prover is currently in a separate branch in the main Vampire source repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.
 
@@
 \subsection{Benchmark Performance}
-Figure \ref{fig:results} shows the performance in terms of solved problems of all the configurations tested.
+Figure \ref{tab:results} shows the performance in terms of solved problems of all the configurations tested.
 These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
@@
 
-\begin{figure}
+\begin{table}
+	\caption{
+Results for the tested configurations.
+\emph{Proved} refers to the total number of problems a configuration solved.
+\emph{Fresh} is the number of problems a configuration solved which were not solved by the baseline.
+\emph{Uniques} is the number of problems a configuration solved which were not solved by any other configuration.
+\emph{\(u\)-score} is a refined unique score which correlates to a configuration's utility in solving new problems, as used in Hoder \emph{et al.}~\cite{u-score}.
+	}
 	\centering
@@
 		decay & 10000 & 13002 & 7 & 0 & 706.3\\
+		\hline
+		converge & (combined) & 13167 & 117 & -- & --\\
+		decay & (combined) & 13106 & 93 & -- & --\\
 	\end{tabular}
-\caption{
-Results for the tested configurations.
-\emph{Proved} refers to the total number of problems a configuration solved.
-\emph{Fresh} is the number of problems a configuration solved which were not solved by the baseline.
-\emph{Uniques} is the number of problems a configuration solved which were not solved by any other configuration.
-\emph{\(u\)-score} is a refined unique score which correlates to a configuration's utility in solving new problems, as used in Hoder \emph{et al.}~\cite{u-score}.
-}
-\label{fig:results}
-\end{figure}
+	\label{tab:results}
+\end{table}
 

### 2019-02-28 papers/CADE19/paper.tex
more fixes

@@
 
-\noindent
- \todo{Can we also capture the solved/unsolved balance? Perhaps just say what \% of solved problems were solved using that ratio.}
-\todo{This discussion should be expanded. Results need more explanation and we should summarise e.g. yes we confirmed assumptions.}\\
+\todo{This discussion should be expanded. Results need more explanation and we should summarise e.g. yes we confirmed assumptions.}
 
 \begin{table}[t]
-	\caption{The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
+	\caption{
+The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
+The percentage of problems in this set solved by the AWR setting is also shown.
 No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
 A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
-\sandm~found that 1:5 had a similar property, but this might be explained by differences in test environment.}
+\sandm~found that 1:5 had a similar property, but this might be explained by differences in test environment.
+}
 	\centering
-	\begin{tabular}{l | c | c  c }
-AWR & Maximum Drop & Mean Drop & (Standard Deviation)\\
+	\begin{tabular}{l | c | c | c  c }
+		AWR & \% Solved & Maximum Drop & Mean Drop & (Standard Deviation)\\
 \hline
-8:1 & 150.67 & 1.37 & 1.98\\
-5:1 & 122.22 & 1.33 & 1.64\\
-4:1 & 101.44 & 1.32 & 1.42\\
-3:1 & 105.00 & 1.29 & 1.41\\
-2:1 & 112.67 & 1.27 & 1.45\\
-3:2 & 119.89 & 1.27 & 1.51\\
-5:4 & 125.00 & 1.26 & 1.55\\
-1 & 133.56 & 1.26 & 1.63\\
-2:3 & 147.67 & 1.28 & 1.79\\
-2 & 162.67 & 1.31 & 1.97\\
-3 & 190.56 & 1.37 & 2.30\\
-4 & 17.41 & 1.38 & 0.67\\
-5 & 133.67 & 1.44 & 1.70\\
-6 & 106.44 & 1.47 & 1.46\\
-7 & 104.89 & 1.49 & 1.44\\
-8 & 101.33 & 1.53 & 1.45\\
-10 & 101.78 & 1.60 & 1.53\\
-12 & 101.67 & 1.65 & 1.62\\
-14 & 103.00 & 1.70 & 1.75\\
-16 & 101.33 & 1.74 & 1.76\\
-20 & 100.89 & 1.82 & 1.94\\
-24 & 101.33 & 1.89 & 2.08\\
-28 & 99.22 & 1.96 & 2.24\\
-32 & 100.00 & 1.99 & 2.36\\
-40 & 98.78 & 2.09 & 2.64\\
-50 & 98.78 & 2.17 & 2.88\\
-64 & 101.22 & 2.28 & 3.19\\
-128 & 107.44 & 2.57 & 4.16\\
-1024 & 222.01 & 2.83 & 7.55
+8:1 & 85.25 & 150.67 & 1.37 & 1.98\\
+5:1 & 86.10 & 122.22 & 1.33 & 1.64\\
+4:1 & 86.93 & 101.44 & 1.32 & 1.42\\
+3:1 & 87.63 & 105.00 & 1.29 & 1.41\\
+2:1 & 88.62 & 112.67 & 1.27 & 1.45\\
+3:2 & 89.83 & 119.89 & 1.27 & 1.51\\
+5:4 & 89.98 & 125.00 & 1.26 & 1.55\\
+1   & 90.56 & 133.56 & 1.26 & 1.63\\
+2:3 & 91.20 & 147.67 & 1.28 & 1.79\\
+2   & 91.68 & 162.67 & 1.31 & 1.97\\
+3   & 91.81 & 190.56 & 1.37 & 2.30\\
+4   & 91.85 & 17.41 & 1.38 & 0.67\\
+5   & 92.00 & 133.67 & 1.44 & 1.70\\
+6   & 91.57 & 106.44 & 1.47 & 1.46\\
+7   & 91.49 & 104.89 & 1.49 & 1.44\\
+8   & 91.09 & 101.33 & 1.53 & 1.45\\
+10  & 90.52 & 101.78 & 1.60 & 1.53\\
+12  & 90.00 & 101.67 & 1.65 & 1.62\\
+14  & 89.29 & 103.00 & 1.70 & 1.75\\
+16  & 89.42 & 101.33 & 1.74 & 1.76\\
+20  & 88.61 & 100.89 & 1.82 & 1.94\\
+24  & 88.26 & 101.33 & 1.89 & 2.08\\
+28  & 87.57 & 99.22 & 1.96 & 2.24\\
+32  & 87.01 & 100.00 & 1.99 & 2.36\\
+40  & 86.23 & 98.78 & 2.09 & 2.64\\
+50  & 84.93 & 98.78 & 2.17 & 2.88\\
+64  & 84.17 & 101.22 & 2.28 & 3.19\\
+128 & 81.34 & 107.44 & 2.57 & 4.16\\
+1024 & 73.11 & 222.01 & 2.83 & 7.55
 	\end{tabular}
@@
 	\caption{Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
-We compare the base settings of 1:1 (Vampire's default), 1:5 (the best-behaved from Table \ref{tab:no-best-awr}), and the worst setting for the problem.}
+We compare the base settings of 1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.}
 	\centering
@@
 In this work we selected two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
-Since a simple linear shape does not extend well to indefinite proof search (it is unclear what should happen after either 1:1 or the target AWR is reached), an exponential decay function is used instead, as shown in Figure \ref{fig:decay-and-converge}.
-
-\todo{What are these figures showing?}
-
-These shapes are further parameterised by an integral \emph{shape frequency} setting, which controls the rate of decay or convergence: every \(n\) steps, the difference between the current and the target AWR is halved, rounding where necessary.
+Since a simple linear shape does not extend well to indefinite proof search (it is unclear what should happen after either 1:1 or the target AWR is reached), an exponential decay function is used instead.
+These exponential shapes are further parameterised by an integral \emph{shape frequency} setting, which controls the rate of decay or convergence: every \(n\) steps, the difference between the current and the target AWR is halved, rounding where necessary.
 In future, this might allow the use of repeating patterns such as a sinusoid, hence \emph{frequency}.
+Figure \ref{fig:decay-and-converge} shows rates at which the new configurations converge or decay from the fixed AWR setting for some indicative frequency settings.
 
@@
 \label{sec:experiments}
-
-\todo{Change marketing here. Success is adding new options to portfolio mode. Only mention CASC once and refer to portfolio mode elsewhere.}
-
-In practice, decaying from an age-biased initial AWR to 1:1 (or converging backwards) was not found to be an improvement.
-However, age/weight shapes decaying from or converging to a weight-based AWR were found to be useful in some cases.
-An experimental evaluation of this approach on the TPTP~\cite{tptp} benchmark follows.
-
+\todo{Only mention CASC once and refer to portfolio mode elsewhere.}
+An experimental evaluation of this approach on the TPTP\footnote{version 7.1.0} benchmark~\cite{tptp} follows.
 Vampire first ran to establish baseline performance in its competition \emph{CASC mode} on all problems in TPTP, with a wallclock time limit of 300 seconds.
 For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.
+
+\subsection{Benchmark Performance}
 The baseline solved 13,057 problems in total.
 No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
+The union of all \emph{converge} and \emph{decay} configurations improved on the baseline, with 13,167 and 13,106 solved problems respectively.
 
-\subsection{Benchmark Performance}
 Figure \ref{tab:results} shows the performance in terms of solved problems of all the configurations tested.
@@
 \subsection{New Problems}
-Some problems were solved which were marked as ``Unknown'' status in the TPTP headers.
+Some problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
 Vampire baseline (and several configurations) were able to solve \texttt{GEO120+1}, but only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.

### 2019-02-28 papers/CADE19/paper.tex
add conclusions section

@@
 
-After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes two main the contributions.
-Firstly, in Section~\ref{sec:awr:study} we experimentally confirm the folklore that (i) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (ii) there is no `best' age/weight ratio, indeed, there exist a large range of pragmatically useful ratios.
+After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes two main contributions.
+Firstly, in Section~\ref{sec:awr:study} we experimentally confirm the folklore that (i) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (ii) there is no `best' age/weight ratio: indeed, a large range of pragmatically useful ratios exists.
 Secondly, we demonstrate (Section~\ref{sec:varying:study}) that varying the age-weight ratio over time can achieve better performance than a fixed ratio and use this to motivate the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
@@
 \label{sec:conclusions}
+Age/weight ratios have a significant effect on the performance of proof search in the Vampire theorem prover, but there is no clear optimal age/weight ratio.
+Further, we have demonstrated that varying the age/weight ratio over time \emph{during} proof search can improve further on an optimal, but fixed age/weight ratio in terms of the number of activations.
+Experiments within Vampire on the TPTP benchmark suggest that these \emph{age/weight shapes} show promise for future developments in this novel approach to proof search.
 
-\todo[inline]{this is a thing, you can do it, you should do it (?), we can probably do better}
-\todo[inline]{more shapes, better ways of doing frequency decay, integrate into existing strategy schedules}
-
+Future directions for research in this area include trying a greater number of ``shapes'' (such as repeating patterns), other approaches for parameterising these shapes, a pseudo-random approach to age/weight instead of the balancing algorithm, and integration of the new approaches into existing strategy schedules.
 \bibliographystyle{plain}

### 2019-02-28 papers/CADE19/paper.tex
more figures

@@
 	\caption{
-The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over an entire problem set causes.
+The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over the problem set (approx. 8,000 problems solved in 1 second) causes.
 The percentage of problems in this set solved by the AWR setting is also shown.
@@
 A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
-\sandm~found that 1:5 had a similar property, but this might be explained by differences in test environment.
+\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment.
 }
 	\centering
-	\begin{tabular}{l | c | c | c  c }
-		AWR & \% Solved & Maximum Drop & Mean Drop & (Standard Deviation)\\
+	\begin{tabular}{l | c | c | c | c  c }
+		AWR & \% Solved & Uniques & Maximum Drop & Mean Drop & (Standard Deviation)\\
 \hline
-8:1 & 85.25 & 150.67 & 1.37 & 1.98\\
-5:1 & 86.10 & 122.22 & 1.33 & 1.64\\
-4:1 & 86.93 & 101.44 & 1.32 & 1.42\\
-3:1 & 87.63 & 105.00 & 1.29 & 1.41\\
-2:1 & 88.62 & 112.67 & 1.27 & 1.45\\
-3:2 & 89.83 & 119.89 & 1.27 & 1.51\\
-5:4 & 89.98 & 125.00 & 1.26 & 1.55\\
-1   & 90.56 & 133.56 & 1.26 & 1.63\\
-2:3 & 91.20 & 147.67 & 1.28 & 1.79\\
-2   & 91.68 & 162.67 & 1.31 & 1.97\\
-3   & 91.81 & 190.56 & 1.37 & 2.30\\
-4   & 91.85 & 17.41 & 1.38 & 0.67\\
-5   & 92.00 & 133.67 & 1.44 & 1.70\\
-6   & 91.57 & 106.44 & 1.47 & 1.46\\
-7   & 91.49 & 104.89 & 1.49 & 1.44\\
-8   & 91.09 & 101.33 & 1.53 & 1.45\\
-10  & 90.52 & 101.78 & 1.60 & 1.53\\
-12  & 90.00 & 101.67 & 1.65 & 1.62\\
-14  & 89.29 & 103.00 & 1.70 & 1.75\\
-16  & 89.42 & 101.33 & 1.74 & 1.76\\
-20  & 88.61 & 100.89 & 1.82 & 1.94\\
-24  & 88.26 & 101.33 & 1.89 & 2.08\\
-28  & 87.57 & 99.22 & 1.96 & 2.24\\
-32  & 87.01 & 100.00 & 1.99 & 2.36\\
-40  & 86.23 & 98.78 & 2.09 & 2.64\\
-50  & 84.93 & 98.78 & 2.17 & 2.88\\
-64  & 84.17 & 101.22 & 2.28 & 3.19\\
-128 & 81.34 & 107.44 & 2.57 & 4.16\\
-1024 & 73.11 & 222.01 & 2.83 & 7.55
+8:1 & 85.25 & 16 & 150.67 & 1.37 & 1.98\\
+5:1 & 86.10 & 1 & 122.22 & 1.33 & 1.64\\
+4:1 & 86.93 & 1 & 101.44 & 1.32 & 1.42\\
+3:1 & 87.63 & 2 & 105.00 & 1.29 & 1.41\\
+2:1 & 88.62 & 3 & 112.67 & 1.27 & 1.45\\
+3:2 & 89.83 & 2 & 119.89 & 1.27 & 1.51\\
+5:4 & 89.98 & 4 & 125.00 & 1.26 & 1.55\\
+1   & 90.56 & 4 & 133.56 & 1.26 & 1.63\\
+2:3 & 91.20 & 9 & 147.67 & 1.28 & 1.79\\
+2   & 91.68 & 0 & 162.67 & 1.31 & 1.97\\
+3   & 91.81 & 5 & 190.56 & 1.37 & 2.30\\
+4   & 91.85 & 3 & \textbf{17.41} & 1.38 & 0.67\\
+5   & 92.00 & 2 & 133.67 & 1.44 & 1.70\\
+6   & 91.57 & 1 & 106.44 & 1.47 & 1.46\\
+7   & 91.49 & 1 & 104.89 & 1.49 & 1.44\\
+8   & 91.09 & 2 & 101.33 & 1.53 & 1.45\\
+10  & 90.52 & 1 & 101.78 & 1.60 & 1.53\\
+12  & 90.00 & 0 & 101.67 & 1.65 & 1.62\\
+14  & 89.29 & 4 & 103.00 & 1.70 & 1.75\\
+16  & 89.42 & 5 & 101.33 & 1.74 & 1.76\\
+20  & 88.61 & 3 & 100.89 & 1.82 & 1.94\\
+24  & 88.26 & 2 & 101.33 & 1.89 & 2.08\\
+28  & 87.57 & 2 & 99.22 & 1.96 & 2.24\\
+32  & 87.01 & 1 & 100.00 & 1.99 & 2.36\\
+40  & 86.23 & 4 & 98.78 & 2.09 & 2.64\\
+50  & 84.93 & 1 & 98.78 & 2.17 & 2.88\\
+64  & 84.17 & 2 & 101.22 & 2.28 & 3.19\\
+128 & 81.34 & 3 & 107.44 & 2.57 & 4.16\\
+1024 & 73.11 & 23 & 222.01 & 2.83 & 7.55
 	\end{tabular}
-
 	\label{tab:no-best-awr}
@@
 \begin{table}
-	\caption{Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
-We compare the base settings of 1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.}
+	\caption{
+Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
+We compare the base settings of 1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
+Where the problem is not solved at all by the base setting, it is ignored.
+}
 	\centering
@@
 		\hline
-		converge & (combined) & 13167 & 117 & -- & --\\
-		decay & (combined) & 13106 & 93 & -- & --\\
+		converge & (combined) & 13167 & 117 & 41 & --\\
+		decay & (combined) & 13106 & 93 & 17 & --\\
 	\end{tabular}

### 2019-02-28 papers/CADE19/paper.tex
tidy

@@
 \usepackage{todonotes}
+%\newcommand{\todo}[1]{}
 
@@
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 As an initial illustrative example of how varying the AWR effects the number of clauses required to be processed before a proof is found consider Figure \ref{fig:example-optimal-awr}.
-This shows \todo{FINISH}
-
-  shows the effect of varying AWR on an illustrative problem from TPTP: a smaller number of activations means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
+This demonstrates the effect that varying AWR can have: a smaller number of activations means that fewer clauses were processed, which in general means that a proof was found faster\footnote{
 It should be noted that if a small number of clauses are extremely expensive to process it may be slower than a larger number of less-expensive clauses, but in general this is a good heuristic measure for prover performance.
@@
 
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=0.8\textwidth]{example-optimal-awr}
-	\caption{
-The number of activations reported by Vampire after successful 1-second runs on a TPTP problem.
-In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but does follow an approximate trend and settles toward the global optimum.
-These properties are typical for the observed set of problems.
-}
-	\label{fig:example-optimal-awr}
-\end{figure}
-
 This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, problems which were not solved by the default mode in under a second for any AWR value, and problems solved in pre-processing.
 The whole set yielded similar AWR data for 7,947 problems.
-These data show that while there is no ``best'' AWR for the whole set (Table \ref{tab:no-best-awr}), choosing a good AWR value for a problem is well-rewarded (Table \ref{tab:awr-improvement}).
+%
+We selected a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode.
+These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
 
-\todo{This discussion should be expanded. Results need more explanation and we should summarise e.g. yes we confirmed assumptions.}
+The results of these experiments are presented in Tables \ref{tab:no-best-awr} and \ref{tab:awr-improvement}. 
+%
+Table \ref{tab:no-best-awr}  shows that there is no ``best'' AWR for the whole set. 
+%
+Table \ref{tab:awr-improvement} shows that choosing a good AWR value for a problem is well-rewarded. 
+%
+Therefore, these experimental results confirm the previous assumptions often made in folklore.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 	\subfloat{\includegraphics[width=0.5\textwidth]{shape-decay}}
@@
 
+\clearpage
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-03-01 papers/CADE19/paper.tex
edits

@@
 \subsection{Logarithmic AWR}
+
 Optimising AWR values is more easily achieved if they have a continuous scale.
@@
 \(L = 0\) represents the middle ground of a 1:1 age/weight ratio.
+Note that we have not updated the balancing algorithm used by Vampire to make use of this value (this still requires two numbers) but use the number in experiments when discussing AWR.
 
@@
 
-This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not support, problems which were not solved by the default mode in under a second for any AWR value, and problems solved in pre-processing.
+This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g.
+higher-order problems).
+We ran Vampire in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode.
+These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
+We then filter out problems not solved by any of these or those solved trivially (e.g in preprocessing).
 The whole set yielded similar AWR data for 7,947 problems.
 %
-We selected a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode.
-These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
+We selected
 

### 2019-03-01 papers/CADE19/paper.tex
section3

@@
 %%%%%%
-\paragraph{Related Work.} \todo{Mention clause selection strategies of Otter, Prover9, E, and SPASS}
+\paragraph{Related Work.} We briefly review the clause selection approaches taken by other solvers. 
+%
+Otter \cite{} and Prover9 \cite{} CHECK BUT THEY PROBABLY  ALLOW BOTH.
+%
+E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue.
+The default is to use a combination of age and weight selection although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}.
+%
+SPASS~\cite{Spass} uses symbol-counting based clause selection. 
+%
+iProver~\cite{} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
+%
+The general idea in this paper of \emph{vary the ratio over time} can be applied to any ratio-based clause selection strategy and our specific results apply to those that take a ratio between age and weight.
 
-\newpage
+%\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{table}[t]
+	\caption{
+Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
+We compare  1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
+Where the problem is not solved at all by the base setting, it is ignored.
+}
+	\centering
+	\begin{tabular}{c | c | c c}
+		& Maximum Gain & Mean Gain & (Standard Deviation)\\
+		\hline
+		1:1 & 133.56 & 1.26 & 1.63 \\
+		1:5 & 133.67 & 1.44 & 1.70 \\
+		(worst) & 222.01 & 3.95 & 7.60 \\
+	\end{tabular}
+	\label{tab:awr-improvement}
+\end{table}
+
 This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g.
 higher-order problems).
-We ran Vampire in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode.
+We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode.
 These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
@@
 %
-We selected
 
-The results of these experiments are presented in Tables \ref{tab:no-best-awr} and \ref{tab:awr-improvement}. 
-%
-Table \ref{tab:no-best-awr}  shows that there is no ``best'' AWR for the whole set. 
-%
-Table \ref{tab:awr-improvement} shows that choosing a good AWR value for a problem is well-rewarded. 
-%
-Therefore, these experimental results confirm the previous assumptions often made in folklore.
+The first result is that choosing a good AWR value for a problem is well-rewarded.
+Table~\ref{tab:awr-improvement} summarises the impact that choosing the best AWR can have.
+Compared to the default we can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr}) shows we also lose just under 10\% of problems by choosing the default.
+It is more relevant to note that there are cases where we can do \emph{much} better by selecting a different AWR value.
+So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster (not necessarily).
+In the worst case we can do almost 4x worse on average by always choosing the worst AWR value.
+
+The second result is that there is no ``best'' AWR across this full set of problems.
+Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance.
+The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best.
+No AWR value solves all problems.
+A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
+\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment.
+It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely.
+This is typically in saturation-based proof search; approaches that do not perform well in general may perform well in specific cases where the general approach does not.
+
+In summary, these results confirm the previous assumptions often made in folklore.
+It should be noted that this is a small experiment (1 second runs in default mode) and the relative performance of different AWR values cannot be generalised, but the general result that they are complementary can.
 
 \begin{table}[t]
-	\caption{
-The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over the problem set (approx. 8,000 problems solved in 1 second) causes.
-The percentage of problems in this set solved by the AWR setting is also shown.
-No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
-A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
-\sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment.
+	\caption{Per-AWR value results on 1 second runs over 7,947 TPTP problems.
+%The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over the problem set (approx. 8,000 problems solved in 1 second) causes.
+%The percentage of problems in this set solved by the AWR setting is also shown.
+%No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
 }
@@
 
-\begin{table}
-	\caption{
-Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
-We compare the base settings of 1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
-Where the problem is not solved at all by the base setting, it is ignored.
-}
-	\centering
-	\begin{tabular}{c | c | c c}
-		& Maximum Gain & Mean Gain & (Standard Deviation)\\
-		\hline
-		1:1 & 133.56 & 1.26 & 1.63 \\
-		1:5 & 133.67 & 1.44 & 1.70 \\
-		(worst) & 222.01 & 3.95 & 7.60 \\
-	\end{tabular}
-	\label{tab:awr-improvement}
-\end{table}
-
-\clearpage
+%\clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-03-01 papers/CADE19/paper.tex
updates

@@
 \usepackage{todonotes}
+%\usepackage{multicol}
 %\newcommand{\todo}[1]{}
@@
 
-Vampire is a portfolio solver and is typically run in a mode that attempts multiple different \emph{strategies} in quick succession e.g.
-in a 30 second run it may attempt 10 or more different strategies.
+Vampire is a portfolio solver and is typically run in a mode that attempts multiple different \emph{strategies} in quick succession, e.g.
+in a 30 second run it may attempt 10 or more different strategies, and may run these in parallel with different priorities~\cite{DBLP:conf/cade/RawsonR18}.
 These strategies employ many different options including different saturation algorithms (including Otter and Discount), preprocessing options, literal selection strategies, inference rules, and clause selection heuristics.
@@
 %
-iProver~\cite{} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
+iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
 %
@@
 The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best.
-No AWR value solves all problems.
+No AWR value solves all problems, with the best being 1:5.
 A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
@@
 
-Although choosing a good AWR value is important, this is covered in part by the use of strategy scheduling~\cite{CADE18} in which many AWR values are tried in sequence (along with other prover options).
+Although choosing a good AWR value is important, this is covered in part by the use of strategy scheduling in which many AWR values are tried in sequence (along with other prover options).
 Additionally, given that varying the AWR can have such a large impact, it seems likely that a constant AWR fixed for the entire proof search is unlikely to be optimal for any given problem.
@@
 Applying this method with 10,000 repetitions to the problem seen earlier (\texttt{PRO017+2}) yields the example AWR trend shown in Figure \ref{fig:random-walk}, which reduces the best number of activations from 330 with a fixed AWR, to 287 with a varying AWR.
-
+%
 Unsurprisingly, in ad-hoc experiments on other problems, the best shape is rarely constant.
@@
 %%%%%%%
-\subsection{Varying AWR in Vampire} \label{sec:varying:implementation}
+\subsection{Varying AWR (in Vampire)} \label{sec:varying:implementation}
 
-If changing AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
+We describe how we vary AWR dynamically and how this is achieved in Vampire. 
+%
+%If dynamically varying AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
 In general we would like to describe any possible sequence that the AWR could follow during proof search.
@@
 
-In this work we selected two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
+In this work we select two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
+In our investigations we found that even fluctuating sequences had a general trend and these two fixed trends are reasonable approximations of this.
+We have chosen to use an `original' AWR value to move away from/towards as this original AWR has already been shown to be `useful' in some sense.
+
 Since a simple linear shape does not extend well to indefinite proof search (it is unclear what should happen after either 1:1 or the target AWR is reached), an exponential decay function is used instead.
@@
 In future, this might allow the use of repeating patterns such as a sinusoid, hence \emph{frequency}.
-Figure \ref{fig:decay-and-converge} shows rates at which the new configurations converge or decay from the fixed AWR setting for some indicative frequency settings.
+Figure \ref{fig:decay-and-converge} illustrates rates at which the new configurations converge or decay from the fixed AWR setting for some indicative frequency settings.
 
@@
 
-Together, these new settings allow a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.
+We introduce two new options: \texttt{--age\_weight\_ratio\_shape} can take the values \emph{constant}, \emph{decay}, or \emph{converge} and selects one of the above shapes; and \texttt{--age\_weight\_ratio\_shape\_frequency} specifies the frequency (rate) or convergance/decay (default is 100).
+These are used with the existing \texttt{--age\_weight\_ratio} option (default 1:1) to give a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.
 This version of the prover is currently in a separate branch in the main Vampire source repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.
 
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:experiments}
-\todo{Only mention CASC once and refer to portfolio mode elsewhere.}
-An experimental evaluation of this approach on the TPTP\footnote{version 7.1.0} benchmark~\cite{tptp} follows.
-Vampire first ran to establish baseline performance in its competition \emph{CASC mode} on all problems in TPTP, with a wallclock time limit of 300 seconds.
-For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.
 
-\subsection{Benchmark Performance}
-The baseline solved 13,057 problems in total.
-No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
-The union of all \emph{converge} and \emph{decay} configurations improved on the baseline, with 13,167 and 13,106 solved problems respectively.
+We perform two experiments to evaluate the new techniques.
+The first compares the various options attempting to draw some conclusions about which option values work well together.
+The second looks at how useful the new options are in the context of portfolio solving.
+Both experiments use the TPTP (version 7.1.0) benchmark~\cite{tptp} and were run on StarExec \cite{starexec}.
 
-Figure \ref{tab:results} shows the performance in terms of solved problems of all the configurations tested.
-These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
-In total, 134 problems were solved by the new configurations that were not solved by the baseline.
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparing New Options}
 
-\begin{table}
+\begin{table}[t]
+\caption{Number of problems solved (top) and unique problems solved (bottom) by various configurations varying AWR, AWR shape, and AWR frequency.\label{tab:short:runs}}
+\centering
+\begin{tabular}{l|l|lllll|lllll}
+&&  \multicolumn{5}{|c}{converge} &  \multicolumn{5}{|c}{decay} \\  
+AWR & constant & \multicolumn{5}{|c|}{Frequency}  & \multicolumn{5}{|c}{Frequency} \\  
+&& 1 & 10 & 100 & 1000 & Union & 1 & 10 & 100 & 1000 &Union  \\ \hline
+\multicolumn{10}{c}{Problems Solved} \\ \hline
+10:1		&5911 & 5908&5910 & 5969 &6135 &  & 5908 &5910  & 5969  &6135  \\
+1:10		&6417& 6412&6421 &6389 & 6325  \\
+1:100	&6026& 6028& 6010& 5971& 6148\\
+1:1000	&5387&5395  &5408 &5510 &6019  \\ \hline
+Union & \\ \hline \hline
+\multicolumn{10}{c}{Uniquely Solved} \\ \hline
+10:1		&1 & 4&2 & 5 &10 & 4&2 & 5&10 \\
+1:10		& \\
+1:100	& \\
+1:1000	&\\ \hline
+Union & \\ 
+\end{tabular}
+\end{table}
+
+We ran Vampire in default mode (with {\tt discount} saturation algorithm) for 10s whilst varying  \texttt{age\_weight\_ratio} and \texttt{age\_weight\_ratio\_shape\_frequency} for both new values of \texttt{age\_weight\_ratio\_shape}.
+We also ran the constant default AWR of 1:1 as a baseline.
+
+The results are given in Table~\ref{tab:short:runs}.
+The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value.
+We separately report the total number of problems solved and those solved uniquely (by a single strategy across all experiments).
+
+The best combination of options overall was X. In general, strategies that converged performed better/worse than those that diverged.
+
+In total X out of all Y strategies contributed some unique solutions. This shows that these new options are complementary in general.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Contribution to Portfolio}
+
+\begin{table}[t]
 	\caption{
@@
 		\hline
-		baseline & -- & 13057 & 0 & 1 & 714.2\\
-		converge & 1 & 13039 & 24 & 3 & 714.3\\
-		converge & 5 & 13029 & 27 & 1 & 709.5\\
-		converge & 10 & 13028 & 35 & 5 & 714.3\\
-		converge & 50 & 13015 & 45 & 5 & 712.8\\
-		converge & 100 & 12976 & 51 & 1 & 705.9\\
-		converge & 500 & 12895 & 63 & 4 & 698.3\\
-		converge & 1000 & 12837 & 52 & 0 & 688.6\\
-		converge & 5000 & 12775 & 53 & 1 & 682.4\\
-		converge & 10000 & 12751 & 53 & 0 & 678.7\\
-		decay & 1 & 12698 & 48 & 1 & 673.6\\
-		decay & 5 & 12702 & 51 & 1 & 674.9\\
-		decay & 10 & 12698 & 48 & 1 & 674.2\\
-		decay & 50 & 12712 & 49 & 2 & 679.1\\
-		decay & 100 & 12726 & 46 & 1 & 678.8\\
-		decay & 500 & 12795 & 29 & 1 & 685.5\\
-		decay & 1000 & 12860 & 29 & 2 & 692.6\\
-		decay & 5000 & 12982 & 16 & 2 & 707.1\\
-		decay & 10000 & 13002 & 7 & 0 & 706.3\\
+		baseline & -- & \textbf{13,057} & 0 & 1 & 714.2\\ \hline
+		converge & 1 & 13,039 & 24 & 3 & 714.3\\
+		converge & 5 & 13,029 & 27 & 1 & 709.5\\
+		converge & 10 & 13,028 & 35 & 5 & 714.3\\
+		converge & 50 & 13,015 & 45 & 5 & 712.8\\
+		converge & 100 & 12,976 & 51 & 1 & 705.9\\
+		converge & 500 & 12,895 & 63 & 4 & 698.3\\
+		converge & 1000 & 12,837 & 52 & 0 & 688.6\\
+		converge & 5000 & 12,775 & 53 & 1 & 682.4\\
+		converge & 10000 & 12,751 & 53 & 0 & 678.7\\ \hline
+		decay & 1 & 12,698 & 48 & 1 & 673.6\\
+		decay & 5 & 12,702 & 51 & 1 & 674.9\\
+		decay & 10 & 12,698 & 48 & 1 & 674.2\\
+		decay & 50 & 12,712 & 49 & 2 & 679.1\\
+		decay & 100 & 12,726 & 46 & 1 & 678.8\\
+		decay & 500 & 12,795 & 29 & 1 & 685.5\\
+		decay & 1000 & 12,860 & 29 & 2 & 692.6\\
+		decay & 5000 & 12,982 & 16 & 2 & 707.1\\
+		decay & 10000 & 13,002 & 7 & 0 & 706.3\\
 		\hline
-		converge & (combined) & 13167 & 117 & 41 & --\\
-		decay & (combined) & 13106 & 93 & 17 & --\\
+		converge & (combined) & 13,167 & 117 & 41 & --\\
+		decay & (combined) & 13,106 & 93 & 17 & --\\
 	\end{tabular}
@@
 
-\subsection{New Problems}
-Some problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
-Vampire baseline (and several configurations) were able to solve \texttt{GEO120+1}, but only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.
+Our next experiment aims to answer the question ``how much can the portfolio mode of Vampire improve using these new options''.
+To address this question we use the portfolio mode used in the most recent CASC competition~\cite{CADE18} as a baseline and run the new options on top of this.
+Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio e.g.
+finite model building~\cite{DBLP:conf/sat/Reger0V16}.
 
-\clearpage
+Vampire first ran to establish baseline performance in the given portfolio mode on all problems in TPTP, with a wallclock time limit of 300 seconds.
+We then forced the new options on top of the portfolio mode, using the existing AWR values in the various strategies as the starting point.
+The purpose is to gauge what impact adding such options to a new portfolio mode could have.
+In this experiment we are mainly hoping to find new solved problems and identify new strategies that could be added to a portfolio mode.
+Therefore, it makes sense to consider the union of all experiments ran.
+%For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.
+
+Overall, the baseline solved the most problems (13,057)
+No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
+The union of all \emph{converge} and \emph{decay} configurations improved on the baseline, with 13,167 and 13,106 solved problems respectively.
+
+Figure \ref{tab:results} shows the performance in terms of solved problems of all the configurations tested.
+These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
+In total, 134 problems were solved by the new configurations that were not solved by the baseline.
+
+%\subsection{New Problems}
+Two problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
+Only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.
+
+%\clearpage
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparison with other Solvers}
+
+\begin{table}[t]
+\caption{Total number of problems solved compared to other solvers.\label{tab:other}}
+\centering
+\begin{tabular}{l|l|l}
+Solver 		& Total solved & Uniquely solved \\ \hline
+Vampire (old)	& 13,057 & \\
+Vampire (new)   & 13,191 & \\
+E 			& \\
+iProver		& \\
+CVC4		& \\
+\end{tabular}
+\end{table}
+
+To place these results in context, we compare the overall number of problems solved by our new strategies with the results of other solvers (using their previous CASC competition strategy).
+These results are from 300s and are given in Table~\ref{tab:other}.
 
@@
 \label{sec:conclusions}
-Age/weight ratios have a significant effect on the performance of proof search in the Vampire theorem prover, but there is no clear optimal age/weight ratio.
+
+Clause selection is a key part of any saturation-based theorem prover and age/weight ratios have a significant effect on the performance of proof search in the Vampire theorem prover.
+We have supported the known folklore that there is no clear optimal age/weight ratio.
 Further, we have demonstrated that varying the age/weight ratio over time \emph{during} proof search can improve further on an optimal, but fixed age/weight ratio in terms of the number of activations.
 Experiments within Vampire on the TPTP benchmark suggest that these \emph{age/weight shapes} show promise for future developments in this novel approach to proof search.
+Indeed, including our relatively simple shapes already leads to significant performance gains.
 
-Future directions for research in this area include trying a greater number of ``shapes'' (such as repeating patterns), other approaches for parameterising these shapes, a pseudo-random approach to age/weight instead of the balancing algorithm, and integration of the new approaches into existing strategy schedules.
+Future directions for research include trying a greater number of ``shapes'' (such as repeating patterns), other approaches for parameterising these shapes, a pseudo-random approach to age/weight instead of the balancing algorithm, and integration of the new approaches into existing strategy schedules.
 \bibliographystyle{plain}

### 2019-03-02 papers/CADE19/paper.tex
proof reading

@@
 \usepackage{amsmath}
-\usepackage{todonotes}
-%\usepackage{multicol}
-%\newcommand{\todo}[1]{}
 
@@
 \begin{abstract}
-Modern saturation-based theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process.
+Modern saturation theorem provers are based on the given-clause algorithm, which iteratively selects new clauses to process.
 This clause selection has a large impact on the performance of proof search and has been the subject of much folklore.
@@
 Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
-Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or shapes) that prove to be pragmatically useful.
+Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or \emph{shapes}) that prove to be pragmatically useful.
 
@@
 \label{sec:background}
-
 This section introduces the relevant background for the rest of the paper.
 
-%%%%%%
 \paragraph{First-Order Logic and Weight.} Our setting is the standard first-order predicate logic with equality.
@@
 
-\iffalse
-Terms are of the form $f(t_1,\ldots,t_n), c$ or $x$ where
-$f$ is a \emph{function symbol} of arity $n \geq 1$, $t_1,\ldots, t_n$ are terms, $c$ is a zero arity function symbol (i.e. a constant) and $x$ is a variable. 
-%
-The \emph{weight} of a term $t$ is defined as $w(t) =1$ if $t$ is a variable or a constant
-and as $w(t) = 1+\sum_{i=1,\ldots,n} w(t_i)$ if $t$ is of the form
-$f(t_1,\ldots,t_n)$. In other words, the weight or a term is the
-number of symbols in it.
-%
-Atoms are of the form $p(t_1,\ldots,t_n), q$ or $t_1 \eql t_2$ where $p$ is a \emph{predicate symbol} of arity $n$, $t_1, \ldots, t_n$ are terms, $q$ is a zero arity predicate symbol and $\eql$ is the \emph{equality symbol}.
-%
-The weight function naturally extends to atoms: $w(p(t_1,\ldots,t_n)) = 1+\sum_{i=1,\ldots,n} w(t_i)$, $w(t_1 \eql t_2) = w(t_1)+w(t_2)$,
-and $w(q)=1$.
-%
-A literal is either an atom $A$, in which case we call it \emph{positive}, or a negation $\notl A$,
-in which case we call it \emph{negative}.
-%We write negated equalities as $t_1 \neql t_2$. 
-%
-The weight of a literal is the weight of the corresponding atom. 
-% 
-%We write $t[s]_p$ and $L[s]_p$ to denote that a term $s$ occurs in a term $t$ (in a literal $L$) at a position $p$.
-%
-A \emph{clause} is a disjunction of literals $L_1 \orl \ldots \orl L_n$ for $n \geq 0$.
-We disregard the order of literals and treat a clause as a multiset. 
-%
-The weight of a clause is the sum of the weight of its literals. 
-\fi
-
-%%%%%%
 \paragraph{Saturation-Based Proof Search.}
-
-Saturation-based theorem provers \emph{saturate} a set of clauses $S$ with respect to an inference system $\mathbb{I}$ e.g.
-they compute a set of clauses $S'$ by applying rules in $\mathbb{I}$ to clauses in $S$ until no new clauses are generated.
-%
+Saturation-based theorem provers \emph{saturate} a set of clauses $S$ with respect to an inference system $\mathbb{I}$: that is, computing a set of clauses $S'$ by applying rules in $\mathbb{I}$ to clauses in $S$ until no new clauses are generated.
 If the empty clause is generated then $S$ is unsatisfiable.
-%
 Calculi such as resolution and superposition have conditions that ensure \emph{completeness}, which means that a saturated set $S$ is satisfiable if it does not contain the empty clause as an element.
-%
 As first-order logic is only semi-decidable, it is not necessarily the case that $S$ has a finite saturation, and even if it does it may be unachievable in practice using the available resources. 
-%
 Therefore, much effort in saturation-based first-order theorem proving involves 
 controlling proof search to make finding the empty clause more likely (within reasonable resource bounds). 
-%
 One important notion is that of \emph{redundancy}, being able to remove clauses from the search space that are not required. 
-%
 Another important notion are literal selections that place restrictions on the inferences that can be performed. 
-%
 Both notions come with additional requirements for completeness.
@@
 
-%%%%%%
 \paragraph{The Given Clause Algorithm and AWR Clause Selection.}
@@
 Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop.
-%
 Realisations of the given clause algorithm generally differ in how they organise simplifications.
-There are two main approaches (both implemented by Vampire): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.
+There are two main approaches (both implemented by Vampire, originally found in the eponymous theorem provers Otter~\cite{otter} and \textsc{Discount}~\cite{discount}): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.
 
@@
 The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are (in some sense) closer to this.
-Furthermore, lighter clauses are friendlier (less explosive) to proof search in various ways e.g.
-they are more general in terms of subsumption and tend to have fewer children.
+Furthermore, lighter clauses are more general in terms of subsumption and tend to have fewer children, making them less explosive in terms of proof search.
 \end{itemize}
-It has been shown (see \sandm{}) that alternating these heuristics is beneficial and we consider this setting here.
+\sandm~show that alternating these heuristics is beneficial and we consider this setting here.
 In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm.
 The balance is initialised to 0 and used as follows: a positive balance means that a clause should be selected by age, whereas a negative balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
-
 Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.
@@
 %%%%%%
-\paragraph{Related Work.} We briefly review the clause selection approaches taken by other solvers. 
-%
-Otter \cite{} and Prover9 \cite{} CHECK BUT THEY PROBABLY  ALLOW BOTH.
-%
+\paragraph{Related Work.}
+We briefly review the clause selection approaches taken by other solvers. 
+Otter 3.3~\cite{otter3} selects \emph{either} by age, by weight or manually.
+Prover9~\cite{prover9} allows a configurable age/weight ratio.
 E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue.
-The default is to use a combination of age and weight selection although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}.
-%
+The default is to use a combination of age and weight selection, although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}. 
 SPASS~\cite{Spass} uses symbol-counting based clause selection. 
-%
 iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
-%
-The general idea in this paper of \emph{vary the ratio over time} can be applied to any ratio-based clause selection strategy and our specific results apply to those that take a ratio between age and weight.
+The general idea in this paper of a \emph{varying age/weight ratio over time} is applicable to any ratio-based clause selection strategy, and our specific results apply to those that take a ratio between age and weight.
 
-%\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Optimising Age/Weight Ratios} \label{sec:awr:study}
+\section{Optimising Age/Weight Ratios}
+\label{sec:awr:study}
 
@@
 The whole set yielded similar AWR data for 7,947 problems.
-%
 
@@
 It is more relevant to note that there are cases where we can do \emph{much} better by selecting a different AWR value.
-So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster (not necessarily).
-In the worst case we can do almost 4x worse on average by always choosing the worst AWR value.
+So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster, but not necessarily.
+In the worst case (choosing the pessimal AWR value) we can perform almost 4 times as many activations.
 
@@
 It is interesting to note that the extreme AWR values solve fewer problems overall but solve the most uniquely.
-This is typically in saturation-based proof search; approaches that do not perform well in general may perform well in specific cases where the general approach does not.
+This is typical in saturation-based proof search: approaches that do not perform well in general may perform well in specific cases where the general approach does not.
 
@@
 \begin{table}[t]
-	\caption{Per-AWR value results on 1 second runs over 7,947 TPTP problems.
-%The mean relative performance drop (in terms of activations compared to an optimal value) choosing a single AWR over the problem set (approx. 8,000 problems solved in 1 second) causes.
-%The percentage of problems in this set solved by the AWR setting is also shown.
-%No setting is significantly ``better'' or ``worse'' than the others, but some have more variance in performance.
+	\caption{
+Per-AWR value results on 1 second runs over 7,947 TPTP problems.
 	}

### 2019-03-02 papers/CADE19/paper.tex
remove comparative section
extra test in experiments

@@
 
-Overall, the baseline solved the most problems (13,057)
+Overall, the baseline solved the most problems (13,057). 
 No experimental configuration improved on this figure, but some problems not solved by baseline were solved by the new configurations, and some entirely new problems were solved.
@@
 In total, 134 problems were solved by the new configurations that were not solved by the baseline.
+This is an impressive result - it is rare to be able to improve portfolio mode by this many new problems with a single new proof search option.
+
+We also report the \emph{u-score} computed by giving $1/n$ points per problem solved where $n$ is the number of strategies solving a problem.
+This gives a measure of contribution per strategy.
+To consider new options for extending the portfolio mode we will consider options from largest u-score to smallest (and only those with unique solutions overall).
 
 %\subsection{New Problems}
-Two problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
+Finally, two problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
 Only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.
@@
 
+\iffalse
 %%%%%%%%%%%%%%%%%%%%%%
@@
 These results are from 300s and are given in Table~\ref{tab:other}.
+\fi
 

### 2019-03-02 papers/CADE19/paper.tex
fixed comparative section

@@
 Secondly, we demonstrate (Section~\ref{sec:varying:study}) that varying the age-weight ratio over time can achieve better performance than a fixed ratio and use this to motivate the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
-Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement, proving many new problems unsolvable by any current configuration of Vampire as well as other competitive solvers.
+Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement, proving many new problems unsolvable by any current configuration of Vampire.
 
@@
 
-%\subsection{New Problems}
 Finally, two problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
@@
 
-%\clearpage
-
-\iffalse
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Comparison with other Solvers}
-
-\begin{table}[t]
-\caption{Total number of problems solved compared to other solvers.\label{tab:other}}
-\centering
-\begin{tabular}{l|l|l}
-Solver 		& Total solved & Uniquely solved \\ \hline
-Vampire (old)	& 13,057 & \\
-Vampire (new)   & 13,191 & \\
-E 			& \\
-iProver		& \\
-CVC4		& \\
-\end{tabular}
-\end{table}
-
-To place these results in context, we compare the overall number of problems solved by our new strategies with the results of other solvers (using their previous CASC competition strategy).
-These results are from 300s and are given in Table~\ref{tab:other}.
-\fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions and Future Work}

### 2019-03-02 papers/CADE19/paper.tex
converge figures for 10secs

@@
 \begin{tabular}{l|l|lllll|lllll}
-&&  \multicolumn{5}{|c}{converge} &  \multicolumn{5}{|c}{decay} \\  
-AWR & constant & \multicolumn{5}{|c|}{Frequency}  & \multicolumn{5}{|c}{Frequency} \\  
-&& 1 & 10 & 100 & 1000 & Union & 1 & 10 & 100 & 1000 &Union  \\ \hline
+	& & \multicolumn{5}{|c}{converge} & \multicolumn{5}{|c}{decay}\\  
+AWR & constant & \multicolumn{5}{|c|}{Frequency} & \multicolumn{5}{|c}{Frequency}\\  
+&& 1 & 10 & 100 & 1000 & Union & 1 & 10 & 100 & 1000 & Union\\
+\hline
 \multicolumn{10}{c}{Problems Solved} \\ \hline
-10:1		&5911 & 5908&5910 & 5969 &6135 &  & 5908 &5910  & 5969  &6135  \\
-1:10		&6417& 6412&6421 &6389 & 6325  \\
-1:100	&6026& 6028& 6010& 5971& 6148\\
-1:1000	&5387&5395  &5408 &5510 &6019  \\ \hline
-Union & \\ \hline \hline
+10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & 8372\\
+1:10		& 8575 & 8565 & 8578 & 8550 & 8489 & 8778\\
+1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & 8560\\
+1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & 8379\\
+\hline
+Union           & 9019 & 9028 & 9016 & 8981 & 8871 & 9194\\
+\hline
+\hline
 \multicolumn{10}{c}{Uniquely Solved} \\ \hline
-10:1		&1 & 4&2 & 5 &10 & 4&2 & 5&10 \\
-1:10		& \\
-1:100	& \\
-1:1000	&\\ \hline
-Union & \\ 
+10:1		& 2 & 4 & 2 & 6 & 15 & 27\\
+1:10		& 2 & 5 & 3 & 2 & 12 & 22\\
+1:100	        & 1 & 2 & 2 & 9 & 13 & 26\\
+1:1000	        & 1 & 2 & 0 & 0 & 5 & 7\\
+\hline
+Union           & 6 & 13 & 7 & 17 & 45 & --\\ 
 \end{tabular}
@@
 We also ran the constant default AWR of 1:1 as a baseline.
+Experimental results for \emph{decay} were not complete at the time of submission but the trend is similar, e.g.
+some values performed better than others with a spread of unique results.
+These results will be added in a final version.
 

### 2019-03-03 papers/CADE19/paper.tex
tidy

@@
 \begin{tabular}{l|l|lllll|lllll}
-	& & \multicolumn{5}{|c}{converge} & \multicolumn{5}{|c}{decay}\\  
-AWR & constant & \multicolumn{5}{|c|}{Frequency} & \multicolumn{5}{|c}{Frequency}\\  
+	& & \multicolumn{5}{|c}{\texttt{converge}} & \multicolumn{5}{|c}{\texttt{decay}}\\  
+AWR & \texttt{constant} & \multicolumn{5}{|c|}{Frequency} & \multicolumn{5}{|c}{Frequency}\\  
 && 1 & 10 & 100 & 1000 & Union & 1 & 10 & 100 & 1000 & Union\\
@@
 \multicolumn{10}{c}{Problems Solved} \\ \hline
-10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & 8372\\
-1:10		& 8575 & 8565 & 8578 & 8550 & 8489 & 8778\\
-1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & 8560\\
-1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & 8379\\
+10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & \emph{8372}\\
+1:10		& 8575 & 8565 & \textbf{8578} & 8550 & 8489 & \emph{8778}\\
+1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & \emph{8560}\\
+1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & \emph{8379}\\
 \hline
-Union           & 9019 & 9028 & 9016 & 8981 & 8871 & 9194\\
+Union           & \emph{9019} & \emph{9028} & \emph{9016} & \emph{8981} & \emph{8871} & \emph{9194}\\
 \hline
@@
 \multicolumn{10}{c}{Uniquely Solved} \\ \hline
-10:1		& 2 & 4 & 2 & 6 & 15 & 27\\
-1:10		& 2 & 5 & 3 & 2 & 12 & 22\\
-1:100	        & 1 & 2 & 2 & 9 & 13 & 26\\
-1:1000	        & 1 & 2 & 0 & 0 & 5 & 7\\
+10:1		& 2 & 4 & 2 & 6 & \textbf{15} & \emph{27}\\
+1:10		& 2 & 5 & 3 & 2 & 12 & \emph{22}\\
+1:100	        & 1 & 2 & 2 & 9 & 13 & \emph{26}\\
+1:1000	        & 1 & 2 & 0 & 0 & 5 & \emph{7}\\
 \hline
-Union           & 6 & 13 & 7 & 17 & 45 & --\\ 
+Union           & \emph{6} & \emph{13} & \emph{7} & \emph{17} & \emph{45} & --\\ 
 \end{tabular}

### 2019-03-03 papers/CADE19/paper.tex
add decay results, fix references and tables

@@
 Vampire is a portfolio solver and is typically run in a mode that attempts multiple different \emph{strategies} in quick succession, e.g.
-in a 30 second run it may attempt 10 or more different strategies, and may run these in parallel with different priorities~\cite{DBLP:conf/cade/RawsonR18}.
+in a 30 second run it may attempt 10 or more different strategies, and may run these in parallel with different priorities~\cite{dynamic-strategy-priority}.
 These strategies employ many different options including different saturation algorithms (including Otter and Discount), preprocessing options, literal selection strategies, inference rules, and clause selection heuristics.
@@
 E~\cite{E} allows the user to specify an arbitrary number of priority queues and a weighted round-robin scheme that determines how many clauses are picked from each queue.
-The default is to use a combination of age and weight selection, although there is also a complex strategy developed by a genetic algorithm~\cite{DBLP:conf/gcai/SchaferS15}. 
+The default is to use a combination of age and weight selection, although there is also a complex strategy developed by a genetic algorithm~\cite{genetic-heuristics}. 
 SPASS~\cite{Spass} uses symbol-counting based clause selection. 
@@
 higher-order problems).
-We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:awr-improvement}) - these are the values used in Vampire's portfolio mode.
+We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:no-best-awr}) --- these are the values used in Vampire's portfolio mode.
 These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
@@
 \multicolumn{10}{c}{Problems Solved} \\ \hline
-10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & \emph{8372}\\
-1:10		& 8575 & 8565 & \textbf{8578} & 8550 & 8489 & \emph{8778}\\
-1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & \emph{8560}\\
-1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & \emph{8379}\\
+10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & \emph{8372} & 8448 & 8441 & 8380 & 8169 & \emph{8614}\\
+1:10		& 8575 & 8565 & \textbf{8578} & 8550 & 8489 & \emph{8778} & 8458 & 8456 & 8484 & 8268 & \emph{8787}\\
+1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & \emph{8560} & 8454 & 8484 & 8537 & \textbf{8636} & \emph{8907}\\
+1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & \emph{8379} & 8470 & 8492 & 8492 & 8473 & \emph{8873}\\
 \hline
-Union           & \emph{9019} & \emph{9028} & \emph{9016} & \emph{8981} & \emph{8871} & \emph{9194}\\
+Union           & \emph{9019} & \emph{9028} & \emph{9016} & \emph{8981} & \emph{8871} & \emph{9194} & \emph{8572} & \emph{8674} & \emph{8725} & \emph{8978} & \emph{9048}\\
 \hline
@@
 \multicolumn{10}{c}{Uniquely Solved} \\ \hline
-10:1		& 2 & 4 & 2 & 6 & \textbf{15} & \emph{27}\\
-1:10		& 2 & 5 & 3 & 2 & 12 & \emph{22}\\
-1:100	        & 1 & 2 & 2 & 9 & 13 & \emph{26}\\
-1:1000	        & 1 & 2 & 0 & 0 & 5 & \emph{7}\\
+10:1		& 1 & 3 & 2 & 3 & 8 & \emph{16} & 0 & 1 & 2 & 4 & \emph{7}\\
+1:10		& 1 & 3 & 2 & 1 & 5 & \emph{11} & 0 & 1 & 0 & 7 & \emph{8}\\
+1:100	        & 1 & 2 & 2 & 7 & \textbf{9} & \emph{20} & 2 & 0 & 3 & 9 & \emph{14}\\
+1:1000	        & 1 & 2 & 0 & 0 & 3 & \emph{5} & 2 & 0 & 0 & \textbf{12} & \emph{14}\\
 \hline
-Union           & \emph{6} & \emph{13} & \emph{7} & \emph{17} & \emph{45} & --\\ 
+Union           & \emph{4} & \emph{10} & \emph{6} & \emph{11} & \emph{25} & \emph{52} & \emph{4} & \emph{2} & \emph{5} & \emph{32} & \emph{43}\\ 
 \end{tabular}
@@
 We also ran the constant default AWR of 1:1 as a baseline.
-Experimental results for \emph{decay} were not complete at the time of submission but the trend is similar, e.g.
-some values performed better than others with a spread of unique results.
-These results will be added in a final version.
 
-The results are given in Table~\ref{tab:short:runs}.
-The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value.
+The results are given in Table~\ref{tab:short:runs}. The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value.
 We separately report the total number of problems solved and those solved uniquely (by a single strategy across all experiments).
 
-The best combination of options overall was X. In general, strategies that converged performed better/worse than those that diverged.
-
-In total X out of all Y strategies contributed some unique solutions. This shows that these new options are complementary in general.
+The best combination of options overall was decaying from an initial age/weight of 1:100 with frequency 1000.
+However, in general strategies that converged performed better than those that decayed, both in terms of total problems solved and in terms of unique solutions.
+In total, 28 out of a total of 36 strategies contributed some unique solutions. This shows that these new options are complementary in general.
 
@@
 
-Our next experiment aims to answer the question ``how much can the portfolio mode of Vampire improve using these new options''.
-To address this question we use the portfolio mode used in the most recent CASC competition~\cite{CADE18} as a baseline and run the new options on top of this.
-Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio e.g.
-finite model building~\cite{DBLP:conf/sat/Reger0V16}.
+Our next experiment aims to answer the question ``How much can the portfolio mode of Vampire improve using these new options?''.
+To address this we use the portfolio mode used in the most recent CASC competition~\cite{CADE18} as a baseline and run the new options on top of this.
+Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio e.g. finite model building~\cite{fmb}.
 
@@
 Therefore, it makes sense to consider the union of all experiments ran.
-%For experimental purposes, the same environment was preserved with the exception of activating the new ``shapes'' settings with a range of different shape frequencies.
 
@@
 In total, 134 problems were solved by the new configurations that were not solved by the baseline.
-This is an impressive result - it is rare to be able to improve portfolio mode by this many new problems with a single new proof search option.
+This is an impressive result --- it is rare to be able to improve portfolio mode by this many new problems with a single new proof search option.
 
-We also report the \emph{u-score} computed by giving $1/n$ points per problem solved where $n$ is the number of strategies solving a problem.
+We also report the \emph{u-score} computed by giving $1/n$ points per problem solved where $n$ is the number of strategies solving a problem~\cite{u-score}.
 This gives a measure of contribution per strategy.

### 2019-03-03 papers/CADE19/paper.tex
spell check, Ahmed's comments

@@
 
-We note here that Vampire's portfolio mode includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of nongoal clauses, thus preferring clauses in or derived from the goal in clause selection.
+We note here that Vampire's portfolio mode includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of non-goal clauses, thus preferring clauses in or derived from the goal in clause selection.
 Use of this heuristic is orthogonal to the age/weight ratio and we do not investigate its impact any further here.
@@
 Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance.
-The drop in performance is the how many times more activations were required for a proof in the worst/average case compared to the best.
+The drop in performance is how many times more activations were required for a proof in the worst/average case compared to the best.
 No AWR value solves all problems, with the best being 1:5.
@@
 
-We introduce two new options: \texttt{--age\_weight\_ratio\_shape} can take the values \emph{constant}, \emph{decay}, or \emph{converge} and selects one of the above shapes; and \texttt{--age\_weight\_ratio\_shape\_frequency} specifies the frequency (rate) or convergance/decay (default is 100).
+We introduce two new options: \texttt{--age\_weight\_ratio\_shape} can take the values \emph{constant}, \emph{decay}, or \emph{converge} and selects one of the above shapes; and \texttt{--age\_weight\_ratio\_shape\_frequency} specifies the frequency (rate) or convergence/decay (default is 100).
 These are used with the existing \texttt{--age\_weight\_ratio} option (default 1:1) to give a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.

### 2019-03-03 papers/CADE19/paper.tex
extra comments

@@
 The results are given in Table~\ref{tab:short:runs}. The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value.
-We separately report the total number of problems solved and those solved uniquely (by a single strategy across all experiments).
+We separately report the total number of problems solved and those solved uniquely (by a single strategy across all experiments, including the baseline).
 
+In all cases varying the AWR over time outperformed keeping it constant.
+As the new strategies always include some reasoning using \verb|1:1| AWR this suggests that all benefit from \emph{some} fair reasoning during proof search and some biased reasoning.
 The best combination of options overall was decaying from an initial age/weight of 1:100 with frequency 1000.
-However, in general strategies that converged performed better than those that decayed, both in terms of total problems solved and in terms of unique solutions.
+However, in general, strategies that converged performed better than those that decayed, both in terms of total problems solved and in terms of unique solutions. 
+Longer frequencies tended to do better, suggesting that more time at the intermediate AWR values is preferable. 
 In total, 28 out of a total of 36 strategies contributed some unique solutions. This shows that these new options are complementary in general.
@@
 To address this we use the portfolio mode used in the most recent CASC competition~\cite{CADE18} as a baseline and run the new options on top of this.
-Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio e.g. finite model building~\cite{fmb}.
+Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio, e.g.
+finite model building~\cite{fmb}, as well as other options related to clause selection, e.g.
+non-goal weight coefficient and set-of-support.
 

### 2019-03-03 papers/CADE19/paper.tex
numbers

@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparison with other Solvers}
+
+\begin{table}[t]
+\caption{Total number of problems solved compared to other solvers.\label{tab:other}}
+\centering
+\begin{tabular}{l|l|l}
+Solver 		& Total solved & Uniquely solved \\ \hline
+Vampire (old)	& 13,057 & \\
+Vampire (new)   & 13,191 & \\
+E 			&10,845\\
+iProver		& 8,143\\
+CVC4		& 9,354\\
+\end{tabular}
+\end{table}
+
+To place these results in context, we compare the overall number of problems solved by our new strategies with the results of other solvers (using their 2018 CASC competition strategy).
+These results are from 300s and are given in Table~\ref{tab:other}.
+In this table Vampire (old) stands for the 2018 competition version whilst Vampire (new) stands for the union of all problems solved by new options in the previous section.
+The main point we want to make here is that the unique problems found in the previous section remain unique when compared to other competitive solvers.
+
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions and Future Work}

### 2019-03-04 papers/CADE19/paper.tex
add comparison to other solvers

@@
 
-In all cases varying the AWR over time outperformed keeping it constant.
+In all cases, for a given start AWR value, varying the AWR over time outperformed keeping it constant.
 As the new strategies always include some reasoning using \verb|1:1| AWR this suggests that all benefit from \emph{some} fair reasoning during proof search and some biased reasoning.
@@
 Solver 		& Total solved & Uniquely solved \\ \hline
-Vampire (old)	& 13,057 & \\
-Vampire (new)   & 13,191 & \\
-E 			&10,845\\
-iProver		& 8,143\\
-CVC4		& 9,354\\
+Vampire (old)	& 13,057 & -\\
+Vampire (new)   & 13,191 & 54\\
+E 			&10,845 & 190\\
+iProver		& 8,143 & 215\\
+CVC4		& 9,354 & 501\\
 \end{tabular}
@@
 In this table Vampire (old) stands for the 2018 competition version whilst Vampire (new) stands for the union of all problems solved by new options in the previous section.
-The main point we want to make here is that the unique problems found in the previous section remain unique when compared to other competitive solvers.
+Between them, the two versions of Vampire solve 1,030 problems uniquely.
+The main point we want to make here is that 54 of the unique problems found in the previous section remain unique when compared to other competitive solvers.
 

### 2019-03-04 papers/CADE19/data/analyze.py
add data

@@ -1,0 +1,88 @@
+#!/usr/bin/env python
+
+import sys, copy, random
+
+import cPickle as pickle
+import gzip
+
+def read_strats(data_file_names):
+
+  solvers = {}   # maps solvers to expers where each exper is a {solved_problems:status} dict
+  
+  # loading
+  for name in data_file_names:
+    with open(name,"r") as f:
+      for line in f:
+        prob,solver,config,_,time,_,memory,stat = line.split(",")[:8]
+
+        if ".rm" in prob:
+          pass
+
+        solver_name = solver+"_"+config
+
+	if "SMT" in prob:
+		logic = prob.split("/")[1]
+		solver_name = logic+"_"+solver_name
+
+        if solver_name in solvers:
+          exper = solvers[solver_name]
+        else:
+          exper = {}
+          solvers[solver_name] = exper
+            
+        if prob in exper: # if there were multiple runs of the same strategy, conut
+          pass
+        else:
+          exper[prob] = (stat.strip(),time) 
+
+  return solvers
+
+if __name__ == "__main__":
+  
+  solvers = read_strats(["others.csv","info.csv"])
+  if "solver_configuration" in solvers:
+    del solvers["solver_configuration"]
+  
+  m = max({len(k) for k in solvers.keys()}) 
+
+  # map from problem to solvers that solve it
+  solved = {}
+  times = {}
+
+  for (solver,exper) in sorted(solvers.items()):
+    count = 0
+    for (prob,(stat,time)) in exper.items(): 
+      if stat=="Refutation" or stat=="Theorem" or stat=="Unsatisfiable" or stat=="Satisfiable" or stat=="CounterSatisfiable": 
+        count +=1
+        if prob not in solved:
+          solved[prob] = set()
+        solved[prob].add(solver)
+        if prob not in times:
+          times[prob] = {} 
+        times[prob][solver] = time
+    print solver,"\t:",count
+             
+  #sys.exit(0)
+
+  uniques = {}
+  for (prob,solutions) in solved.items():    
+    vnew = 0
+    other = 0
+    for s in solutions:
+      if '_awrs' in s or 'baseline' in s:
+	vnew = 1 
+      else:
+        other = other+1
+    if (vnew+other) == 1: 
+      s = "Vampire (new)"
+      if other==1:
+	s = solutions.pop()
+      if s not in uniques:
+        uniques[s] = set()
+      uniques[s].add(prob)
+
[…]

### 2019-03-04 papers/CADE19/paper.tex
proofread, tidy up

--- 2019-03-03 papers/CADE19/paper.tex
@@
 \usepackage[caption=false]{subfig}
-\usepackage[dvipsnames]{xcolor}
 \usepackage{amssymb}
@@
 
-We restrict our attention to clause selection strategies that alternate between selecting clauses based on age (e.g.
-in a first-in-first-out manner) and weight (e.g.
+We restrict our attention to clause selection strategies that alternate between selecting clauses based on age (i.e.
+in a first-in-first-out manner) and weight (i.e.
 those with the fewest symbols first).
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Comparison with other Solvers}
+
+\begin{table}[t]
+	\caption{
+Total number of problems solved compared to other solvers.
+}
+	\centering
+	\begin{tabular}{l l l}
+	Solver~~ & Total solved~~~ & Uniquely solved \\
+	\hline
+	Vampire (old)	& 13,057 & 0\\
+	Vampire (new)   & 13,191 & 54\\
+	E 		& 10,845 & 190\\
+	iProver		& 8,143 & 215\\
+	CVC4		& 9,354 & 501\\
+	\end{tabular}
+	\label{tab:other}
+\end{table}
+
+To place these results in context, we compare the overall number of problems solved by our new strategies with the results of other solvers, using their 2018 CASC competition mode.
+These results are from 300-second runs in identical conditions and are given in Table~\ref{tab:other}.
+In this table \emph{Vampire (old)} stands for the 2018 competition version whilst \emph{Vampire (new)} stands for the union of all problems solved by new options in the previous section.
+Between them, the two versions of Vampire solve 1,030 problems uniquely.
+54 unique problems found in the previous section remain unique when compared to other competitive solvers.
+
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions and Future Work}

### 2019-05-08 papers/FroCoS19/paper.tex
FroCoS '19 skeleton and introduction

\documentclass[runningheads]{llncs}
\newcommand{\lerna}{\textsc{Lerna}}
\begin{document}
\title{A Neural, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
	Giles Reger
}
\institute{University of Manchester, UK}
\maketitle

\begin{abstract}
	We present the \lerna{} system, a neurally-guided automatic theorem prover for first-order logic with equality.
	The system uses a neural network trained on previous proof search attempts to evaluate subgoals based directly on their structure, and hence bias proof search toward success.
	An existing first-order theorem prover is employed to dispatch easy subgoals and prune branches which cannot be solved.
	Exploration of the search space is asynchronous with respect to both the evaluation network and the existing prover, allowing for efficient batched neural network execution and for natural parallelism within the prover.
\keywords{ATP \and Graph Convolutional Network \and Tableaux \and MCTS}
\end{abstract}

\section{Background}

\subsection{First-Order Logic}
(TODO Giles copy traditional FOL-with-equality background here)

\subsection{Automatic Theorem Provers}
(TODO Giles talk about ATP background/history here)

\subsection{Machine Learning and Theorem Proving}
(TODO talk about the intersection of ATP and ML, in particular with respect to proof search guidance)

TODO The Deep Network Guided Proof Search (DNGPS) system~\cite{DNGPS} should be mentioned.

\section{Introduction}
Recent advances~\cite{graph-cnn,gcn,gcn-relational} in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers~\cite{formula-graph}, suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs.
\begin{enumerate}
	\item Proof state in such systems may be very large, such as in saturation-based provers~\cite{Vampire}, leading to training data which is impractical to learn from and slow to evaluate. The DNGPS system side-stepped this by only evaluating a single clause and the conjecture, reducing the amount of information available for the neural network in exchange for brevity.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than comprehension.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance. Even with the reduced amount of data processed, DNGPS employed a two-phase approach in which the prover ran for some time with network guidance and then for the remainder without guidance.
\end{enumerate}
%
By avoiding these issues with a novel prover architecture and exploring several options to improve efficiency, the \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} system explores a step toward useful neural automatic theorem provers.

\section{Design}
\subsection{Rationale}
\subsection{Architecture}
\subsection{Implementation}

\section{Calculus}
\section{Oracle}

\section{Learning}
TODO should ML results go here?

\subsection{Data Collection and Augmentation}
\subsection{Translation to Graphs}
\subsection{Neural Architecture}

\subsection{Neural Network Throughput}
To measure network throughput, the network evaluated \(\approx\) 500MB of pre-recorded data with various batch sizes --- batch size represents a tradeoff between latency of evaluations and parallelism opportunity for a GPU.
Mean throughput (measured in evaluations/second) for each batch size is shown in Table \ref{table:network-throughput}, representing the total number of tableaux sub-goals evaluated in a single second.
As expected, neural network throughput for this task increases rapidly when allowed to process inputs in batches.
In practice a larger batch size was more helpful for proof search, suggesting that a small amount of latency does not significantly affect search in the long term.

\begin{table}
	\caption{Mean neural network throughput versus batch size: evaluations per second.}
	\centering
	\begin{tabular}{c | c}
		\textbf{Batch Size} & \textbf{Mean Throughput (evals./sec)}\\
		\hline
		1 & 254.2\\
		8 & 887.9\\
		32 & 2,128.3\\
		128 & 3,015.0\\
	\end{tabular}
	\label{table:network-throughput}
\end{table}

\section{Results}
\subsection{Hardware}
All results collected on commodity hardware (TODO talk about specs).

\subsection{Benchmark Results}

Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of Z3 and LeRNA on the M2k dataset.
Z3 ran for a full 10 seconds to establish baseline performance, then as an oracle for 20 milliseconds to determine the number of ``trivial'' problems.
LeRNA ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.

\begin{table}
	\caption{Total theorems proved on the M2k dataset.}
	\centering
	\begin{tabular}{r | c}
		\textbf{Configuration} & \textbf{Theorems Proved}\\
		\hline
		Z3 (10s, as baseline) & 1216\\
		Z3 (20ms, as oracle) & 711\\
		LeRNA, unguided (10s, with oracle) & 969\\
		LeRNA, guided (10s, with oracle) & 1012\\
						      & FIXME there is a bug, these ought be better\\
	\end{tabular}
	\label{table:m2k-results}
\end{table}

\section{Conclusion}
\section{Future Work}

\bibliographystyle{splncs04}
\bibliography{references}
\end{document}

### 2019-05-08 papers/FroCoS19/paper.tex
future work section

@@
 \newcommand{\lerna}{\textsc{Lerna}}
+\newcommand{\z}[1]{\textsc{Z3}}
 \begin{document}
@@
 
-Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of Z3 and LeRNA on the M2k dataset.
-Z3 ran for a full 10 seconds to establish baseline performance, then as an oracle for 20 milliseconds to determine the number of ``trivial'' problems.
+Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and LeRNA on the M2k dataset.
+\z3 ran for a full 10 seconds to establish baseline performance, then as an oracle for 20 milliseconds to determine the number of ``trivial'' problems.
 LeRNA ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.
@@
 		\hline
-		Z3 (10s, as baseline) & 1216\\
-		Z3 (20ms, as oracle) & 711\\
+		\z3 (10s, as baseline) & 1216\\
+		\z3 (20ms, as oracle) & 711\\
 		LeRNA, unguided (10s, with oracle) & 969\\
@@
 \section{Future Work}
+The style of theorem prover described opens many new directions for future work and improvements to the system.
+As \lerna{} is a very new system, there is likely much to be gained by simple engineering and tuning: for example, the UCT exploration parameter \(c\) has been left at its theoretical optimum value \(\sqrt{2}\), but it is likely that a higher value will account for neural network inaccuracies and hence improve performance.
+Training on, benchmarking with, and optimising for other datasets (such as the famous TPTP benchmark~\cite{TPTP}) is also left as future work.
 
+\subsection{Proof Search}
+\lerna{} is well-suited for long-term proof search attempts in mathematics, such as those employed in the AIM project~\cite{AIM}: search is stable over time and does not produce a combinatorial explosion in the same way that some traditional systems tend to after a short period.
+Additionally, the amount of information (``confidence'') in the system grows over time, as a result of a growing number of oracle invocations and neural network evaluations.
+Proof search can in principle be manually inspected more easily than in saturation-based provers to examine promising subgoals and remove known falsehoods from the search space.
+The authors hope to explore applying the system to this interesting domain.
+
+Another future direction for proof search is a principled incomplete mode, in which branches deemed sufficiently uninteresting by the heuristic are pruned, perhaps in response to time or memory constraints as seen in limited resource strategies~\cite{LRS}.
+This approach, while clearly incomplete, would significantly accelerate proof search in the direction of more promising search within the available resources.
+
+\subsection{Prover Calculus}
+The calculus currently employed is deliberately na\"ive, so much work is planned here.
+In particular, the simplification routines can be improved to remove more trivial sub-formulae.
+While in general the oracles' pre-processing will remove these, they serve as noise for the neural network and might also increase the number of inference steps required to reach a proof.
+
+As one possible view of this approach is as an intelligent pre-processor for an existing efficient ATP, a greater number of aggressive and/or weakening inferences might be included in the calculus.
+It is well-known (TODO find citation) that \emph{prenexing} (or conversely \emph{miniscoping}) formulae can have a significant effect on proof search for some theorem provers, so including suitable quantifier-manipulation rules might prove to be a useful extension.
+
+Generalising further, ideas from other refutation-tableaux calculi could well be suitable for this system.
+In particular, the authors are attempting to integrate an adapted connection rule from the non-clausal connection calculus~\cite{non-clausal-connections}, as used in nanoCoP~\cite{nanocop}, in order to reduce the number of proof steps required to instantiate universal quantifiers.
+
+Finally, this prover architecture can support other logics without excessive modification.
+Given that \z3 is already capable of supporting many \emph{theories}, such as arithmetic or datatypes, a many-sorted first order logic such as those described by SMTLIB~\cite{smtlib} or the TFF0 dialect of TPTP~\cite{tff0} seems an appropriate first step.
+
+\subsection{Oracle}
+While \z3 is a strong theorem prover in its own right and performs well here, it remains to be seen if it is the best for this application.
+Other ATPs (or indeed counter-example-finding systems) should be tested as it involves minimal engineering effort.
+A \emph{portfolio} of several oracle systems working in tandem might also be considered, although of course this will eventually retard proof search linearly in the number of systems present.
+
+Reducing the number of oracle invocations is another area for optimisation.
+Currently, the system calls an oracle for every new subgoal generated.
+It seems unlikely that the subgoal is materially easier to dispatch than its parent (especially in the case of propositional inferences that do not split the goal), so heuristically or probabilistically removing such subgoals from the oracle's queue is a possible area for improvement.
+
+\lerna{} does not currently use any information from the oracle beyond its status: using auxiliary information such as satisfying models or unused formulae could well aid proof search.
+
+\subsection{Machine-Learned Heuristic}
+Many other graph-based neural architectures are possible.
+PyTorch Geometric alone currently includes nearly 40 other graph-specific neural layers pre-programmed from the literature.
+Neural models specifically for theorem proving are relatively under-studied.
+To combat this, data used for this paper will be published in the near future so that the machine-learning community can improve upon our simple models.
+
+Different approaches to formula-to-graph translation, symbol embeddings, data augmentation, and model integration may also be explored.
 \bibliographystyle{splncs04}

### 2019-05-09 papers/FroCoS19/paper.tex
Oracle section, started on learning

@@
 \documentclass[runningheads]{llncs}
+\usepackage{graphviz}
+\usepackage{hyperref}
 \newcommand{\lerna}{\textsc{Lerna}}
 \newcommand{\z}[1]{\textsc{Z3}}
+\newcommand{\mizarlarge}{\textit{M40k}}
+\newcommand{\mizarsmall}{\textit{M2k}}
 \begin{document}
@@
 \author{
-	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
-	Giles Reger
+	Michael Rawson*\orcidID{0000-0001-7834-1567}\inst{1} \and
+	Giles Reger\inst{2}
 }
-\institute{University of Manchester, UK}
+\institute{
+	University of Manchester, UK\\
+	\email{michael@rawsons.uk}\and
+	University of Manchester, UK
+}
 \maketitle
@@
 \subsection{First-Order Logic}
-(TODO Giles copy traditional FOL-with-equality background here)
+(TODO copy Giles' traditional FOL-with-equality background here)
 
 \subsection{Automatic Theorem Provers}
-(TODO Giles talk about ATP background/history here)
+(TODO copy Giles' ATP background/history here)
 
@@
 
-TODO The Deep Network Guided Proof Search (DNGPS) system~\cite{DNGPS} should be mentioned.
+TODO The Deep Network Guided Proof Search (DNGPS) system~\cite{DNGPS} should be mentioned, as should rlCoP~\cite{rlCoP}
+
+\subsection{The MPTP Problem Set}
+For training and evaluation purposes a set of problems exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
+Urban et al.~\cite{rlCoP} took a subset\footnote{\url{https://github.com/JUrban/deepmath/blob/master/M2k_list}} of the large \mizarlarge{} problem set (containing 32,524 problems) and called it \mizarsmall{} (containing 2004 problems).
 
@@
 Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} system explores a step toward useful neural automatic theorem provers.
-
 \section{Design}
+\label{section:Design}
 \subsection{Rationale}
@@
 \section{Calculus}
+
 \section{Oracle}
+\label{section:oracle}
+One problem with the calculus as described is that proofs can be quite lengthy, even if the goal is relatively trivial.
+To rectify the situation, new goals generated by ongoing proof search are enqueued for attempted proof by an existing \emph{oracle} ATP system, as described in Section \ref{section:Design}.
 
-\section{Learning}
-TODO should ML results go here?
+\subsection{Z3}
+\z3{}\cite{Z3} is an efficient and mature SMT solver, supporting a core propositional logic extended with ``theories'' such as arithmetic and datatypes.
+It also notably supports first-order logic via a combination of decision procedures for decidable fragments (such as the Bernays-Sch\"onfinkel class of formulae), and heuristic quantifier instantiation routines~\cite{quantifier-instantiation}.
+\z3{} is attractive for this application due to its low startup times and its ability to produce both satisfiable and unsatisfiable results.
 
-\subsection{Data Collection and Augmentation}
+In the \lerna{} system, we ran \z3{} as an external program (thus reducing the coupling to any particular ATP, although \z3{} has an API which might improve efficiency) with its Model-Based Quantifier Instantiation heuristic for 20 milliseconds.
+This was chosen as the shortest time in which the oracle can dispatch a reasonable amount of trivial goals (and in fact \z3{} is so strong it dispatches some goals immediately: see Section \ref{section:results}).
+Longer oracle runtimes might produce better performance in future, but for this experiment longer runtimes begin to conflate the performance of the oracle and the performance of the system as a whole.
+This application is unusual for ATP systems --- very short runtimes, and a mix of true and false problem statements.
+Running in this setting also amounts to fuzz-testing: in development of this system a bug was rediscovered in \z3{}\footnote{\url{https://github.com/Z3Prover/z3/issues/2101}} which resulted in non-termination of the prover: happily, the bug was already fixed in a newer version.
+
+\subsection{Acting as a Preprocessor}
+\lerna{} as a whole might also be seen as an intelligent pre-processor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
+The system can therefore act as an adaptor for any existing first-order ATP, adding parallelism opportunities and ``smoothing out'' sensitivity to input syntax.
+
+\section{Learned Heuristic}
+A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
+Although the data is collected by running the system itself (and therefore might be considered \emph{reinforcement} learning~\cite{reinforcement}), for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.
+
+\subsection{Data Collection}
+A large dataset of satisfiable and unsatisfiable goals were collected by running the unguided prover on the \mizarlarge{} dataset for 10 seconds.
+As soon as the prover determines the satisfiability of any sub-goal, the formula it represents and its status is recorded.
+This resulted in 18,340 unsatisfiable examples and 1,845,267 satisfiable examples, occupying 6GB of disk space.
+The dataset is very imbalanced (due to a combination of weakening rules producing a large number of trivially-satisfiable examples, and to prover termination after after the goal is shown to be unsatisfiable), at a ratio of around 100:1.
+
 \subsection{Translation to Graphs}
+Wang et al.~\cite{formula-graph} give a translation from higher-order formulae to directed graphs, and a similar scheme is used here.
+Constants, function symbols, predicate symbols, and bound variables are given their own node.
+Application of functions and predicates to arguments are represented as an ``application node'' with two children: the symbol node and an ``argument list'' node representing the list of arguments.
+Propositional connectives and equality have the obvious representation, while quantifiers have two children: the variable they bind and their sub-formula.
+Common sub-graphs are merged.
+
+An example formula's translation from
+\[
+	\forall X, \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]
+\]
+is shown in figure...
+
+\begin{figure}
+\end{figure}
+
+\subsection{Augmentation}
 \subsection{Neural Architecture}
@@
 \section{Results}
+\label{section:results}
 \subsection{Hardware}
@@
 
-\section{Conclusion}
+\section{Conclusions}
 \section{Future Work}
@@
 
-As one possible view of this approach is as an intelligent pre-processor for an existing efficient ATP, a greater number of aggressive and/or weakening inferences might be included in the calculus.
-It is well-known (TODO find citation) that \emph{prenexing} (or conversely \emph{miniscoping}) formulae can have a significant effect on proof search for some theorem provers, so including suitable quantifier-manipulation rules might prove to be a useful extension.
+As one possible view of this approach is as an intelligent pre-processor (as mentioned in Section \ref{section:oracle}) for an existing efficient ATP, a greater number of aggressive and/or weakening inferences might be included in the calculus.
+For instance, \emph{prenexing} (or conversely \emph{miniscoping}) formulae can have a significant effect on proof search~\cite{clausification} for some theorem provers, so including suitable quantifier-manipulation rules might prove to be a useful extension.
 
@@
 Different approaches to formula-to-graph translation, symbol embeddings, data augmentation, and model integration may also be explored.
+
+\section{Acknowledgements}
+The authors wish to thank Josef Urban and his group in \v{C}VUT, Prague for their help and encouragement with early iterations of this work and for supplying the Mizar dataset used in this paper.
+
 \bibliographystyle{splncs04}

### 2019-05-10 papers/FroCoS19/paper.tex
added ML section

@@
 \documentclass[runningheads]{llncs}
-\usepackage{graphviz}
+\usepackage{graphicx}
 \usepackage{hyperref}
+\usepackage{subcaption}
 \newcommand{\lerna}{\textsc{Lerna}}
@@
 This resulted in 18,340 unsatisfiable examples and 1,845,267 satisfiable examples, occupying 6GB of disk space.
-The dataset is very imbalanced (due to a combination of weakening rules producing a large number of trivially-satisfiable examples, and to prover termination after after the goal is shown to be unsatisfiable), at a ratio of around 100:1.
+The dataset is very imbalanced (due to a combination of weakening rules producing a large number of trivially-satisfiable examples, and to prover termination after the goal is shown to be unsatisfiable), at a ratio of around 100:1.
 
@@
 Propositional connectives and equality have the obvious representation, while quantifiers have two children: the variable they bind and their sub-formula.
-Common sub-graphs are merged.
+
+To produce an input graph from a formula \(F\), the formula is first parsed into an abstract syntax tree.
+Common sub-trees up to \(\alpha\)-equivalence are merged, then the resulting directed acyclic graph has any named-symbol nodes replaced with an opaque, nameless label such as "predicate" or "variable" --- since distinct symbols remain as distinct nodes under this scheme, no information is lost other than the natural-language semantics of the symbol name.
+In practice, undirected graphs improved model performance so the graph is made undirected before encoding node labels as one-hot inputs to produce the final input graphs.
 
@@
 \[
-	\forall X, \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]
+	F = \forall X, \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]
 \]
-is shown in figure...
+is shown in Figure \ref{figure:translation}.
 
 \begin{figure}
+	\centering
+	\begin{subfigure}{.4\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{translation-ast}
+		\caption{The AST for \(F\).}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{translation-graph}
+		\caption{De-duplicated DAG.}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{translation-nameless}
+		\caption{Nameless DAG.}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{translation-undirected}
+		\caption{Final result.}
+	\end{subfigure}
+	\caption{The translation process from \(F\) to a graph as seen by the neural network.}
+	\label{figure:translation}
 \end{figure}
@@
 \subsection{Augmentation}
+One possible solution~\cite{imbalanced} to the problem of classification on imbalanced domains is to synthesize new data for under-represented classes --- in this case unsatisfiable formulae --- from existing data by augmenting it.
+A classic example is augmenting image data by cropping, flipping or adding noise to existing images.
+
+There are many possible ways to augment formulae graphs.
+For this prototype, a simple approach is taken in which a small number of nonsense formulae are added to the graph by randomly adding nodes/edges where appropriate.
+This approach has the advantage of exposing the network to ``noise'' such as additional axioms which might well occur in practice, but if the network is adequately capable of filtering these then no new formulae are actually seen.
+
 \subsection{Neural Architecture}
+TODO set fixed architecture and report on this.
 
-\subsection{Neural Network Throughput}
-To measure network throughput, the network evaluated \(\approx\) 500MB of pre-recorded data with various batch sizes --- batch size represents a tradeoff between latency of evaluations and parallelism opportunity for a GPU.
-Mean throughput (measured in evaluations/second) for each batch size is shown in Table \ref{table:network-throughput}, representing the total number of tableaux sub-goals evaluated in a single second.
-As expected, neural network throughput for this task increases rapidly when allowed to process inputs in batches.
-In practice a larger batch size was more helpful for proof search, suggesting that a small amount of latency does not significantly affect search in the long term.
-
-\begin{table}
-	\caption{Mean neural network throughput versus batch size: evaluations per second.}
-	\centering
-	\begin{tabular}{c | c}
-		\textbf{Batch Size} & \textbf{Mean Throughput (evals./sec)}\\
-		\hline
-		1 & 254.2\\
-		8 & 887.9\\
-		32 & 2,128.3\\
-		128 & 3,015.0\\
-	\end{tabular}
-	\label{table:network-throughput}
-\end{table}
+\subsection{Implementation}
+TODO discuss implementation with PyTorch~\cite{pytorch} and its graph-processing extension PyTorch Geometric~\cite{pytorch-geometric}.
 

### 2019-05-12 papers/FroCoS19/paper.tex
added figure source and PDF renders

@@
 \usepackage{subcaption}
+\usepackage{textcomp}
 \newcommand{\lerna}{\textsc{Lerna}}
@@
 	\email{michael@rawsons.uk}\and
-	University of Manchester, UK
+	University of Manchester, UK\\
+	\email{regerg@cs.man.ac.uk}
 }
@@
 \begin{abstract}
-	We present the \lerna{} system, a neurally-guided automatic theorem prover for first-order logic with equality.
-	The system uses a neural network trained on previous proof search attempts to evaluate subgoals based directly on their structure, and hence bias proof search toward success.
+	We present a prototype of a neurally-guided automatic theorem prover for first-order logic with equality.
+	The prototype uses a neural network trained on previous proof search attempts to evaluate subgoals based directly on their structure, and hence bias proof search toward success.
 	An existing first-order theorem prover is employed to dispatch easy subgoals and prune branches which cannot be solved.
 	Exploration of the search space is asynchronous with respect to both the evaluation network and the existing prover, allowing for efficient batched neural network execution and for natural parallelism within the prover.
+	Evaluation on the MPTP dataset shows that the prover can improve with learning.
 \keywords{ATP \and Graph Convolutional Network \and Tableaux \and MCTS}
@@
 
+\section{Introduction}
+Recent advances~\cite{graph-cnn,gcn,gcn-relational} in neural network systems allow for processing graph-structured data in a neural context.
+Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs)~\cite{formula-graph}, suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
+Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
+By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
+However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs.
+Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
+
+\begin{enumerate}
+	\item Proof state in such systems may be very large, such as in saturation-based provers~\cite{Vampire}, leading to training data which is impractical to learn from and slow to evaluate.
+side-stepped this by only evaluating a single clause at a time (instead of large set) and the conjecture, reducing the amount of information available for the neural network in exchange for brevity.
+	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
+	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
+Even with the reduced amount of data processed, DNGPS employed a two-phase approach in which the prover ran for some time with network guidance and then for the remainder without guidance.
+\end{enumerate}
+%
+Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures.
+Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.
+
 \section{Background}
-
 \subsection{First-Order Logic}
@@
 
-\section{Introduction}
-Recent advances~\cite{graph-cnn,gcn,gcn-relational} in neural network systems allow for processing graph-structured data in a neural context.
-Graphs are a natural representation for logical formulae as found in automatic theorem provers~\cite{formula-graph}, suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
-Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
-By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
-However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs.
-\begin{enumerate}
-	\item Proof state in such systems may be very large, such as in saturation-based provers~\cite{Vampire}, leading to training data which is impractical to learn from and slow to evaluate.
-The DNGPS system side-stepped this by only evaluating a single clause and the conjecture, reducing the amount of information available for the neural network in exchange for brevity.
-	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than comprehension.
-	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
-Even with the reduced amount of data processed, DNGPS employed a two-phase approach in which the prover ran for some time with network guidance and then for the remainder without guidance.
-\end{enumerate}
-%
-By avoiding these issues with a novel prover architecture and exploring several options to improve efficiency, the \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures.
-Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} system explores a step toward useful neural automatic theorem provers.
 \section{Design}
@@
 This resulted in 18,340 unsatisfiable examples and 1,845,267 satisfiable examples, occupying 6GB of disk space.
-The dataset is very imbalanced (due to a combination of weakening rules producing a large number of trivially-satisfiable examples, and to prover termination after the goal is shown to be unsatisfiable), at a ratio of around 100:1.
+The dataset is very imbalanced (due to a combination of weakening rules producing a large number of trivially-satisfiable examples, and to immediate prover termination after the goal is shown to be unsatisfiable), at a ratio of around 100:1.
 
@@
 In practice, undirected graphs improved model performance so the graph is made undirected before encoding node labels as one-hot inputs to produce the final input graphs.
-
-An example formula's translation from
-\[
-	F = \forall X, \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]
-\]
-is shown in Figure \ref{figure:translation}.
+An example formula's translation is shown in Figure \ref{figure:translation}.
 
@@
 	\centering
-	\begin{subfigure}{.4\textwidth}
+	\begin{subfigure}{.5\textwidth}
 		\centering
 		\includegraphics[width=\linewidth]{translation-ast}
-		\caption{The AST for \(F\).}
+		\caption{AST.}
 	\end{subfigure}
-	\begin{subfigure}{.3\textwidth}
+	\begin{subfigure}{.39\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\begin{subfigure}{.3\textwidth}
+	\begin{subfigure}{.4\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\begin{subfigure}{.3\textwidth}
+	\begin{subfigure}{.35\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\caption{The translation process from \(F\) to a graph as seen by the neural network.}
+	\caption{The translation process for \(\forall X \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]\) to a graph, as seen by the neural network.}
 	\label{figure:translation}
@@
 \subsection{Augmentation}
+\label{section:augmentation}
 One possible solution~\cite{imbalanced} to the problem of classification on imbalanced domains is to synthesize new data for under-represented classes --- in this case unsatisfiable formulae --- from existing data by augmenting it.
-A classic example is augmenting image data by cropping, flipping or adding noise to existing images.
+An example is augmenting image data by cropping, flipping or adding noise to existing images.
 
@@
 \subsection{Neural Architecture}
-TODO set fixed architecture and report on this.
+Convolutional neural networks~\cite{cnn} are well-known as a method for processing images.
+In a typical convolutional architecture, there are a series of filtering stages, followed by a densely-connected neural network.
+Each filtering stage intuitively combines data from local features (via \emph{convolution}), then reduces the dimensions of the image (via \emph{pooling}) for the next stage.
+Graph neural networks have analagous convolution~\cite{gcn} (combining information from neighbouring nodes) and pooling~\cite{top-k-pooling} (merging nodes to reduce the size of the input graph) operators.
 
-\subsection{Implementation}
-TODO discuss implementation with PyTorch~\cite{pytorch} and its graph-processing extension PyTorch Geometric~\cite{pytorch-geometric}.
+A brief period of experimentation with these operators yielded the following network architecture, shown in Figure \ref{figure:network}.
+It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimize the network parameters.
+The network is shown processing a formula in Figure \ref{figure:computation}.
+\begin{enumerate}
+	\item Input. a graph \(G\) consisting of one-hot encoded nodes \(N\) and edges \(E\).
+	\item Embedding. Each node is mapped to an embedding vector of size 64 via a trained dense embedding.
+	\item Initial Convolution. 4 convolution layers are applied to the graph with rectified linear activations. This yields a graph of the same size, but with information exchanged between nodes.
+	\item Convolution/Pooling.
+Similar convolution layers are then passed through \emph{top-\(k\)}~\cite{top-k-pooling} layers, retaining \(k = 60\%\) of the graph's nodes.
+This is repeated 3 times, reducing the size of the graph considerably.
+	\item Convolution/Max-Pooling. A final convolution layer feeds into a max-pooling layer, combining all remaining node data into one datum, and dropping the edge data.
+	\item Fully Connected. A fully-connected hidden layer with rectified linear activation halves the input size.
+	\item Fully Connected/Softmax. A fully-connected final layer outputs two class labels, with softmax activation.
+\end{enumerate}
+
+To reduce over-fitting, dropout~\cite{dropout} is applied in convolutional and fully-connected layers, \(p = 0.1\).
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{network}
+	\caption{The neural network architecture.
+Initially there are \(N\) nodes, then after pooling there are \(P_1, P_2, P_3\) nodes.
+Node-level embedding layers are shown per-node, graph-level convolutional and pooling layers are shown per-graph.}
+	\label{figure:network}
+\end{figure}
+
+\subsection{Implementation and Training}
+This architecture was implemented with the PyTorch~\cite{pytorch} neural network library, combined with a graph-processing (``geometric'') extension library, PyTorch Geometric~\cite{pytorch-geometric}, which together provide facilities for automatic differentiation, GPU-accelerated training, pre-programmed layers for graph processing, and various utilities.
+
+The dataset is split into a large training set and a smaller test set (200 balanced examples), since unsatisfiable examples were difficult to obtain.
+The training set was then augmented as described in section \ref{section:augmentation} to produce a total training set of 3.5 million examples.
+The network was trained on commodity desktop hardware with a mid-range GPU~\footnote{NVIDIA\textsuperscript{\textregistered} GeForce\textsuperscript{\textregistered} GT 730.} for 8 epochs/24 hours, optimizing a negative log-likelihood loss function.
+
+\begin{figure}
+	\centering
+	\captionsetup[subfigure]{justification=centering}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{embedding}
+		\caption{Output of embedding layer.}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{conv}
+		\caption{Output after initial convolutions.}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{pool0}
+		\caption{After first pooling.}
+	\end{subfigure}
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{pool1}
+		\caption{After second pooling.}
+	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{fc0}
+		\caption{After max-pooling.}
+	\end{subfigure}
+	\caption{Computation in the neural network, showing intermediate values involved in the network (correctly) predicting that an input formula from the test set is unsatisfiable.
+Vectors of floating-point numbers are shown as grids.}
+	\label{figure:computation}
+\end{figure}
+
+\subsection{Evaluation}
+The network was evaluated on the balanced test set of 200 examples, as described.
+Various metrics for accuracy are shown in Table \ref{table:network-evaluation}.
+While these results are very promising, it should be emphasised that it is unclear how effective a train/test split is in this setting (since similar subgoals may occur in both sets, even with proper data hygiene), and that this network is not attempting to determine the satisfiability of arbitrary formulae, merely those that occur in proof attempts on the \mizarlarge{} dataset.
+
+The higher precision and lower recall values are likely an artefact of the augmentation process.
+However, even with these caveats, the network performance is surpising and is practically useful for improving proof search in this dataset.
+
+\begin{table}
+	\caption{Accuracy metrics for the neural heuristic.}
+	\centering
+	\begin{tabular}{r | c}
+		\textbf{Metric} & \textbf{Score}\\
+		\hline
+		Accuracy & 93.0\%\\
+		True Positive&99\\
+		True Negative&87\\
+		False Positive&13\\
+		False Negative&1\\
+		Precision&0.990\\
+		Recall&0.884\\
+		\(F_1\)&0.934\\
+	\end{tabular}
+	\label{table:network-evaluation}
+\end{table}
 
@@
 \label{section:results}
-\subsection{Hardware}
-All results collected on commodity hardware (TODO talk about specs).
+In order to evaluate the effect of heuristic guidance on the existing prover, the prover ran both with and without guidance for 10 seconds on all available CPU cores.
+All results were collected on commodity desktop hardware\footnote{Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-6700 CPU @ 3.40GHz, 16GB RAM.}.
 
 \subsection{Benchmark Results}
-
 Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and LeRNA on the M2k dataset.
@@
 Many other graph-based neural architectures are possible.
-PyTorch Geometric alone currently includes nearly 40 other graph-specific neural layers pre-programmed from the literature.
+PyTorch Geometric alone currently includes nearly 40 other graph-specific neural layers pre-programmed from the literature\footnote{\url{https://rusty1s.github.io/pytorch_geometric/build/html/modules/nn.html}}.
 Neural models specifically for theorem proving are relatively under-studied.
 To combat this, data used for this paper will be published in the near future so that the machine-learning community can improve upon our simple models.
-
 Different approaches to formula-to-graph translation, symbol embeddings, data augmentation, and model integration may also be explored.
@@
 \section{Acknowledgements}
-The authors wish to thank Josef Urban and his group in \v{C}VUT, Prague for their help and encouragement with early iterations of this work and for supplying the Mizar dataset used in this paper.
+The authors wish to thank Josef Urban and his group in \v{C}VUT, Prague for their help and encouragement with early iterations of this work, and for supplying the Mizar dataset used in this paper.
 

### 2019-05-13 papers/FroCoS19/paper.tex
added figure, spellcheck, clean up to 16 pages

@@
 \usepackage{hyperref}
+\usepackage{mathpartir}
 \usepackage{subcaption}
@@
 \section{Introduction}
-Recent advances~\cite{graph-cnn,gcn,gcn-relational} in neural network systems allow for processing graph-structured data in a neural context.
-Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs)~\cite{formula-graph}, suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
-Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
-By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
-However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs.
-Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
-
+Recent advances~ in neural network systems allow for processing graph-structured data in a neural context.
+Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
+However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
 \begin{enumerate}
-	\item Proof state in such systems may be very large, such as in saturation-based provers~\cite{Vampire}, leading to training data which is impractical to learn from and slow to evaluate.
-side-stepped this by only evaluating a single clause at a time (instead of large set) and the conjecture, reducing the amount of information available for the neural network in exchange for brevity.
+	\item Proof state in such systems may be very large, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
 	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
 	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
-Even with the reduced amount of data processed, DNGPS employed a two-phase approach in which the prover ran for some time with network guidance and then for the remainder without guidance.
 \end{enumerate}
@@
 \section{Background}
-\subsection{First-Order Logic}
-(TODO copy Giles' traditional FOL-with-equality background here)
+An understanding of neural networks~\cite{neural-survey}, and of first-order logic and automatic theorem provers~\cite{fol-atp-survey} is assumed.
+The particular syntax of first-order logic used in this work is presented here as clarification.
 
-\subsection{Automatic Theorem Provers}
-(TODO copy Giles' ATP background/history here)
+\subsection{Syntax of First-Order Logic}
+Terms \(t\) are either
+\begin{enumerate}
+	\item Variables \(x\), bound by a quantifier.
+	\item Function applications \(f\left(t_1, t_2, \ldots, t_n\right)\) where \(f\) is an \(n\)-ary function symbol. Constants \(c\) are a degenerate case in which \(n = 0\).
+\end{enumerate}
+%
+Formulae \(\phi, \psi\) are one of
+\begin{enumerate}
+	\item Truth \(\top\) or falsehood \(\bot\).
+	\item Equalities \(t_1 = t_2\).
+	\item Predicate applications \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol.
+	\item Negations \(\lnot \phi\), implications \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
+	\item Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
+	\item Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).
+\end{enumerate}
+
+\subsection{Modern Automatic Theorem Provers}
+State-of-the-art ATP systems (e.g.
+E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}, CVC4~\cite{CVC4}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
+Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
+Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.
 
 \subsection{Machine Learning and Theorem Proving}
-(TODO talk about the intersection of ATP and ML, in particular with respect to proof search guidance)
+Despite the efficiency of such systems, they can still spend time exploring areas that a human mathematician would discard, and tuning such systems manually is in general extremely difficult~\cite{portfolio}.
+There is therefore interest in applying machine-learning techniques to automatic theorem provers, with the eventual aim of an ``intelligent'' theorem prover which can learn from past experience to develop an intuition, discard uninteresting search space, and tune performance in a more principled way.
 
-TODO The Deep Network Guided Proof Search (DNGPS) system~\cite{DNGPS} should be mentioned, as should rlCoP~\cite{rlCoP}
+Work has proceeded in the fields of premise selection~\cite{DeepMath,MaSh,MaLARea,formula-graph}, static strategy selection~\cite{Bridge,MaLeS,E-MaLeS}, dynamic (run-time) strategy selection~\cite{dynamic-strategy-priority} and more recently, direct proof guidance~\cite{MaLeCoP,FEMaLeCoP,rlCoP,ENIGMA,DNGPS}.
+Proof guidance typically involves some form of machine-learned \emph{heuristic} which biases proof search in some way, allowing the prover to avoid parts of the search space deemed uninteresting by the heuristic.
+
+Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
+By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
+Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
+DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.
+rlCoP~\cite{rlCoP} is a connection-based reinforcement-learning prover which is not presently neurally-guided, but takes a similar approach to that taken in this paper and achieves impressive results.
+
+\subsection{Neural Networks for Formulae}
+Neural networks are well-known tools for supervised learning~\cite{neural-survey}, and combined with trainable convolution/pooling operators are suitable for processing large-scale data such as images~\cite{cnn}.
+
+Processing structured data such as logical formulae is a relatively new domain for neural networks.
+Some work attempts to use unstructured representations of such formulae, such as text, or build entirely-new models for a specific logic~\cite{logical-entailment}, whereas others attempt to re-use neural techniques for generic structures such as trees~\cite{logical-tree}.
+A promising direction in this area is recent research on neural methods working with graphs~\cite{graph-cnn,gcn,gcn-relational}, which have already been applied to premise selection~\cite{formula-graph}.
+Graph neural networks tend to include network layers inspired by convolution operators in image-processing networks, combining information from neighbouring nodes (pixels).
+For example, the Graph Convolutional Network~\cite{gcn} includes the following convolution layer, where \(H_i\) is the network output at layer \(i\), \(\bar{A}\) is the self-connected adjacency matrix, \(\bar{D}\) is the diagonal node degree matrix of \(\bar{A}\), \(W_i\) is a trainable weight matrix and \(\sigma\) is a non-linear activation function:
+\[
+	H_{i + 1} = \sigma\left(\bar{D}^{-\frac{1}{2}}\bar{A}\bar{D}^{-\frac{1}{2}}H_iW_i\right)
+\]
 
@@
 \label{section:Design}
-\subsection{Rationale}
+In order to achieve the goal of a neural theorem prover without the disadvantages associated with neural approaches, a new design of theorem prover is required.
+Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
+For such a system, we require the following from the calculus:
+\begin{enumerate}
+        \item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of computation and resources.
+Saturation-based provers can have very large proof states, for example.
+        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms tend to operate more efficiently in batches.
+Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
+        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space, without e.g.
+unifiers.
+Otherwise, the learning system is trying to learn while blind to the context of the search.
+	\item \emph{Subgoals must be intelligible.} Adding ``noise'' such as clausification obscures the original intuition behind a goal, at least for human observers.
+While this is not necessarily the case for machine-learning algorithms, it seems likely that removing structure and adding artefacts will reduce model performance.
+\end{enumerate}
+We therefore implement a refutation prover based on a first-order tableaux calculus without unification, on non-clausal formulae.
+Each goal in this case is the set of formulae present on the tableau branch.
+In this context, proof state is small (only the current branch), evaluation of states is possible in parallel, each branch is independent and contains all information required, and all available structure from the original problem is kept.
+
+\subsection{Search}
+In the calculus (see Section \ref{section:Calculus}) for this prover, there are two branching factors: each goal has a set of possible inferences, and each inference contains a set of possible sub-goals.
+To prove a goal, at least one inference must be proved.
+To prove an inference, all the inferences' sub-goals must be proved.
+A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.6\linewidth]{search}
+	\caption{Search in the \lerna{} system, showing shared sub-goals.}
+\end{figure}
+
+Now the search graph can be explored: in each step, a leaf (goal) node is selected for expansion, and all resulting inferences and sub-goals are added to the graph.
+If a goal has no possible inferences, it is satisfiable and can be removed from the search space.
+On the other hand, if a goal is trivial (i.e. contains a contradiction), it is unsatisfiable and can be marked as proven.
+This idea can be lifted to inferences: if an inference contains any satisfiable sub-goal, it too is satisfiable, whereas if an inference contains all unsatisfiable goals, it is unsatisfiable.
+Proof search continues until the timeout is reached or the root goal is shown to be (un-)satisfiable.
+In order to dispatch trivial sub-goals quickly, an existing fast \emph{oracle} ATP is used (see Section \ref{section:oracle}).
+This may mark goals as (un-)satisfiable, at which point no further exploration is required.
+
+Search is biased by heuristic evaluation.
+The neural heuristic function (see Section \ref{section:heuristic}) evaluates each goal and assigns a score corresponding to whether the network believes that the goal is satisfiable or unsatisfiable.
+In order to balance exploitation of promising directions and exploration of all parts of the search space, a principled UCT-based search algorithm is used, as in MonteCoP~\cite{MonteCoP}.
+At each sub-goal, the prover chooses the inference maximising the expression
+\[
+	s + c\sqrt{\frac{\ln N}{n}}
+\]
+where \(s\) is the score associated with the inference, \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)), \(N\) is the number of visits to the current node, and \(n\) is the number of visits to the child.
+After an inference is selected, the least-promising sub-goal is selected, and the selection process continues until a leaf node is reached.
+The score \(s\) of a leaf node is the score assigned to it by the heuristic function.
+An inference's score is the minimum of their sub-goals' scores, and non-leaf sub-goal's score is the maximum of their inference's scores.
+
 \subsection{Architecture}
+The system aims to consume all available CPU and GPU resources as efficiently as possible.
+To that end, proof search is asynchronous: the search algorithm generates new sub-goals, which are placed on two separate queues: one for the oracle ATP, another for heuristic evaluation.
+Proof search then continues elsewhere, while the oracle ATP is called in parallel on each sub-goal (consuming all available CPU) while the heuristic consumes batches of subgoals, efficiently utilising the available GPU resource.
+As information flows backwards from these processes, the search process updates its information about a given sub-goal and propagates that information upwards to the sub-goal's parent inferences, to influence future proof search: see Figure \ref{figure:architecture}.
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{architecture}
+	\caption{A cartoon showing information flow in the system.}
+	\label{figure:architecture}
+\end{figure}
+
 \subsection{Implementation}
+The Rust~\cite{rust} programming language was chosen to implement the main prover (minus the heuristic), for its memory and thread-safety guarantees, as well as its support for managing memory allocation and algebraic datatypes.
+The implementation is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
+Python 3~\cite{python3} was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
+The heuristic is implemented as a server, communicating with the main prover via a TCP socket.
+In principle this allows for the heuristic to be a shared resource with a centralised heuristic server, or a load-balanced cluster.
 
 \section{Calculus}
+\label{section:Calculus}
+The proof calculus used in this context may be extremely general: in fact, any function from goals to a finite set of possible inferences (themselves finite sets of sub-goals) will suffice, as long as each goal remains independent of any other such that the heuristic function can process all available information.
+If the inference system is complete, there are no additional constraints such as orderings or fairness to ensure the completeness of the prover, as the balanced search algorithm (see Section \ref{section:Design}) will ensure this.
+
+\lerna{} presently implements a refutation tableaux calculus~\cite{handbook-of-automated-reasoning} without unification.
+The calculus described is deliberately na\"ive in order to easily satisfy the design constraints described, but this is a weak point of this work and will be improved in future.
+A na\"ive calculus is not necessarily a problem in this context, as the heuristic should in principle select promising areas to explore and ignore uninteresting sub-goals.
+However, a more efficient calculus would improve performance where the heuristic fails.
+
+\subsection{Refutation Tableaux}
+In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
+A set of inference rules of the form
+\begin{mathpar}
+	\inferrule{\Gamma_1 \\ \Gamma_2 \\ \ldots \\ \Gamma_n}{\Delta}
+\end{mathpar}
+where \(\Gamma_i, \Delta\) are sets of formulae and \(\lnot\left(\Gamma_1 \wedge \Gamma_2 \wedge \ldots \Gamma_n\right) \Rightarrow \lnot \Delta\) is an unconditional tautology, form a refutation calculus.
+Proofs in this calculus can be expressed by closed trees of inference rules.
+
+\subsection{Complete Inferences}
+The inference rules in Figure \ref{figure:complete-inferences} form a complete inference system, by analogy with a first-order tableaux calculus without unification.
+A difference and point of interest is the rule for instantiating universal quantifiers: instead of instantiating a variable with any possible term \(t\) --- an infinite space --- it is instantiated with one function symbol (or constant) at a time, quantifying over new variables as needed.
+This allows for instantiating any term over multiple inference rules (effectively enumerating the Herbrand universe for the goal), but without an infinite number of possible inferences at any point.
+Equality is handled by a rule rewriting classes of equal ground terms.
+Both of these rules are complete yet inefficient, but both are likely to be used only a few times in order to provide enough of a ``hint'' to the oracle system for it to find a proof.
+
+\begin{figure}
+	\centering
+	\begin{mathpar}
+		\inferrule[contradiction]{ }{\phi, \lnot \phi, \Gamma}\and
+		\inferrule[equal]{t = s, \phi\left[t/s\right], \Gamma}{t = s, \phi, \Gamma}\and
+		\inferrule[implies]{\lnot \phi, \psi, \Gamma}{\phi \Rightarrow \psi, \Gamma}\and
+		\inferrule[equivalent]{\lnot \phi, \lnot \psi, \Gamma\\\phi, \psi, \Gamma}{\phi \equiv \psi, \Gamma}\and
+		\inferrule[conjunction]{\phi_1, \phi_2, \ldots, \phi_n, \Gamma}{\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n, \Gamma}\and
+		\inferrule[disjunction]{\phi_1, \Gamma \\ \phi_2, \Gamma \\ \ldots \\ \phi_n, \Gamma}{\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n, \Gamma}\and
+		\inferrule*[lab={instantiation}]{\forall x_1, x_2, \ldots x_n. \phi[f(x_1, x_2, \ldots x_n)/x], \forall x. \phi, \Gamma}{\forall x. \phi, \Gamma}\and
+		\inferrule[non-empty]{\phi[k/x], \Gamma}{\forall x. \phi, \Gamma}\and
+		\inferrule[exists]{\phi[k/x], \Gamma}{\exists x. \phi, \Gamma}
+	\end{mathpar}
+	\caption{A complete inference system for \lerna{}.
+Rules for negation are as usual and not shown here for brevity.
+In rule \textsc{instantiation}, \(f\) is a function symbol of arity \(n\) in the conclusion's signature and \(x_1 \ldots x_n\) are fresh for the conclusion.
+In rules \textsc{non-empty} and \textsc{exists}, \(k\) is fresh for the conclusion.
+\(\phi[t/s]\) is a capture-avoiding substitution replacing \(t\) for \(s\) in \(\phi\).}
+	\label{figure:complete-inferences}
+\end{figure}
+
+\subsection{Weakening}
+A weakening rule is an important part of \lerna{}'s calculus, since the \textsc{instantiation} and \textsc{equal} rule can produce a large number of formulae, some of which must be removed to help the oracle to prove the goal.
+Each application of the rule removes some amount of information from the goal in order to simplify it --- this is sound and corresponds to removing an axiom from proof search.
+The rule is merely
+\begin{mathpar}
+	\inferrule[weaken]{\Gamma}{\phi, \Gamma}
+\end{mathpar}
+
+\subsection{Simplifications}
+Before each inferred goal is added to proof search, it is simplified with a simplification function mapping a goal to a possibly-simpler goal.
+This allows for efficient removal of tedious inferences such as double-negation elimination and generally reduces the search space.
+A set of example simplification rules are shown in figure \ref{figure:simplification-rules}.
+
+\begin{figure}
+	\centering
+	\begin{mathpar}
+		\inferrule[double-negation]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
+		\inferrule[conjunction-associative]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
+		\inferrule[disjunction-propagation]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
+		\inferrule[reflexivity]{\top, \Gamma}{t = t, \Gamma}\and
+		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
+	\end{mathpar}
+	\caption{Some simplification rules implemented in \lerna{}. In rule \textsc{free}, \(x\) is free in \(\phi\). Several other rules are implemented.}
+	\label{figure:simplification-rules}
+\end{figure}
 
@@
 \subsection{Acting as a Preprocessor}
-\lerna{} as a whole might also be seen as an intelligent pre-processor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
-The system can therefore act as an adaptor for any existing first-order ATP, adding parallelism opportunities and ``smoothing out'' sensitivity to input syntax.
+\lerna{} as a whole might also be seen as an intelligent preprocessor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
+The system can therefore act as an adaptor for any existing ATP, adding parallelism opportunities and ``smoothing out'' sensitivity to input syntax.
 
 \section{Learned Heuristic}
+\label{section:heuristic}
 A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
@@
 	\centering
-	\begin{subfigure}{.5\textwidth}
+	\begin{subfigure}{.35\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\begin{subfigure}{.39\textwidth}
-		\centering
-		\includegraphics[width=\linewidth]{translation-graph}
-		\caption{De-duplicated DAG.}
-	\end{subfigure}
-	\begin{subfigure}{.4\textwidth}
+%	\begin{subfigure}{.39\textwidth}
+%%		\centering
+%		\includegraphics[width=\linewidth]{translation-graph}
+%		\caption{De-duplicated DAG.}
+%	\end{subfigure}
+	\begin{subfigure}{.3\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\begin{subfigure}{.35\textwidth}
+	\begin{subfigure}{.25\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\caption{The translation process for \(\forall X \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]\) to a graph, as seen by the neural network.}
+	\caption{The translation process for \(\forall x. \left[p\left(f(c), x\right) \wedge \lnot\left(f(c) = d\right)\right]\) to a graph, as seen by the neural network.}
 	\label{figure:translation}
@@
 \label{section:augmentation}
-One possible solution~\cite{imbalanced} to the problem of classification on imbalanced domains is to synthesize new data for under-represented classes --- in this case unsatisfiable formulae --- from existing data by augmenting it.
+One possible solution~\cite{imbalanced} to the problem of classification on imbalanced domains is to synthesise new data for under-represented classes --- in this case unsatisfiable formulae --- from existing data by augmenting it.
 An example is augmenting image data by cropping, flipping or adding noise to existing images.
-
 There are many possible ways to augment formulae graphs.
@@
 \subsection{Neural Architecture}
-Convolutional neural networks~\cite{cnn} are well-known as a method for processing images.
-In a typical convolutional architecture, there are a series of filtering stages, followed by a densely-connected neural network.
+In a typical convolutional network architecture for images~\cite{cnn}, there are a series of filtering stages, followed by a densely-connected neural network.
 Each filtering stage intuitively combines data from local features (via \emph{convolution}), then reduces the dimensions of the image (via \emph{pooling}) for the next stage.
-Graph neural networks have analagous convolution~\cite{gcn} (combining information from neighbouring nodes) and pooling~\cite{top-k-pooling} (merging nodes to reduce the size of the input graph) operators.
-
+Graph neural networks have analogous convolution~\cite{gcn} (combining information from neighbouring nodes) and pooling~\cite{top-k-pooling} (merging nodes to reduce the size of the input graph) operators.
 A brief period of experimentation with these operators yielded the following network architecture, shown in Figure \ref{figure:network}.
-It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimize the network parameters.
-The network is shown processing a formula in Figure \ref{figure:computation}.
+It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimise the network parameters.
 \begin{enumerate}
@@
 This architecture was implemented with the PyTorch~\cite{pytorch} neural network library, combined with a graph-processing (``geometric'') extension library, PyTorch Geometric~\cite{pytorch-geometric}, which together provide facilities for automatic differentiation, GPU-accelerated training, pre-programmed layers for graph processing, and various utilities.
-
-The dataset is split into a large training set and a smaller test set (200 balanced examples), since unsatisfiable examples were difficult to obtain.
-The training set was then augmented as described in section \ref{section:augmentation} to produce a total training set of 3.5 million examples.
-The network was trained on commodity desktop hardware with a mid-range GPU~\footnote{NVIDIA\textsuperscript{\textregistered} GeForce\textsuperscript{\textregistered} GT 730.} for 8 epochs/24 hours, optimizing a negative log-likelihood loss function.
+The dataset is split into a large training set and a smaller test set (200 balanced examples), since unsatisfiable examples were time-consuming to obtain in this setting.
+The unsatisfiable training data were then augmented as described in section \ref{section:augmentation} to produce a balanced total training set of 3.5 million examples.
+The network was trained on commodity desktop hardware with a mid-range GPU~\footnote{NVIDIA\textsuperscript{\textregistered} GeForce\textsuperscript{\textregistered} GT 730.} for 8 epochs/24 hours, optimising a negative log-likelihood loss function.
 
@@
 	\end{subfigure}
-	\caption{Computation in the neural network, showing intermediate values involved in the network (correctly) predicting that an input formula from the test set is unsatisfiable.
-Vectors of floating-point numbers are shown as grids.}
+	\caption{Computation in the neural network, showing intermediate values involved in the network (correctly) predicting the satisfiability of an input formula.}
 	\label{figure:computation}
@@
 While these results are very promising, it should be emphasised that it is unclear how effective a train/test split is in this setting (since similar subgoals may occur in both sets, even with proper data hygiene), and that this network is not attempting to determine the satisfiability of arbitrary formulae, merely those that occur in proof attempts on the \mizarlarge{} dataset.
-
 The higher precision and lower recall values are likely an artefact of the augmentation process.
-However, even with these caveats, the network performance is surpising and is practically useful for improving proof search in this dataset.
+However, even with these caveats, the network performance is surprising and is practically useful for improving proof search in this dataset.
 
@@
 \subsection{Benchmark Results}
-Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and LeRNA on the M2k dataset.
+Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and \lerna{} on the \mizarsmall{} dataset.
 \z3 ran for a full 10 seconds to establish baseline performance, then as an oracle for 20 milliseconds to determine the number of ``trivial'' problems.
-LeRNA ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.
+\lerna{} ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.
 
 \begin{table}
-	\caption{Total theorems proved on the M2k dataset.}
+	\caption{Total successful proof attempts on the \mizarsmall{} dataset.}
 	\centering
 	\begin{tabular}{r | c}
-		\textbf{Configuration} & \textbf{Theorems Proved}\\
+		\textbf{Configuration} & \textbf{Proofs}\\
 		\hline
@@
 		\z3 (20ms, as oracle) & 711\\
-		LeRNA, unguided (10s, with oracle) & 969\\
-		LeRNA, guided (10s, with oracle) & 1012\\
-						      & FIXME there is a bug, these ought be better\\
+		\lerna{}, unguided (10s, with oracle) & 969\\
+		\lerna{}, guided (10s, with oracle) & 1023\\
 	\end{tabular}
@@
 
+Although clearly worse, \lerna{} solves a significantly different set of problems compared to the baseline: 114 problems are uniquely solved by \lerna{}.
+It should be noted that the impressive results reported by the rlCoP system~\cite{rlCoP} are not directly comparable here due to differences in experimental setup.
+
 \section{Conclusions}
+The prototype system successfully implements a theorem prover with a neural heuristic processing the entire proof state, structured as a graph.
+After training on data automatically generated by the prover system, the neural network approach is shown to be practically useful for improving proof search performance.
+A number of approaches (batching, oracle invocations, parallelism) are employed to improve system efficiency.
+While the prototype is not yet a successful state-of-the-art ATP, it has some unique desirable properties, among them simplicity, parallelism, parametricity with respect to calculus/oracle/heuristic, and introspection of proof state.
+The general approach is flexible and presently unexplored.
+
 \section{Future Work}
@@
 In particular, the simplification routines can be improved to remove more trivial sub-formulae.
-While in general the oracles' pre-processing will remove these, they serve as noise for the neural network and might also increase the number of inference steps required to reach a proof.
-
-As one possible view of this approach is as an intelligent pre-processor (as mentioned in Section \ref{section:oracle}) for an existing efficient ATP, a greater number of aggressive and/or weakening inferences might be included in the calculus.
+While in general the oracles' preprocessing will remove these, they serve as noise for the neural network and might also increase the number of inference steps required to reach a proof.
+As one possible view of this approach is as an intelligent preprocessor (as mentioned in Section \ref{section:oracle}) for an existing efficient ATP, a greater number of aggressive and/or weakening inferences might be included in the calculus.
 For instance, \emph{prenexing} (or conversely \emph{miniscoping}) formulae can have a significant effect on proof search~\cite{clausification} for some theorem provers, so including suitable quantifier-manipulation rules might prove to be a useful extension.
 
-Generalising further, ideas from other refutation-tableaux calculi could well be suitable for this system.
-In particular, the authors are attempting to integrate an adapted connection rule from the non-clausal connection calculus~\cite{non-clausal-connections}, as used in nanoCoP~\cite{nanocop}, in order to reduce the number of proof steps required to instantiate universal quantifiers.
-
+Ideas from other refutation-tableaux calculi could well be suitable for this system.
+The authors are attempting to integrate an adapted connection rule from the non-clausal connection calculus~\cite{non-clausal-connections}, as used in nanoCoP~\cite{nanocop}, in order to reduce the number of proof steps required to instantiate universal quantifiers.
 Finally, this prover architecture can support other logics without excessive modification.
-Given that \z3 is already capable of supporting many \emph{theories}, such as arithmetic or datatypes, a many-sorted first order logic such as those described by SMTLIB~\cite{smtlib} or the TFF0 dialect of TPTP~\cite{tff0} seems an appropriate first step.
+Given that \z3 is already capable of supporting many \emph{theories}, such as arithmetic or datatypes, a many-sorted first order logic such as those described by SMTLIB~\cite{smtlib} or the TFF0 dialect of TPTP~\cite{tff0} seems appropriate.
 
@@
 While \z3 is a strong theorem prover in its own right and performs well here, it remains to be seen if it is the best for this application.
-Other ATPs (or indeed counter-example-finding systems) should be tested as it involves minimal engineering effort.
+Other ATPs (or counter-example-finding systems) should be tested as it involves minimal engineering effort.
 A \emph{portfolio} of several oracle systems working in tandem might also be considered, although of course this will eventually retard proof search linearly in the number of systems present.
-
 Reducing the number of oracle invocations is another area for optimisation.
-Currently, the system calls an oracle for every new subgoal generated.
-It seems unlikely that the subgoal is materially easier to dispatch than its parent (especially in the case of propositional inferences that do not split the goal), so heuristically or probabilistically removing such subgoals from the oracle's queue is a possible area for improvement.
-
+Currently, the system calls an oracle for every new sub-goal generated.
+It seems unlikely that the sub-goal is materially easier to dispatch than its parent (especially in the case of propositional inferences that do not split the goal), so heuristically or probabilistically removing such subgoals from the oracle's queue is a possible area for improvement.
 \lerna{} does not currently use any information from the oracle beyond its status: using auxiliary information such as satisfying models or unused formulae could well aid proof search.
@@
 Different approaches to formula-to-graph translation, symbol embeddings, data augmentation, and model integration may also be explored.
-
 \section{Acknowledgements}
 The authors wish to thank Josef Urban and his group in \v{C}VUT, Prague for their help and encouragement with early iterations of this work, and for supplying the Mizar dataset used in this paper.
-
+\clearpage
 \bibliographystyle{splncs04}

### 2019-05-13 papers/FroCoS19/paper.tex
self citation yay

@@
 Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
-However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
+However, in practice there are several implementation issues~\cite{lerna-discussion} which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
 \begin{enumerate}

### 2019-05-14 papers/FroCoS19/paper.tex
superficial presentation issues and make it fit

@@
-\documentclass[runningheads]{llncs}
+\documentclass{llncs}
 \usepackage{graphicx}
@@
 \usepackage{textcomp}
+\usepackage{wrapfig}
+
 \newcommand{\lerna}{\textsc{Lerna}}
@@
 \author{
-	Michael Rawson*\orcidID{0000-0001-7834-1567}\inst{1} \and
-	Giles Reger\inst{2}
+	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
+	Giles Reger
 }
@@
 	University of Manchester, UK\\
-	\email{michael@rawsons.uk}\and
-	University of Manchester, UK\\
-	\email{regerg@cs.man.ac.uk}
+	\email{michael@rawsons.uk}, \email{giles.reger@manchester.ac.uk}
 }
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Background}
-An understanding of neural networks~\cite{neural-survey}, and of first-order logic and automatic theorem provers~\cite{fol-atp-survey} is assumed.
-The particular syntax of first-order logic used in this work is presented here as clarification.
 
-\subsection{Syntax of First-Order Logic}
-Terms \(t\) are either
-\begin{enumerate}
-	\item Variables \(x\), bound by a quantifier.
-	\item Function applications \(f\left(t_1, t_2, \ldots, t_n\right)\) where \(f\) is an \(n\)-ary function symbol. Constants \(c\) are a degenerate case in which \(n = 0\).
-\end{enumerate}
-%
-Formulae \(\phi, \psi\) are one of
-\begin{enumerate}
-	\item Truth \(\top\) or falsehood \(\bot\).
-	\item Equalities \(t_1 = t_2\).
-	\item Predicate applications \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol.
-	\item Negations \(\lnot \phi\), implications \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
-	\item Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
-	\item Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).
-\end{enumerate}
+We assume basic familiarity with first-order logic, theorem proving, and  neural networks~\cite{neural-survey}.
 
-\subsection{Modern Automatic Theorem Provers}
-State-of-the-art ATP systems (e.g.
-E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}, CVC4~\cite{CVC4}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
-Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
-Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.
+\subsection{Logic and Theorem Proving}
+
+\paragraph{First-Order Logic}
+
+\lerna{} works with formulas in standard first-order logic with equality. Terms $t$ and formulas $\phi$ are recursively defined as follows
+\[
+\begin{array}{ll}
+t = & x \mid f\left(t_1, \ldots, t_n\right) \mid c \\
+\phi  =  & \top \mid \bot \mid t_1 = t_2 \mid p\left(t_1, \ldots, t_n\right) \mid \phi_1 \Rightarrow \phi_2 \mid \phi_1 \equiv \phi_2\mid \neg \phi \mid \\ &   \phi_1 \wedge \ldots \wedge \phi_n \mid \phi_1  \vee \ldots \vee \phi_n \mid \forall x.
+\phi \mid \exists x.
+\phi
+\end{array}
+\]
+where $x$ is a variable, $f$ is a $n$-ary function symbol, $c$ is a constant, and $p$ is a $n$-ary predicate symbol. Their meaning is defined as usual.
+
+%A formula is either true \(\top\), false \(\bot\), an equality between terms \(t_1 = t_2\), a predicate applied to terms \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol, a negated formula \(\lnot \phi\), an impliciation \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
+%Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
+%Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).
+
+\paragraph{Modern Automatic Theorem Provers}
+An automated theorem prover (ATP) is a system able to automatically establish whether a formula (in first-order logic) is \emph{satisfiable} or \emph{unsatisfiable}; although, given the undecidability of this problem, ATPs may also return \emph{unknown}.
+Both saturation-based provers (e.g.
+E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}) and SMT solvers (e.g.
+CVC4~\cite{CVC4} and Z3~\cite{Z3}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
+%Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
+%Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.
 
 \subsection{Machine Learning and Theorem Proving}
-Despite the efficiency of such systems, they can still spend time exploring areas that a human mathematician would discard, and tuning such systems manually is in general extremely difficult~\cite{portfolio}.
-There is therefore interest in applying machine-learning techniques to automatic theorem provers, with the eventual aim of an ``intelligent'' theorem prover which can learn from past experience to develop an intuition, discard uninteresting search space, and tune performance in a more principled way.
+Despite the efficiency of modern ATP systems, they can still spend time exploring areas that a human mathematician would discard, and tuning such systems is, in general, extremely difficult~\cite{portfolio}.
+This has led to the application of machine-learning techniques, with the eventual aim of an ``intelligent'' theorem prover able to learn from past experience to develop an intuition, discard uninteresting search space, and tune performance in a more principled way.
 
-Work has proceeded in the fields of premise selection~\cite{DeepMath,MaSh,MaLARea,formula-graph}, static strategy selection~\cite{Bridge,MaLeS,E-MaLeS}, dynamic (run-time) strategy selection~\cite{dynamic-strategy-priority} and more recently, direct proof guidance~\cite{MaLeCoP,FEMaLeCoP,rlCoP,ENIGMA,DNGPS}.
-Proof guidance typically involves some form of machine-learned \emph{heuristic} which biases proof search in some way, allowing the prover to avoid parts of the search space deemed uninteresting by the heuristic.
+Previous work has focussed on premise selection~\cite{DeepMath,MaSh,MaLARea,formula-graph}, static strategy selection~\cite{Bridge,MaLeS,E-MaLeS}, dynamic (run-time) strategy selection~\cite{dynamic-strategy-priority} and more recently, direct proof guidance~\cite{MaLeCoP,FEMaLeCoP,rlCoP,ENIGMA,DNGPS}.
+Proof guidance typically involves a form of machine-learned \emph{heuristic} which biases proof search in some way, allowing the prover to avoid parts of the search space deemed uninteresting by the heuristic.
 
 Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
-By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
+By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
 Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
 DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.
-rlCoP~\cite{rlCoP} is a connection-based reinforcement-learning prover which is not presently neurally-guided, but takes a similar approach to that taken in this paper and achieves impressive results.
 
-\subsection{Neural Networks for Formulae}
+\paragraph{rlCoP} The rlCoP system~\cite{rlCoP} is a connection-based reinforcement-learning prover which is not presently neurally-guided, but takes a similar approach to that taken in this paper and achieves impressive results.
+
+\paragraph{Neural Networks for Formulae}
 Neural networks are well-known tools for supervised learning~\cite{neural-survey}, and combined with trainable convolution/pooling operators are suitable for processing large-scale data such as images~\cite{cnn}.
@@
 
-\subsection{The MPTP Problem Set}
+\paragraph{The MPTP Problem Set}
 For training and evaluation purposes a set of problems exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Design}
@@
 Saturation-based provers can have very large proof states, for example.
-        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms tend to operate more efficiently in batches.
+        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms operate more efficiently in batches.
 Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
-        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space, without e.g.
+        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space.%, without e.g.
 unifiers.
@@
 To prove an inference, all the inferences' sub-goals must be proved.
-A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences.
+A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences (illustrated in Fig.~\ref{fig:search}).
 
-\begin{figure}
+%\begin{figure}[t]
+\begin{wrapfigure}[12]{r}{0.5\textwidth} 
+\vspace{-26pt}
 	\centering
-	\includegraphics[width=0.6\linewidth]{search}
-	\caption{Search in the \lerna{} system, showing shared sub-goals.}
-\end{figure}
+	\includegraphics[width=0.9\linewidth]{search2.pdf}
+	\caption{Search in the \lerna{} system, showing shared sub-goals.\label{fig:search}}
+\end{wrapfigure}	
+%\end{figure}
 
@@
 On the other hand, if a goal is trivial (i.e. contains a contradiction), it is unsatisfiable and can be marked as proven.
-This idea can be lifted to inferences: if an inference contains any satisfiable sub-goal, it too is satisfiable, whereas if an inference contains all unsatisfiable goals, it is unsatisfiable.
+This idea is lifted to inferences: if an inference contains any satisfiable sub-goal, it too is satisfiable, whereas if an inference contains all unsatisfiable goals, it is unsatisfiable.
 Proof search continues until the timeout is reached or the root goal is shown to be (un-)satisfiable.
@@
 
-\subsection{Architecture}
+\subsection{Architecture and Prototype Implementation}
 The system aims to consume all available CPU and GPU resources as efficiently as possible.
 To that end, proof search is asynchronous: the search algorithm generates new sub-goals, which are placed on two separate queues: one for the oracle ATP, another for heuristic evaluation.
-Proof search then continues elsewhere, while the oracle ATP is called in parallel on each sub-goal (consuming all available CPU) while the heuristic consumes batches of subgoals, efficiently utilising the available GPU resource.
-As information flows backwards from these processes, the search process updates its information about a given sub-goal and propagates that information upwards to the sub-goal's parent inferences, to influence future proof search: see Figure \ref{figure:architecture}.
+Proof search then continues elsewhere, while the oracle ATP is called in parallel on each sub-goal (consuming all available CPU) while the heuristic consumes batches of subgoals, efficiently utilising the available computational resource.
+As information flows backwards from these processes, the search process updates its information about a given sub-goal and propagates that information upwards to the sub-goal's parent inferences, to influence future proof search: see Fig.
+\ref{figure:architecture}.
 
-\begin{figure}
-	\includegraphics[width=\linewidth]{architecture}
-	\caption{A cartoon showing information flow in the system.}
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=0.8\linewidth]{architecture}
+	\caption{Illustrating information-flow in the system.}
 	\label{figure:architecture}
@@
 
-\subsection{Implementation}
-The Rust~\cite{rust} programming language was chosen to implement the main prover (minus the heuristic), for its memory and thread-safety guarantees, as well as its support for managing memory allocation and algebraic datatypes.
-The implementation is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
-Python 3~\cite{python3} was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
+%\subsection{}
+
+The prototype implementation (minus the heuristic) %was built in the Rust programming language.% was chosen to implement the main prover , for its memory and thread-safety guarantees, as well as its support for managing memory allocation and algebraic datatypes.
+%The implementation 
+is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
+Python 3 was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
 The heuristic is implemented as a server, communicating with the main prover via a TCP socket.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Calculus}
 \label{section:Calculus}
-The proof calculus used in this context may be extremely general: in fact, any function from goals to a finite set of possible inferences (themselves finite sets of sub-goals) will suffice, as long as each goal remains independent of any other such that the heuristic function can process all available information.
+The proof calculus used in the above architecture may be extremely general: in fact, any function from goals to a finite set of possible inferences (themselves finite sets of sub-goals) will suffice, as long as each goal remains independent of any other such that the heuristic function can process all available information.
 If the inference system is complete, there are no additional constraints such as orderings or fairness to ensure the completeness of the prover, as the balanced search algorithm (see Section \ref{section:Design}) will ensure this.
@@
 \lerna{} presently implements a refutation tableaux calculus~\cite{handbook-of-automated-reasoning} without unification.
-The calculus described is deliberately na\"ive in order to easily satisfy the design constraints described, but this is a weak point of this work and will be improved in future.
-A na\"ive calculus is not necessarily a problem in this context, as the heuristic should in principle select promising areas to explore and ignore uninteresting sub-goals.
+The calculus described is deliberately na\"ive in order to easily satisfy the design constraints given above, but may be replaced by a stronger calculus in the future. 
+A na\"ive calculus is not necessarily a problem as the heuristic should select promising areas to explore and ignore uninteresting sub-goals.
 However, a more efficient calculus would improve performance where the heuristic fails.
 
-\subsection{Refutation Tableaux}
+\paragraph{Refutation Tableaux}
 In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
@@
 
-\subsection{Complete Inferences}
-The inference rules in Figure \ref{figure:complete-inferences} form a complete inference system, by analogy with a first-order tableaux calculus without unification.
-A difference and point of interest is the rule for instantiating universal quantifiers: instead of instantiating a variable with any possible term \(t\) --- an infinite space --- it is instantiated with one function symbol (or constant) at a time, quantifying over new variables as needed.
-This allows for instantiating any term over multiple inference rules (effectively enumerating the Herbrand universe for the goal), but without an infinite number of possible inferences at any point.
-Equality is handled by a rule rewriting classes of equal ground terms.
-Both of these rules are complete yet inefficient, but both are likely to be used only a few times in order to provide enough of a ``hint'' to the oracle system for it to find a proof.
-
-\begin{figure}
+\begin{figure}[t]
 	\centering
@@
 
-\subsection{Weakening}
+\paragraph{Complete Inferences}
+The inference rules in Fig. \ref{figure:complete-inferences} form a complete inference system, by analogy with a first-order tableaux calculus without unification.
+A difference and point of interest is the rule for instantiating universal quantifiers: instead of instantiating a variable with any possible term \(t\) --- an infinite space --- it is instantiated with one function symbol (or constant) at a time, quantifying over new variables as needed.
+This allows for instantiating any term over multiple inference rules (effectively enumerating the Herbrand universe for the goal), but without an infinite number of possible inferences at any point.
+Equality is handled by a rule rewriting classes of equal ground terms.
+Both of these rules are complete yet inefficient, but both are likely to be used only a few times in order to provide enough of a ``hint'' to the oracle system for it to find a proof.
+
+\paragraph{Weakening}
 A weakening rule is an important part of \lerna{}'s calculus, since the \textsc{instantiation} and \textsc{equal} rule can produce a large number of formulae, some of which must be removed to help the oracle to prove the goal.
 Each application of the rule removes some amount of information from the goal in order to simplify it --- this is sound and corresponds to removing an axiom from proof search.
-The rule is merely
+The rule is merely\vspace{-0.5em}
 \begin{mathpar}
@@
 
-\subsection{Simplifications}
-Before each inferred goal is added to proof search, it is simplified with a simplification function mapping a goal to a possibly-simpler goal.
-This allows for efficient removal of tedious inferences such as double-negation elimination and generally reduces the search space.
-A set of example simplification rules are shown in figure \ref{figure:simplification-rules}.
+\paragraph{Simplifications}
+Before each inferred goal is added to proof search, it is simplified, % with a simplification function mapping a goal to a possibly-simpler goal.
+%This allows for efficient removal of 
+removing 
+tedious inferences such as double-negation elimination and generally %reduces 
+reducing 
+the search space.
+Fig. \ref{figure:simplification-rules} gives example simplification rules.
 
-\begin{figure}
+\begin{figure}[t]
 	\centering
 	\begin{mathpar}
-		\inferrule[double-negation]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
-		\inferrule[conjunction-associative]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
-		\inferrule[disjunction-propagation]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
-		\inferrule[reflexivity]{\top, \Gamma}{t = t, \Gamma}\and
+%		\inferrule[double-negation]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
+%		\inferrule[conjunction-associative]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
+%		\inferrule[disjunction-propagation]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
+%		\inferrule[reflexivity]{\top, \Gamma}{t = t, \Gamma}\and
+%		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
+		\inferrule[double-neg]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
+		\inferrule[conj-assoc]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
+		\inferrule[disj-prop]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
+		\inferrule[refl]{\top, \Gamma}{t = t, \Gamma}\and
 		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Oracle}
@@
 To rectify the situation, new goals generated by ongoing proof search are enqueued for attempted proof by an existing \emph{oracle} ATP system, as described in Section \ref{section:Design}. 
+%
+In our prototype implementation we use the mature
+%
+%\subsection{Z3}
+\z3{} SMT solver~\cite{Z3}, which % is an efficient and mature SMT solver, supporting a core propositional logic extended with ``theories'' such as arithmetic and datatypes.
+%It also notably 
+supports quantified first-order logic via a combination of decision procedures for decidable fragments (such as the Bernays-Sch\"onfinkel class of formulae), and heuristic quantifier instantiation routines~\cite{quantifier-instantiation}.
+\z3{} is attractive for this application due to its low startup times and its ability to produce both satisfiable and unsatisfiable results. 
+%
+%In the \lerna{} system, we ran \z3{} as an external program (thus reducing the coupling to any particular ATP, although \z3{} has an API which might improve efficiency) with its Model-Based Quantifier Instantiation heuristic for 20 milliseconds.
 
-\subsection{Z3}
-\z3{}\cite{Z3} is an efficient and mature SMT solver, supporting a core propositional logic extended with ``theories'' such as arithmetic and datatypes.
-It also notably supports first-order logic via a combination of decision procedures for decidable fragments (such as the Bernays-Sch\"onfinkel class of formulae), and heuristic quantifier instantiation routines~\cite{quantifier-instantiation}.
-\z3{} is attractive for this application due to its low startup times and its ability to produce both satisfiable and unsatisfiable results.
+\lerna{} uses \z3{} as an external system (it could be replaced by an alternative ATP), running it with its Model-Based Quantifier Instantiation heuristic for 20 milliseconds. 
+%
+This was chosen as the shortest time in which the oracle can dispatch a reasonable amount of trivial goals (and in fact \z3{} is so strong it dispatches some goals immediately: see Section \ref{section:results}).
+Longer oracle runtimes might produce better performance in future, but for this work longer runtimes begin to conflate the performance of the oracle and the performance of the system as a whole.
+This application is unusual for ATP systems --- very short runtimes, and a mix of true and false problem statements.
+%Running in this setting also amounts to fuzz-testing: in development of this system a bug was rediscovered in \z3{}\footnote{\url{https://github.com/Z3Prover/z3/issues/2101}} which resulted in non-termination of the prover: happily, the bug was already fixed in a newer version.
 
-In the \lerna{} system, we ran \z3{} as an external program (thus reducing the coupling to any particular ATP, although \z3{} has an API which might improve efficiency) with its Model-Based Quantifier Instantiation heuristic for 20 milliseconds.
-This was chosen as the shortest time in which the oracle can dispatch a reasonable amount of trivial goals (and in fact \z3{} is so strong it dispatches some goals immediately: see Section \ref{section:results}).
-Longer oracle runtimes might produce better performance in future, but for this experiment longer runtimes begin to conflate the performance of the oracle and the performance of the system as a whole.
-This application is unusual for ATP systems --- very short runtimes, and a mix of true and false problem statements.
-Running in this setting also amounts to fuzz-testing: in development of this system a bug was rediscovered in \z3{}\footnote{\url{https://github.com/Z3Prover/z3/issues/2101}} which resulted in non-termination of the prover: happily, the bug was already fixed in a newer version.
-
-\subsection{Acting as a Preprocessor}
-\lerna{} as a whole might also be seen as an intelligent preprocessor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
+\paragraph{Acting as a Preprocessor}
+\lerna{} might also be seen as an intelligent preprocessor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
 The system can therefore act as an adaptor for any existing ATP, adding parallelism opportunities and ``smoothing out'' sensitivity to input syntax.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Learned Heuristic}
@@
 A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
-Although the data is collected by running the system itself (and therefore might be considered \emph{reinforcement} learning~\cite{reinforcement}), for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.
+Although the data is collected by running the system itself and might be considered \emph{reinforcement} learning, for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.
 
@@
 In practice, undirected graphs improved model performance so the graph is made undirected before encoding node labels as one-hot inputs to produce the final input graphs.
-An example formula's translation is shown in Figure \ref{figure:translation}.
+An example formula's translation is shown in Fig. \ref{figure:translation}.
 
-\begin{figure}
+\begin{figure}[t]
 	\centering
@@
 Graph neural networks have analogous convolution~\cite{gcn} (combining information from neighbouring nodes) and pooling~\cite{top-k-pooling} (merging nodes to reduce the size of the input graph) operators.
-A brief period of experimentation with these operators yielded the following network architecture, shown in Figure \ref{figure:network}.
-It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimise the network parameters.
+A brief period of experimentation with these operators yielded the following network architecture, shown in Fig. \ref{figure:network}.
 \begin{enumerate}
@@
 \end{enumerate}
-
+It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimise the network parameters. 
 To reduce over-fitting, dropout~\cite{dropout} is applied in convolutional and fully-connected layers, \(p = 0.1\).
@@
 
-\subsection{Evaluation}
+\subsection{Network Evaluation}
 The network was evaluated on the balanced test set of 200 examples, as described.
@@
 
-\begin{table}
+\begin{table}[t]
 	\caption{Accuracy metrics for the neural heuristic.}
@@
 		Accuracy & 93.0\%\\
+		Precision&0.990\\	
+		Recall&0.884\\
+		\(F_1\)&0.934\\			
+	\end{tabular}
+	\quad
+	\begin{tabular}{r | c}
+		\textbf{Metric} & \textbf{Score}\\
+		\hline
 		True Positive&99\\
@@
 		False Negative&1\\
-		Precision&0.990\\
-		Recall&0.884\\
-		\(F_1\)&0.934\\
 	\end{tabular}	
@@
 \end{table}
-
-\section{Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Experimental Results}
 \label{section:results}
@@
 
-\subsection{Benchmark Results}
+%\subsection{Benchmark Results}
 Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and \lerna{} on the \mizarsmall{} dataset.
@@
 
-\begin{table}
+\begin{table}[t]
 	\caption{Total successful proof attempts on the \mizarsmall{} dataset.}
@@
 
-\section{Conclusions}
-The prototype system successfully implements a theorem prover with a neural heuristic processing the entire proof state, structured as a graph.
-After training on data automatically generated by the prover system, the neural network approach is shown to be practically useful for improving proof search performance.
-A number of approaches (batching, oracle invocations, parallelism) are employed to improve system efficiency.
-While the prototype is not yet a successful state-of-the-art ATP, it has some unique desirable properties, among them simplicity, parallelism, parametricity with respect to calculus/oracle/heuristic, and introspection of proof state.
-The general approach is flexible and presently unexplored.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Future Work}
 
-\section{Future Work}
-The style of theorem prover described opens many new directions for future work and improvements to the system.
+Given the prototype nature of this work, we have included a detailed discussion on future directions. 
+%The style of theorem prover described opens many new directions for future work and improvements to the system.
 As \lerna{} is a very new system, there is likely much to be gained by simple engineering and tuning: for example, the UCT exploration parameter \(c\) has been left at its theoretical optimum value \(\sqrt{2}\), but it is likely that a higher value will account for neural network inaccuracies and hence improve performance.
-Training on, benchmarking with, and optimising for other datasets (such as the famous TPTP benchmark~\cite{TPTP}) is also left as future work.
+Training on, benchmarking with, and optimising for other datasets (such TPTP or SMT-LIB) is also left as future work.
 
-\subsection{Proof Search}
+\paragraph{Proof Search}
 \lerna{} is well-suited for long-term proof search attempts in mathematics, such as those employed in the AIM project~\cite{AIM}: search is stable over time and does not produce a combinatorial explosion in the same way that some traditional systems tend to after a short period.
@@
 
-Another future direction for proof search is a principled incomplete mode, in which branches deemed sufficiently uninteresting by the heuristic are pruned, perhaps in response to time or memory constraints as seen in limited resource strategies~\cite{LRS}.
+Another future direction for proof search is a principled incomplete mode where branches deemed sufficiently uninteresting by the heuristic are pruned, perhaps in response to resource constraints as in limited resource strategies~\cite{LRS}.
 This approach, while clearly incomplete, would significantly accelerate proof search in the direction of more promising search within the available resources.
 
-\subsection{Prover Calculus}
-The calculus currently employed is deliberately na\"ive, so much work is planned here.
-In particular, the simplification routines can be improved to remove more trivial sub-formulae.
-While in general the oracles' preprocessing will remove these, they serve as noise for the neural network and might also increase the number of inference steps required to reach a proof.
-As one possible view of this approach is as an intelligent preprocessor (as mentioned in Section \ref{section:oracle}) for an existing efficient ATP, a greater number of aggressive and/or weakening inferences might be included in the calculus.
-For instance, \emph{prenexing} (or conversely \emph{miniscoping}) formulae can have a significant effect on proof search~\cite{clausification} for some theorem provers, so including suitable quantifier-manipulation rules might prove to be a useful extension.
+\paragraph{Prover Calculus}
+The calculus currently employed is deliberately na\"ive and extensions should be explored. 
+In particular, the simplification routines can be improved to remove more trivial sub-formulae as, while in general the oracles' preprocessing will remove these, they serve as noise for the neural network and might also increase the number of inference steps required to reach a proof.
+As one possible view of this approach is as an intelligent preprocessor for an existing ATP, more aggressive and/or weakening inferences might be included in the calculus.
+For instance, \emph{prenexing} (or conversely \emph{miniscoping}) formulae can have a significant effect on proof search for some theorem provers, so including suitable quantifier-manipulation rules might prove to be a useful extension.
 
@@
 Finally, this prover architecture can support other logics without excessive modification.
-Given that \z3 is already capable of supporting many \emph{theories}, such as arithmetic or datatypes, a many-sorted first order logic such as those described by SMTLIB~\cite{smtlib} or the TFF0 dialect of TPTP~\cite{tff0} seems appropriate.
+Given that \z3 is already capable of supporting many \emph{theories}, such as arithmetic or datatypes, a many-sorted first order logic such as those described by SMTLIB or the TFF0 dialect of TPTP seems appropriate.
 
-\subsection{Oracle}
+\paragraph{Oracle}
 While \z3 is a strong theorem prover in its own right and performs well here, it remains to be seen if it is the best for this application.
-Other ATPs (or counter-example-finding systems) should be tested as it involves minimal engineering effort.
+Other ATPs (or counter-example-finding systems) should be explored. 
 A \emph{portfolio} of several oracle systems working in tandem might also be considered, although of course this will eventually retard proof search linearly in the number of systems present.
@@
 
-\subsection{Machine-Learned Heuristic}
+\paragraph{Machine-Learned Heuristic}
 Many other graph-based neural architectures are possible.
@@
 Different approaches to formula-to-graph translation, symbol embeddings, data augmentation, and model integration may also be explored.
-\section{Acknowledgements}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusions}
+The introduced prototype \lerna{} system successfully implements a theorem prover with a neural heuristic processing the entire proof state, structured as a graph.
+After training on data automatically generated by the prover system, the neural network approach is shown to be practically useful for improving proof search performance.
+A number of approaches (batching, oracle invocations, parallelism) are employed to improve system efficiency.
+While the prototype is not yet a successful state-of-the-art ATP, it has some unique desirable properties, among them simplicity, parallelism, parametricity with respect to calculus/oracle/heuristic, and introspection of proof state.
+The general approach is flexible and presently unexplored.
+
+\paragraph{Acknowledgements}
 The authors wish to thank Josef Urban and his group in \v{C}VUT, Prague for their help and encouragement with early iterations of this work, and for supplying the Mizar dataset used in this paper.
-\clearpage
+%\clearpage
+
 \bibliographystyle{splncs04}

### 2019-05-15 papers/FroCoS19/paper.tex
small edits

@@
 To prove a goal, at least one inference must be proved.
-To prove an inference, all the inferences' sub-goals must be proved.
-A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences (illustrated in Fig.~\ref{fig:search}).
+To prove an inference, all the inferences' sub-goals must be proved (e.g. shown to be unsatisfiable).
+A simple optimisation is that sub-goals may be shared between inferences, so search becomes a directed acyclic graph, alternating between goals and inferences (illustrated in Fig.~\ref{fig:search}).
 
@@
 where \(s\) is the score associated with the inference, \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)), \(N\) is the number of visits to the current node, and \(n\) is the number of visits to the child.
+An inference's score is the minimum of their sub-goals' scores, and non-leaf sub-goal's score is the maximum of their inference's scores.
+The score \(s\) of a goal node is the score assigned to it by the heuristic function.
 After an inference is selected, the least-promising sub-goal is selected, and the selection process continues until a leaf node is reached.
-The score \(s\) of a leaf node is the score assigned to it by the heuristic function.
-An inference's score is the minimum of their sub-goals' scores, and non-leaf sub-goal's score is the maximum of their inference's scores.
 

### 2019-05-15 papers/FroCoS19/paper.tex
typos

@@
 \begin{enumerate}
-	\item Input. a graph \(G\) consisting of one-hot encoded nodes \(N\) and edges \(E\).
+	\item Input. A graph \(G\) consisting of one-hot encoded nodes \(N\) and edges \(E\).
 	\item Embedding. Each node is mapped to an embedding vector of size 64 via a trained dense embedding.
@@
 \end{enumerate}
-It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimise the network parameters. 
+It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimise the network architecture or hyper-parameters. 
 To reduce over-fitting, dropout~\cite{dropout} is applied in convolutional and fully-connected layers, \(p = 0.1\).

### 2019-05-15 papers/FroCoS19/paper.tex
giles and michael did some stuff

@@
 In order to balance exploitation of promising directions and exploration of all parts of the search space, a principled UCT-based search algorithm is used, as in MonteCoP~\cite{MonteCoP}.
-At each sub-goal, the prover chooses the inference maximising the expression
+At each sub-goal $g$, the prover chooses the inference $i$ and subgoal $sg$ with the maximal expression
 \[
-	s + c\sqrt{\frac{\ln N}{n}}
+{\sf max}_{i \in {\sf inferences}(g)}  	{\sf min}_{sg \in {\sf subgoals}(i)} \left ( {\sf score}(sg) + c~\sqrt{\frac{\ln {\sf visits}(g)}{{\sf visits}(sg)}} \right )  
 \]
-where \(s\) is the score associated with the inference, \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)), \(N\) is the number of visits to the current node, and \(n\) is the number of visits to the child.
-An inference's score is the minimum of their sub-goals' scores, and non-leaf sub-goal's score is the maximum of their inference's scores.
-The score \(s\) of a goal node is the score assigned to it by the heuristic function.
-After an inference is selected, the least-promising sub-goal is selected, and the selection process continues until a leaf node is reached.
+where ${\sf score}$ gives the heuristic score, ${\sf visits}$ gives the total number of visits to that node so far, and \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)).
+This prioritises subgoals considered satisfiable by the heuristic as establishing satisfiability makes the subgoals redundancy, allowing them to be pruned.
 
@@
 \label{section:results}
-In order to evaluate the effect of heuristic guidance on the existing prover, the prover ran both with and without guidance for 10 seconds on all available CPU cores.
+
+To show that neural guidance can improve the performance of \lerna{} the system was run with and without guidance for 10 seconds on all available CPU cores.
 All results were collected on commodity desktop hardware\footnote{Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark} i7-6700 CPU @ 3.40GHz, 16GB RAM.}.
@@
 \lerna{} ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.
-
+%
 \begin{table}[t]
@@
 \end{table}
+%
+With neural guidance \lerna{} was able to solve an additional 54 problems and overall \lerna{} was shown to be complementary to \z3{}, proving 114 problems that \z3{} was unable to solve on its own.
+Conversely, \z3{} was able to solve more problems in total, which unsurprising given the maturity of the tool.
+These results show that \lerna{} is able to learn from experience and complement an existing ATP.
 
-Although clearly worse, \lerna{} solves a significantly different set of problems compared to the baseline: 114 problems are uniquely solved by \lerna{}.
-It should be noted that the impressive results reported by the rlCoP system~\cite{rlCoP} are not directly comparable here due to differences in experimental setup.
+%Although clearly worse, \lerna{} solves a significantly different set of problems compared to the baseline: 114 problems are uniquely solved by \lerna{}.
+%It should be noted that the impressive results reported by the rlCoP system~\cite{rlCoP} are not directly comparable here due to differences in experimental setup.
 

### 2019-05-15 papers/FroCoS19/paper.tex
final figures, cosmetics

@@
 \documentclass{llncs}
+\usepackage{amsmath}
 \usepackage{graphicx}
@@
 
-\paragraph{Modern Automatic Theorem Provers}
+\paragraph{Automatic Theorem Provers}
 An automated theorem prover (ATP) is a system able to automatically establish whether a formula (in first-order logic) is \emph{satisfiable} or \emph{unsatisfiable}; although, given the undecidability of this problem, ATPs may also return \emph{unknown}.
@@
 
-Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
+Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
 By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
@@
 In order to balance exploitation of promising directions and exploration of all parts of the search space, a principled UCT-based search algorithm is used, as in MonteCoP~\cite{MonteCoP}.
-At each sub-goal $g$, the prover chooses the inference $i$ and subgoal $sg$ with the maximal expression
+At each sub-goal $g$, the prover chooses the inference \(i\) with subgoals \(s\) according to
 \[
-{\sf max}_{i \in {\sf inferences}(g)}  	{\sf min}_{sg \in {\sf subgoals}(i)} \left ( {\sf score}(sg) + c~\sqrt{\frac{\ln {\sf visits}(g)}{{\sf visits}(sg)}} \right )  
+	\max_{i \in g}\left[
+		\underbrace{\min_{s \in i}\left(\textsf{score}(s)\right)}_\text{exploitation} +~
+		c \times \underbrace{\sqrt{\frac{\ln\textsf{visits}\left(g\right)}{\textsf{visits}\left(i\right)}}}_\text{exploration}
+	\right]
 \]
 where ${\sf score}$ gives the heuristic score, ${\sf visits}$ gives the total number of visits to that node so far, and \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)).
-This prioritises subgoals considered satisfiable by the heuristic as establishing satisfiability makes the subgoals redundancy, allowing them to be pruned.
+The sub-goal with the minimal score is then selected: this prioritises subgoals considered possibly satisfiable by the heuristic, as satisfiable subgoals allows large parts of the search space to be pruned.
 
@@
 %The implementation 
-is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
+is currently just under 3,000 lines of Rust, not including the TPTP format parser or the implementation of perfect sharing.
 Python 3 was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
@@
 %
-With neural guidance \lerna{} was able to solve an additional 54 problems and overall \lerna{} was shown to be complementary to \z3{}, proving 114 problems that \z3{} was unable to solve on its own.
-Conversely, \z3{} was able to solve more problems in total, which unsurprising given the maturity of the tool.
+With neural guidance \lerna{} was able to solve an additional 54 problems and overall \lerna{} was shown to be complementary to \z3{}, proving 114 problems that \z3{} was unable to solve on its own, and 40 that neither unguided \lerna{} nor \z3{} could solve.
+Conversely, \z3{} was able to solve more problems in total, which is unsurprising given the maturity of the tool.
 These results show that \lerna{} is able to learn from experience and complement an existing ATP.

### 2019-05-24 papers/CADE19/paper.tex
update table 5

@@
 	\centering
-	\begin{tabular}{l l l}
-	Solver~~ & Total solved~~~ & Uniquely solved \\
+	\begin{tabular}{l l l l}
+	Solver~~ & Total solved~~~ & \multicolumn{2}{c}{Uniquely solved}  \\
+	& & All ~~~~~& Excluding Vampire (old) \\
 	\hline
-	Vampire (old)	& 13,057 & 0\\
-	Vampire (new)   & 13,191 & 54\\
-	E 		& 10,845 & 190\\
-	iProver		& 8,143 & 215\\
-	CVC4		& 9,354 & 501\\
+	Vampire (old)	& 13,057 & 0 & -\\
+	Vampire (new)   & 13,191 & 54 & 1030\\
+	E 		& 10,845 & 190 & 190\\
+	iProver		& 8,143 & 215 & 215\\
+	CVC4		& 9,354 & 501 & 502\\
 	\end{tabular}

### 2019-05-24 papers/CADE19/data/analyze.py

@@ -39,7 +39,7 @@ def read_strats(data_file_names):
 
 if __name__ == "__main__":
   
-  solvers = read_strats(["others.csv","info.csv"])
+  solvers = read_strats(["others.csv","info_no_baseline.csv"])
   if "solver_configuration" in solvers:
     del solvers["solver_configuration"]
   

### 2019-05-28 papers/CADE19/paper.tex
CADE final

@@
 \usepackage{amsmath}
+\usepackage{romannum}
 
@@
 The standard approach is to alternate between selecting the \emph{oldest} clause and the \emph{lightest} clause with a fixed, but configurable \emph{age/weight ratio} (AWR).
-We show that an optimal fixed value of this ratio can produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
-Based on these observations we develop several new modes for the Vampire prover  which vary AWR according to a ``shape'' during proof search.
-The modes solve a number of new problems in the TPTP benchmark.
+An optimal fixed value of this ratio is shown to produce proofs significantly more quickly on a given problem, and further that varying AWR during proof search can improve upon a fixed ratio.
+Several new modes for the Vampire prover which vary AWR according to a ``shape'' during proof search are developed based on these observations.
+The modes solve a number of new problems in the TPTP benchmark set.
 \end{abstract}
@@
 
-Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, or SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
+Currently, the most successful theorem provers (such as Vampire~\cite{vampire}, E~\cite{E}, and SPASS~\cite{Spass}) for first-order logic are saturation-based, utilising the well-known \emph{given-clause algorithm}.
 Simply, this algorithm \emph{saturates} a set of clauses by iteratively selecting a clause and performing all non-redundant inferences with it until all clauses have been selected or the empty clause (witnessing inconsistency) has been found.
@@
 Over the past few decades a certain amount of folklore has built up around the best methods for clause selection and recent work by \sandm{} \cite{clause-selection-heuristics} systematically studied these.
-Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time) which we instantiate with two simple patterns (or \emph{shapes}) that prove to be pragmatically useful.
+Our work extends this study with new results and also introduces the concept of a \emph{variable} clause selection strategy (one that changes over time), instantiated with two simple patterns (or \emph{shapes}) that prove to be pragmatically useful.
 
-We restrict our attention to clause selection strategies that alternate between selecting clauses based on age (i.e.
+Clause selection strategies that alternate between selecting clauses based on age (i.e.
 in a first-in-first-out manner) and weight (i.e.
-those with the fewest symbols first).
+those with the fewest symbols first) are the subject of this work.
 It was confirmed by \sandm{} that alternating these two heuristics outperforms either by itself.
-We refer to the ratio with which we use either heuristic as the \emph{age/weight ratio} (AWR) as this is the terminology employed by the Vampire theorem prover, which we use as a vehicle for our study.
+The ratio of these selections is the \emph{age/weight ratio} (AWR), as this is the terminology employed by the Vampire theorem prover, the vehicle for our study.
 
 After covering relevant background material in Section~\ref{sec:background} the remainder of the paper makes two main contributions.
-Firstly, in Section~\ref{sec:awr:study} we experimentally confirm the folklore that (i) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (ii) there is no `best' age/weight ratio: indeed, a large range of pragmatically useful ratios exists.
-Secondly, we demonstrate (Section~\ref{sec:varying:study}) that varying the age-weight ratio over time can achieve better performance than a fixed ratio and use this to motivate the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
-Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement, proving many new problems unsolvable by any current configuration of Vampire.
+Firstly, Section~\ref{sec:awr:study} experimentally confirms the folklore that (\romannum{1}) the choice of age/weight ratio often has a significant effect on the performance of proof search, and (\romannum{2}) there is no ``best'' age/weight ratio: indeed, a large range of pragmatically useful ratios exist.
+Section~\ref{sec:varying:study} demonstrates that varying the age-weight ratio over time can achieve better performance than a fixed ratio, and therefore motivates the addition of so-called \emph{age/weight shapes} for varying the ratio over time.
+Experiments (Section~\ref{sec:experiments}) with these new options implemented in the Vampire theorem prover show a significant improvement in coverage, proving many new problems unsolvable by any previous configuration of Vampire.
 
@@
 
-\paragraph{First-Order Logic and Weight.} Our setting is the standard first-order predicate logic with equality.
+\paragraph{First-Order Logic and Weight.}
+Our setting is the standard first-order predicate logic with equality.
 A formal definition of this logic is not required for this paper but an important notion is that of the \emph{weight} of a clause.
 In first-order logic,  terms are built from function symbols and variables, literals are built from terms, and clauses are disjunctions of literals.
-The weight of a term is the number of function and variable symbols occurring in it and this notion is lifted to clauses.
+The weight of a term/literal is the number of symbols (function, variable, or predicate) occurring in it.
+The weight of a clause is the sum of the weights of its literals.
 
@@
 \paragraph{The Given Clause Algorithm and AWR Clause Selection.}
-To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, whereas the \Passive\ clauses are those that have not.
+To achieve saturation, the \emph{given clause algorithm} organises the set of clauses into two sets: the \Active\ clauses are those that have been active in inferences, and the \Passive\ clauses are those that have not.
 Typically, a further \Unprocessed\ set is required in order to manage the clauses produced during a single iteration of the loop.
 Realisations of the given clause algorithm generally differ in how they organise simplifications.
-There are two main approaches (both implemented by Vampire, originally found in the eponymous theorem provers Otter~\cite{otter} and \textsc{Discount}~\cite{discount}): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop only uses \Active.
+There are two main approaches (both implemented by Vampire, originally found in the eponymous theorem provers Otter~\cite{otter} and \textsc{Discount}~\cite{discount}): the Otter loop uses both \Active\ and \Passive\ for simplifications, whereas the Discount loop uses only \Active.
 
@@
 	\item \emph{By Age (or First-in/First-out)} clause selection prefers the oldest clause (produced earlier in proof search), simulating a \emph{breadth-first} search of the clause space.
-In Vampire the age of a clause is the number of inferences performed to produce it (e.g.
-input clauses have age 0).
-	\item \emph{By Weight (or symbol-counting)} clause selection prefers the youngest clause (with the fewest symbols).
-The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are (in some sense) closer to this.
+In Vampire the age of a clause is the number of inferences performed to produce it (input clauses have age 0).
+	\item \emph{By Weight (or symbol-counting)} clause selection prefers the lightest clause.
+The intuition behind this approach is that the sought empty clause has zero symbols and lighter clauses are in some sense closer to this.
 Furthermore, lighter clauses are more general in terms of subsumption and tend to have fewer children, making them less explosive in terms of proof search.
 \end{itemize}
-\sandm~show that alternating these heuristics is beneficial and we consider this setting here.
-In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm.
-The balance is initialised to 0 and used as follows: a positive balance means that a clause should be selected by age, whereas a negative balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
-Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.
 
-\begin{figure}[t]
+\begin{figure}
 \begin{tabbing}
@@
 \>\>\>\IF\ $\Retained(\New)$ \THEN \` (* retention test *) \\ %\inc
-\>\>\>\>simplify $\New$ by clauses in $\Active \cup \Passive$ \semicol 
+\>\>\>\>simplify $\New$ by clauses in \Active \semicol 
                         \` (* forward simplification *) \\
@@
                         \` (* another retention test *) \\
-\>\>\>\>\>simplify $\Active$ and  $\Passive$ using $\New$ \semicol
+\>\>\>\>\>simplify $\Active$ using $\New$ \semicol
                         \` (* backward simplification *) \\
@@
 
+\sandm~show that alternating these heuristics is beneficial.
+In Vampire this alternation is achieved by an age/weight ratio (AWR) implemented by a simple \emph{balancing} algorithm.
+The balance is initialised to 0 and used as follows: a negative balance means that a clause should be selected by age, whereas a positive balance means that a clause should be selected by weight; given a ratio of $a:w$ the balance is incremented by $a$ when selecting by age and decremented by $w$ when selecting by weight.
+Figure~\ref{fig:discount} gives the Discount algorithm along with balance-based AWR clause selection. The lines relevant to clause selection are marked with \Mark.
+
 %%%%%%
@@
 
-We note here that Vampire's portfolio mode includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of non-goal clauses, thus preferring clauses in or derived from the goal in clause selection.
-Use of this heuristic is orthogonal to the age/weight ratio and we do not investigate its impact any further here.
+Vampire's portfolio mode also includes an additional option relevant to clause selection: the \texttt{--nongoal\_weight\_coefficient} option specifies a multiplier to apply to the weight of non-goal clauses, thus preferring clauses in or derived from the problem conjecture in clause selection.
+Use of this heuristic is orthogonal to the age/weight ratio and is not investigated further here.
 
@@
 \paragraph{Related Work.}
-We briefly review the clause selection approaches taken by other solvers. 
+Many clause selection approaches are taken by other solvers.
 Otter 3.3~\cite{otter3} selects \emph{either} by age, by weight or manually.
@@
 SPASS~\cite{Spass} uses symbol-counting based clause selection. 
-iProver~\cite{Iprover} follows in E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
+iProver~\cite{Iprover} follows E in having a number of configurable queues but relies mainly on age and weight heuristics in those queues. 
 The general idea in this paper of a \emph{varying age/weight ratio over time} is applicable to any ratio-based clause selection strategy, and our specific results apply to those that take a ratio between age and weight.
@@
 \label{sec:awr:study}
-
-We set out to experimentally confirm two assumptions from folklore:
+Two assumptions from folklore are confirmed experimentally:
 \begin{enumerate}
@@
 \end{enumerate}
-These were supported by the work of \sandm{} but here we focus on these points and explore them more deeply.
+These are supported by the work of \sandm{} but are explored in more depth here.
 
 \subsection{Logarithmic AWR}
-
-Optimising AWR values is more easily achieved if they have a continuous scale.
-AWR values are mathematically \(\mathbb{Q}^{+}\), the positive rational numbers, but in practice are more easily seen logarithmically: 1:16 is the same distance in terms of the effect on proof search from 1:8 as it is from 1:32.
-Therefore, we introduce the \emph{logarithmic} AWR \(L\), defined in terms of age \(A\) and weight \(W\) as
+Visualising AWR values is more easily achieved if they have a continuous scale.
+AWR values are mathematically \(\mathbb{Q}^{+}\), the positive rational numbers, but in practice are more easily visualised logarithmically.
+Therefore, the \emph{logarithmic} AWR \(L\) is defined in terms of age \(A\) and weight \(W\) as
 \[
@@
 \]
-With this definition, AWR can now be seen continuously.
 As \(L\) tends to positive infinity, Vampire selects only by weight, whereas if \(L\) tends to negative infinity Vampire selects only by age.
 \(L = 0\) represents the middle ground of a 1:1 age/weight ratio.
-Note that we have not updated the balancing algorithm used by Vampire to make use of this value (this still requires two numbers) but use the number in experiments when discussing AWR.
+Note that the balancing algorithm used by Vampire does not make use of this value (it still requires two numbers) but the quantity is used in this work to show continuous AWR values.
 
@@
 	\caption{
-The number of activations reported by Vampire after successful 1-second runs on a TPTP problem.
+		The number of given-clause loops reported by Vampire after finding a proof with 1-second runs on a TPTP problem, \texttt{PRO017+2}.
 In between the peaks on either side, the function of \(L\) is discontinuous with large peaks and troughs, but follows an approximate trend and settles toward the global optimum.
-This is typical for the observed set of problems.
+\texttt{PRO017+2} exhibits typifying behaviour for TPTP, based on manual inspection of several hundred such plots.
 }
@@
 }.
-On this problem, a good AWR value is over 400\% better by this metric than the worst AWR value.
+On the problem shown, a good AWR value is over 400\% better by this metric than the worst AWR value.
 
@@
 Relative performance gain, showing the ratio in activations between the best AWR setting for a given problem and another base setting.
-We compare  1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
+A comparison is drawn between 1:1 (Vampire's default), 1:5 (the best-behaved from \sandm), and the worst setting for the problem.
 Where the problem is not solved at all by the base setting, it is ignored.
@@
 	\begin{tabular}{c | c | c c}
-		& Maximum Gain & Mean Gain & (Standard Deviation)\\
+		Base Setting & \% Maximum Gain & \% Mean Gain & (Standard Deviation)\\
 		\hline
-		1:1 & 133.56 & 1.26 & 1.63 \\
-		1:5 & 133.67 & 1.44 & 1.70 \\
-		(worst) & 222.01 & 3.95 & 7.60 \\
+		1:1 & 13,356 & 126 & 163 \\
+		1:5 & 13,367 & 144 & 170 \\
+		(worst) & 22,201 & 395 & 760
 	\end{tabular}
@@
 
-This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g.
-higher-order problems).
-We ran Vampire for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic so we avoid it for such experiments.} using a sensible set of AWR values (see Table \ref{tab:no-best-awr}) --- these are the values used in Vampire's portfolio mode.
+This experiment was repeated on the whole TPTP problem set, excluding problems Vampire does not currently support (e.g. higher-order problems).
+Vampire ran for 1 second in default mode with the \texttt{discount} saturation algorithm\footnote{The default LRS~\cite{LRS} saturation algorithm can be non-deterministic.} using a sensible set of AWR values (see Table \ref{tab:no-best-awr}) --- these are the values used in Vampire's portfolio mode.
 These tend to favour weight-first over age-first as this has been experimentally shown to be preferable.
-We then filter out problems not solved by any of these or those solved trivially (e.g in preprocessing).
-The whole set yielded similar AWR data for 7,947 problems.
+Problems not solved by any of these, or those solved trivially (e.g in preprocessing) are removed.
+The whole set yielded data for 7,947 problems.
 
@@
 Table~\ref{tab:awr-improvement} summarises the impact that choosing the best AWR can have.
-Compared to the default we can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr}) shows we also lose just under 10\% of problems by choosing the default.
-It is more relevant to note that there are cases where we can do \emph{much} better by selecting a different AWR value.
-So choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster, but not necessarily.
-In the worst case (choosing the pessimal AWR value) we can perform almost 4 times as many activations.
+Compared to the default, Vampire can perform, on average, 1.26 times fewer activations, which is modest but (as Table \ref{tab:no-best-awr} shows) just under 10\% of problems are no longer proven by choosing the default.
+It is more relevant to note that there are cases where Vampire can do \emph{much} better by selecting a different AWR value.
+Therefore, choosing a better AWR value \emph{can} go from no solution to a solution and \emph{can} do so faster, but not necessarily.
+In the worst case (choosing the pessimal AWR value) Vampire performs almost 4 times as many activations.
 
 The second result is that there is no ``best'' AWR across this full set of problems.
-Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, the maximum and mean drop in performance.
-The drop in performance is how many times more activations were required for a proof in the worst/average case compared to the best.
+\emph{Drop in performance} is defined to be how many times more activations were required for a proof under a given AWR, compared to the best AWR.
+Table \ref{tab:no-best-awr} shows, for each AWR value, the \% of problems solved, the number solved uniquely, and the maximum and mean drop in performance.
 No AWR value solves all problems, with the best being 1:5. 
-A ratio of 1:4 produces an unusually well-behaved maximum performance drop.
+A ratio of 1:4 produces an unusually small maximum performance drop.
 \sandm~found that 1:5 had a similar property, but this might be explained by differences in prover and test environment.
@@
 	\begin{tabular}{l | c | c | c | c  c }
-		AWR & \% Solved & Uniques & Maximum Drop & Mean Drop & (Standard Deviation)\\
+		AWR & \% Solved & Uniques & \% Maximum Drop & \% Mean Drop & (Standard Deviation)\\
 \hline
-8:1 & 85.25 & 16 & 150.67 & 1.37 & 1.98\\
-5:1 & 86.10 & 1 & 122.22 & 1.33 & 1.64\\
-4:1 & 86.93 & 1 & 101.44 & 1.32 & 1.42\\
-3:1 & 87.63 & 2 & 105.00 & 1.29 & 1.41\\
-2:1 & 88.62 & 3 & 112.67 & 1.27 & 1.45\\
-3:2 & 89.83 & 2 & 119.89 & 1.27 & 1.51\\
-5:4 & 89.98 & 4 & 125.00 & 1.26 & 1.55\\
-1   & 90.56 & 4 & 133.56 & 1.26 & 1.63\\
-2:3 & 91.20 & 9 & 147.67 & 1.28 & 1.79\\
-2   & 91.68 & 0 & 162.67 & 1.31 & 1.97\\
-3   & 91.81 & 5 & 190.56 & 1.37 & 2.30\\
-4   & 91.85 & 3 & \textbf{17.41} & 1.38 & 0.67\\
-5   & 92.00 & 2 & 133.67 & 1.44 & 1.70\\
-6   & 91.57 & 1 & 106.44 & 1.47 & 1.46\\
-7   & 91.49 & 1 & 104.89 & 1.49 & 1.44\\
-8   & 91.09 & 2 & 101.33 & 1.53 & 1.45\\
-10  & 90.52 & 1 & 101.78 & 1.60 & 1.53\\
-12  & 90.00 & 0 & 101.67 & 1.65 & 1.62\\
-14  & 89.29 & 4 & 103.00 & 1.70 & 1.75\\
-16  & 89.42 & 5 & 101.33 & 1.74 & 1.76\\
-20  & 88.61 & 3 & 100.89 & 1.82 & 1.94\\
-24  & 88.26 & 2 & 101.33 & 1.89 & 2.08\\
-28  & 87.57 & 2 & 99.22 & 1.96 & 2.24\\
-32  & 87.01 & 1 & 100.00 & 1.99 & 2.36\\
-40  & 86.23 & 4 & 98.78 & 2.09 & 2.64\\
-50  & 84.93 & 1 & 98.78 & 2.17 & 2.88\\
-64  & 84.17 & 2 & 101.22 & 2.28 & 3.19\\
-128 & 81.34 & 3 & 107.44 & 2.57 & 4.16\\
-1024 & 73.11 & 23 & 222.01 & 2.83 & 7.55
+8:1 & 85.25 & 16 & 15,067 & 137 & 198\\
+5:1 & 86.10 & 1 & 12,222 & 133 & 164\\
+4:1 & 86.93 & 1 & 10,144 & 132 & 142\\
+3:1 & 87.63 & 2 & 10,500 & 129 & 141\\
+2:1 & 88.62 & 3 & 11,267 & 127 & 145\\
+3:2 & 89.83 & 2 & 11,989 & 127 & 151\\
+5:4 & 89.98 & 4 & 12,500 & 126 & 155\\
+1:1 & 90.56 & 4 & 13,356 & 126 & 163\\
+2:3 & 91.20 & 9 & 14,767 & 128 & 179\\
+1:2 & 91.68 & 0 & 16,267 & 131 & 197\\
+1:3 & 91.81 & 5 & 19,056 & 137 & 230\\
+1:4 & 91.85 & 3 & \textbf{1,741} & 138 & 67\\
+1:5 & 92.00 & 2 & 13,367 & 144 & 170\\
+1:6 & 91.57 & 1 & 10,644 & 147 & 146\\
+1:7 & 91.49 & 1 & 10,489 & 149 & 144\\
+1:8 & 91.09 & 2 & 10,133 & 153 & 145\\
+1:10& 90.52 & 1 & 10,178 & 160 & 153\\
+1:12& 90.00 & 0 & 10,167 & 165 & 162\\
+1:14& 89.29 & 4 & 10,300 & 170 & 175\\
+1:16& 89.42 & 5 & 10,133 & 174 & 176\\
+1:20& 88.61 & 3 & 10,089 & 182 & 194\\
+1:24& 88.26 & 2 & 10,133 & 189 & 208\\
+1:28& 87.57 & 2 & 9,922 & 196 & 224\\
+1:32& 87.01 & 1 & 10,000 & 199 & 236\\
+1:40& 86.23 & 4 & 9,878 & 209 & 264\\
+1:50& 84.93 & 1 & 9,878 & 217 & 288\\
+1:64& 84.17 & 2 & 10,122 & 228 & 319\\
+1:128&81.34 & 3 & 10,744 & 257 & 416\\
+1:1024&73.11 & 23 & 22,201 & 283 & 755
 	\end{tabular}
@@
 \section{Variable AWR for Vampire} 
-
-In this section we motivate and define a variable clause selection approach.
+This section motivates and defines a clause selection approach which varies the AWR value over time.
 
@@
 \subsection{Varying AWR (in Vampire)} \label{sec:varying:implementation}
-
-We describe how we vary AWR dynamically and how this is achieved in Vampire. 
-%
-%If dynamically varying AWR is sometimes useful for finding proofs more quickly, this could be useful as a Vampire option.
-In general we would like to describe any possible sequence that the AWR could follow during proof search.
+An implementation of dynamically-varying AWR values in Vampire is described below.
+In general any possible sequence that the AWR could follow during proof search can be used.
 However, some details constrain the design space:
@@
 
-In this work we select two general trends to explore: a trend away from the original (fixed) AWR toward 1:1 (``decay''), and a trend from 1:1 toward the original setting (``converge'').
-In our investigations we found that even fluctuating sequences had a general trend and these two fixed trends are reasonable approximations of this.
-We have chosen to use an `original' AWR value to move away from/towards as this original AWR has already been shown to be `useful' in some sense.
+In this work two general trends are explored: a trend away from a given \emph{start} AWR toward 1:1 (``decay''), and a trend from 1:1 toward a given \emph{end} AWR (``converge'').
+Investigation showed that even fluctuating sequences had a general trend, and further that these two fixed trends are reasonable approximations of these trends.
+The start/end AWR values are taken from the portfolio mode: these values are known to be useful in a fixed-AWR context, and while this may not generalise to a dynamic-AWR context, it is a useful starting point pending integration of AWR shape parameters into strategy scheduling.
 
@@
 Our approach here was restricted by the balancing algorithm used internally, as AWR steps must be discrete and do not take effect immediately.
-An alternative approach might use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducibly) taken with the use of a seeded PRNG.
-This would permit use of continuous age/weight functions, but would also introduce the possibility of ``getting unlucky'' in which a clause preferred by the age/weight probability is not selected due to an improbable-but-possible series of PRNG samples.
+An alternative approach might be to use an age/weight probability, rather than a ratio, from which age or weight decisions would be pseudo-randomly (but reproducibly) taken with the use of a seeded pseudo-random number generator, permitting use of continuous age/weight functions.
 
@@
 
-We introduce two new options: \texttt{--age\_weight\_ratio\_shape} can take the values \emph{constant}, \emph{decay}, or \emph{converge} and selects one of the above shapes; and \texttt{--age\_weight\_ratio\_shape\_frequency} specifies the frequency (rate) or convergence/decay (default is 100).
+Two new options are implemented: \texttt{--age\_weight\_ratio\_shape} can take the values \emph{constant}, \emph{decay}, or \emph{converge} and selects one of the above shapes; and \texttt{--age\_weight\_ratio\_shape\_frequency} specifies the frequency (rate) or convergence/decay (default is 100).
 These are used with the existing \texttt{--age\_weight\_ratio} option (default 1:1) to give a number of new option combinations, which can be used in conjunction with Vampire's portfolio mode pending integration into the strategy schedules.
 This version of the prover is currently in a separate branch in the main Vampire source repository\footnote{\texttt{https://github.com/vprover/vampire/tree/awr-shapes}}.
+Another option, \texttt{--age\_weight\_ratio\_b} is implemented (default 1:1), controlling the initial AWR value of \emph{converge} or the final AWR value of \emph{decay}.
+This parameter was not evaluated for reasons of computational complexity.
 
@@
 
-We perform two experiments to evaluate the new techniques.
+Two experiments evaluate the new techniques.
 The first compares the various options attempting to draw some conclusions about which option values work well together.
@@
 \begin{table}[t]
-\caption{Number of problems solved (top) and unique problems solved (bottom) by various configurations varying AWR, AWR shape, and AWR frequency.\label{tab:short:runs}}
+\caption{
+	Number of problems solved (top) and unique problems solved (bottom) by various configurations varying AWR, AWR shape, and AWR frequency.
+	Bold numbers indicate the best result within a given shape.
+}
+\label{tab:short:runs}
 \centering
@@
 
-We ran Vampire in default mode (with {\tt discount} saturation algorithm) for 10s whilst varying  \texttt{age\_weight\_ratio} and \texttt{age\_weight\_ratio\_shape\_frequency} for both new values of \texttt{age\_weight\_ratio\_shape}.
-We also ran the constant default AWR of 1:1 as a baseline.
+Vampire ran in default mode (with the \texttt{discount} saturation algorithm) for 10s whilst varying  \texttt{age\_weight\_ratio} and \texttt{age\_weight\_ratio\_shape\_frequency} for both new values of \texttt{age\_weight\_ratio\_shape}.
+Constant AWR values also ran as a baseline.
 
-The results are given in Table~\ref{tab:short:runs}. The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value.
-We separately report the total number of problems solved and those solved uniquely (by a single strategy across all experiments, including the baseline).
+Results are given in Table~\ref{tab:short:runs}.
+The results for the different shapes are grouped into columns and then by frequency with rows giving results per AWR value.
+The total number of problems solved and those solved uniquely are also reported.
 
@@
 
-Our next experiment aims to answer the question ``How much can the portfolio mode of Vampire improve using these new options?''.
-To address this we use the portfolio mode used in the most recent CASC competition~\cite{CADE18} as a baseline and run the new options on top of this.
-Note that this portfolio mode contains techniques completely unrelated to the age/weight ratio, e.g.
+Our next experiment aims to answer the question ``How much can the portfolio mode of Vampire be improved using these new options?''.
+To address this the new options ran on top of the portfolio mode used in the most recent CASC competition CASC-J9~\cite{CASC}.
+Note that the CASC-J9 portfolio mode contains techniques completely unrelated to the age/weight ratio, e.g.
 finite model building~\cite{fmb}, as well as other options related to clause selection, e.g.
@@
 Vampire first ran to establish baseline performance in the given portfolio mode on all problems in TPTP, with a wallclock time limit of 300 seconds.
-We then forced the new options on top of the portfolio mode, using the existing AWR values in the various strategies as the starting point.
+New options were applied on top of the portfolio mode options, using the existing AWR values in the various strategies as the starting point.
 The purpose is to gauge what impact adding such options to a new portfolio mode could have.
-In this experiment we are mainly hoping to find new solved problems and identify new strategies that could be added to a portfolio mode.
-Therefore, it makes sense to consider the union of all experiments ran.
+In this experiment the aim was to find new solved problems and identify new strategies that could be added to a portfolio mode.
+Therefore, it makes sense to consider the union of all experiments.
 
@@
 Figure \ref{tab:results} shows the performance in terms of solved problems of all the configurations tested.
-These data show that configurations which are more similar to the baseline (i.e. slow decay or fast convergence) achieve more similar performance, as expected.
-In total, 134 problems were solved by the new configurations that were not solved by the baseline.
+These data show that configurations which select clauses in a similar way to the baseline (i.e. slow decay or fast convergence) achieve similar performance, as expected.
+In total, 134 (\(117 + 17\), \(93 + 41\))  problems were solved by the new configurations that were not solved by the baseline.
 This is an impressive result --- it is rare to be able to improve portfolio mode by this many new problems with a single new proof search option.
 
-We also report the \emph{u-score} computed by giving $1/n$ points per problem solved where $n$ is the number of strategies solving a problem~\cite{u-score}.
+The \emph{\(u\)-score} is computed by giving $1/n$ points per problem solved where $n$ is the number of strategies solving a problem~\cite{u-score}.
 This gives a measure of contribution per strategy.
-To consider new options for extending the portfolio mode we will consider options from largest u-score to smallest (and only those with unique solutions overall).
+Options with the largest \(u\)-score will be prioritised for extending the existing portfolio mode, but only those with unique solutions overall.
 
-Finally, two problems were solved which were marked with an ``Unknown'' status in the TPTP headers.
+Finally, two problems were solved which were marked with an ``Unknown'' status (with rating 1.00) in the TPTP headers.
 Only converging with frequency 50 solved \texttt{SET345-6} and only decaying with frequency 1 solved \texttt{LAT320+3}.
@@
 
-To place these results in context, we compare the overall number of problems solved by our new strategies with the results of other solvers, using their 2018 CASC competition mode.
+To place these results in context, the overall number of problems solved by our new strategies are compared with the results of other solvers, using their CASC-J9
 These results are from 300-second runs in identical conditions and are given in Table~\ref{tab:other}.
-In this table \emph{Vampire (old)} stands for the 2018 competition version whilst \emph{Vampire (new)} stands for the union of all problems solved by new options in the previous section.
+In this table \emph{Vampire (old)} stands for the CASC-J9 competition version whilst \emph{Vampire (new)} stands for the union of all problems solved by new options in the previous section.
 Between them, the two versions of Vampire solve 1,030 problems uniquely.
@@
 Clause selection is a key part of any saturation-based theorem prover and age/weight ratios have a significant effect on the performance of proof search in the Vampire theorem prover.
-We have supported the known folklore that there is no clear optimal age/weight ratio.
-Further, we have demonstrated that varying the age/weight ratio over time \emph{during} proof search can improve further on an optimal, but fixed age/weight ratio in terms of the number of activations.
-Experiments within Vampire on the TPTP benchmark suggest that these \emph{age/weight shapes} show promise for future developments in this novel approach to proof search.
+Known folklore that there is no clear optimal age/weight ratio is supported.
+Further, varying the age/weight ratio over time \emph{during} proof search can improve further on an optimal, but fixed age/weight ratio in terms of the number of activations.
+Experiments within Vampire on the TPTP benchmark set suggest that these \emph{age/weight shapes} show promise for future developments in this novel approach to proof search.
 Indeed, including our relatively simple shapes already leads to significant performance gains.

### 2019-06-05 papers/CADE19/paper.tex
added 1:4 figures

@@
 \usepackage{amsmath}
+\usepackage{afterpage}
+\usepackage{pdflscape}
 \usepackage{romannum}
@@
 Another option, \texttt{--age\_weight\_ratio\_b} is implemented (default 1:1), controlling the initial AWR value of \emph{converge} or the final AWR value of \emph{decay}.
-This parameter was not evaluated for reasons of computational complexity.
 
@@
 \label{sec:experiments}
+
+\afterpage{
+\begin{landscape}
+\begin{table}[p]
+\caption{
+	Number of problems solved (top) and unique problems solved (bottom) by various configurations varying start/end AWR values, AWR shape, and AWR frequency.
+	Bold numbers indicate the best result within a given shape.
+}
+\label{tab:short:runs}
+\centering
+\begin{tabular}{l|l|lllll|lllll|lllll|lllll}
+	& & \multicolumn{5}{|c}{\texttt{converge from 1:1}} & \multicolumn{5}{|c}{\texttt{decay to 1:1}} & \multicolumn{5}{|c}{\texttt{converge from 1:4}} & \multicolumn{5}{|c}{\texttt{decay to 1:4}}\\  
+	AWR & \texttt{constant} & \multicolumn{5}{|c|}{Frequency} & \multicolumn{5}{|c}{Frequency} & \multicolumn{5}{|c}{Frequency} & \multicolumn{5}{|c}{Frequency}\\  
+&
+& 1 & 10 & 100 & 1000 & Union
+& 1 & 10 & 100 & 1000 & Union
+& 1 & 10 & 100 & 1000 & Union
+& 1 & 10 & 100 & 1000 & Union
+\\
+\hline
+\multicolumn{20}{c}{Problems Solved} \\ \hline
+10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & \emph{8372} & 8448 & 8441 & 8380 & 8169 & \emph{8614} & 7983 & 7990 & 8094 & 8323 & \emph{8485} & 8579 & 8574 & 8493 & 8272 & \emph{8797}\\
+1:10		& 8575 & 8565 & \textbf{8578} & 8550 & 8489 & \emph{8778} & 8458 & 8456 & 8484 & 8268 & \emph{8787} & 8575 & \textbf{8584} & 8582 & 8535 & \emph{8729} & 8584 & 8574 & 8590 & 8608 & \emph{8764}\\
+1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & \emph{8560} & 8454 & 8484 & 8537 & \textbf{8636} & \emph{8907} & 8088 & 8084 & 8071 & 8216 & \emph{8399} & 8572 & 8584 & \textbf{8615} & 8592 & \emph{8855}\\
+1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & \emph{8379} & 8470 & 8492 & 8492 & 8473 & \emph{8873} & 7283 & 7300 & 7364 & 7927 & \emph{8076} & 8567 & 8567 & 8566 & 8446 & \emph{8830}\\
+\hline
+Union           & \emph{9019} & \emph{9028} & \emph{9016} & \emph{8981} & \emph{8871} & \emph{9194} & \emph{8572} & \emph{8674} & \emph{8725} & \emph{8978} & \emph{9048} & \emph{9038} & \emph{9038} & \emph{9016} & \emph{8967} & \emph{9180} & \emph{8697} & \emph{8759} & \emph{8800} & \emph{8927} & \emph{9026}\\
+\hline
+\hline
+\multicolumn{20}{c}{Uniquely Solved} \\ \hline
+10:1 & 1 & 2 & 1 & 1 & \textbf{6} & \emph{10} & 0 & 0 & 0 & 2 & \emph{2} & 1 & 0 & 0 & 1 & \emph{2} & 0 & 0 & 2 & \textbf{4} & \emph{6}\\
+1:10 & 0 & 1 & 0 & 0 & 1 & \emph{2} & 0 & 0 & 0 & \textbf{6} & \emph{6} & 0 & 1 & 0 & 0 & \emph{1} & 0 & 0 & 0 & 0 & \emph{0}\\
+1:100 & 1 & 0 & 0 & 5 & 4 & \emph{9} & 2 & 0 & 3 & 3 & \emph{8} & 0 & 0 & 0 & 2 & \emph{2} & 0 & 1 & 1 & 0 & \emph{2}\\
+1:1000 & 0 & 0 & 0 & 0 & 1 & \emph{1} & 2 & 0 & 0 & 2 & \emph{4} & 0 & 1 & \textbf{3} & \textbf{3} & \emph{7} & 0 & 0 & 0 & 0 & \emph{0}\\
+\hline
+Union & \emph{2} & \emph{3} & \emph{1} & \emph{6} & \emph{12} & \emph{22} & \emph{4} & \emph{0} & \emph{3} & \emph{13} & \emph{20} & \emph{1} & \emph{2} & \emph{3} & \emph{6} & \emph{12} & \emph{0} & \emph{1} & \emph{3} & \emph{4} & \emph{8}\\
+\end{tabular}
+\end{table}
+\end{landscape}
+}
 
@@
 
-\begin{table}[t]
-\caption{
-	Number of problems solved (top) and unique problems solved (bottom) by various configurations varying AWR, AWR shape, and AWR frequency.
-	Bold numbers indicate the best result within a given shape.
-}
-\label{tab:short:runs}
-\centering
-\begin{tabular}{l|l|lllll|lllll}
-	& & \multicolumn{5}{|c}{\texttt{converge}} & \multicolumn{5}{|c}{\texttt{decay}}\\  
-AWR & \texttt{constant} & \multicolumn{5}{|c|}{Frequency} & \multicolumn{5}{|c}{Frequency}\\  
-&& 1 & 10 & 100 & 1000 & Union & 1 & 10 & 100 & 1000 & Union\\
-\hline
-\multicolumn{10}{c}{Problems Solved} \\ \hline
-10:1		& 7967 & 7972 & 7976 & 8050 & 8245 & \emph{8372} & 8448 & 8441 & 8380 & 8169 & \emph{8614}\\
-1:10		& 8575 & 8565 & \textbf{8578} & 8550 & 8489 & \emph{8778} & 8458 & 8456 & 8484 & 8268 & \emph{8787}\\
-1:100	        & 8079 & 8084 & 8079 & 8039 & 8279 & \emph{8560} & 8454 & 8484 & 8537 & \textbf{8636} & \emph{8907}\\
-1:1000     	& 7276 & 7279 & 7297 & 7418 & 8133 & \emph{8379} & 8470 & 8492 & 8492 & 8473 & \emph{8873}\\
-\hline
-Union           & \emph{9019} & \emph{9028} & \emph{9016} & \emph{8981} & \emph{8871} & \emph{9194} & \emph{8572} & \emph{8674} & \emph{8725} & \emph{8978} & \emph{9048}\\
-\hline
-\hline
-\multicolumn{10}{c}{Uniquely Solved} \\ \hline
-10:1		& 1 & 3 & 2 & 3 & 8 & \emph{16} & 0 & 1 & 2 & 4 & \emph{7}\\
-1:10		& 1 & 3 & 2 & 1 & 5 & \emph{11} & 0 & 1 & 0 & 7 & \emph{8}\\
-1:100	        & 1 & 2 & 2 & 7 & \textbf{9} & \emph{20} & 2 & 0 & 3 & 9 & \emph{14}\\
-1:1000	        & 1 & 2 & 0 & 0 & 3 & \emph{5} & 2 & 0 & 0 & \textbf{12} & \emph{14}\\
-\hline
-Union           & \emph{4} & \emph{10} & \emph{6} & \emph{11} & \emph{25} & \emph{52} & \emph{4} & \emph{2} & \emph{5} & \emph{32} & \emph{43}\\ 
-\end{tabular}
-\end{table}
-
-Vampire ran in default mode (with the \texttt{discount} saturation algorithm) for 10s whilst varying  \texttt{age\_weight\_ratio} and \texttt{age\_weight\_ratio\_shape\_frequency} for both new values of \texttt{age\_weight\_ratio\_shape}.
-Constant AWR values also ran as a baseline.
+Vampire ran in default mode (with the \texttt{discount} saturation algorithm) for 10s whilst varying  \texttt{age\_weight\_ratio} and \texttt{age\_weight\_ratio\_shape\_frequency} for several AWR shapes: constant, converging from 1:1, decaying to 1:1, converging from 1:4 and decaying from 1:4.
 
@@
 The total number of problems solved and those solved uniquely are also reported.
-
-In all cases, for a given start AWR value, varying the AWR over time outperformed keeping it constant.
-As the new strategies always include some reasoning using \verb|1:1| AWR this suggests that all benefit from \emph{some} fair reasoning during proof search and some biased reasoning.
 The best combination of options overall was decaying from an initial age/weight of 1:100 with frequency 1000.
-However, in general, strategies that converged performed better than those that decayed, both in terms of total problems solved and in terms of unique solutions. 
 Longer frequencies tended to do better, suggesting that more time at the intermediate AWR values is preferable. 
-In total, 28 out of a total of 36 strategies contributed some unique solutions. This shows that these new options are complementary in general.
+Unique solutions are distributed well in general, showing that the new options are complementary.
 
@@
 New options were applied on top of the portfolio mode options, using the existing AWR values in the various strategies as the starting point.
+Three shapes are employed: constant (baseline), converging from 1:1 and decaying to 1:1.
 The purpose is to gauge what impact adding such options to a new portfolio mode could have.

### 2019-06-10 papers/ARW19/abstract.tex
ARW 2019 abstract

\documentclass[twocolumn,a4paper,10pt]{article}
\usepackage{tcolorbox}
\newcommand{\vampire}{\textsc{Vampire}}
\usepackage{arw19}
\title{Reinforcement-Learned Input for Saturation Provers}
\institute{University of Manchester, UK}
\author{Michael Rawson \and Giles Reger}
\institute{University of Manchester, UK}
\abstract{
	Many of today's best-performing automatic theorem provers (ATPs) for first-order logic rely on \emph{saturation} algorithms.
	However, to date the most successful work on applying machine-learned guidance to ATPs relies on tableaux methods, as these are friendlier toward machine learning algorithms.
	We describe a reinforcement-learning system which selectively infers new clauses from an input problem.
	The system (in progress) is rewarded if generated clauses are subsequently used by the first-order saturation prover \vampire{} in a proof.
	In this way, the system learns to generate new clauses which are important in the proof search, but might otherwise not be selected for some time in Vampire's proof search algorithm.
	The system is implemented via \(Q\)-learning, with a graph neural net processing the structure of the clause set acting as an approximator.
}
\begin{document}
\maketitle{}
\section{Saturation Provers}
Saturation-based ATPs such as \vampire{}~\cite{vampire} utilise an algorithm which attempts to produce a \emph{saturated} set of first-order clauses from their input: that is, clauses closed under all possible inferences in their calculus (usually ordered resolution and superposition).
If the empty clause is found, then this witnesses refutation and a proof may be given.
Conversely, if a saturated set is produced, then the input is satisfiable and therefore not a theorem.
On hard problems, neither case will occur and the prover will diverge and continue to generate a very large set of clauses until either time or memory runs out.

This approach lends itself to efficiency, but it is inherently unguided and ill-conditioned with respect to the input.
In \vampire{} and other similar ATP systems, a number of options and manually-programmed heuristics are available in order to try and delay divergence or find a proof faster on certain problem classes: these are made more useful by \emph{portfolio modes}~\cite{portfolio} in which many combinations of options are run in sequence in order to ameliorate the brittle performance of the underlying prover.
Such provers have accumulated several pieces of experimentally-validated ``folklore''~\cite{slowly-growing,fewest-children}, notably
\begin{tcolorbox}
	\centering
	\emph{``There is (at least) one clause, which, once found, significantly accelerates finding a proof.''}
\end{tcolorbox}
\noindent
A system which learns to find such a clause early would therefore be very useful.

\section{Guidance for ATP systems}
The general idea of guiding ATP systems with a machine-learned heuristic is hardly novel: experiments with premise selection~\cite{premise-selection} and internal guidance of tableau-based provers~\cite{rlcop} have produced successful results.
However, guiding saturation-based theorem provers has proved more difficult: there is a tradeoff between learning power~\cite{dngps} and throughput~\cite{enigma}: the better the heuristic, the more computationally expensive it is and therefore the greater the impact on prover throughput.

The system described falls somewhere between premise selection and internal guidance: while it can select clauses, it can also perform inferences which may help proof search directly (even acting as an poor man's theorem prover in its own right).
However, it does not affect internal proof search routines and therefore does not impact prover throughput.

\section{Reinforcement Learning Clausal Search}
Reinforcement learning aims to produce agents which learn to perform actions in an environment while maximising a reward function~\cite{reinforcement}.
Theorem proving with clauses can be viewed as performing actions (generating inferences) in an environment (the current clause set), aiming to find a reward (the empty clause).
In principle, this would be a good way to build a learning theorem prover in its own right.
However, this approach has a few problems:
\begin{enumerate}
	\item\label{problem:sparse-reward} The reward function here is very sparse, with only a single reward at the end of proof search.
	\item\label{problem:deep-proofs} Proofs may be very deep, which an agent at the start of its training may not be able to find in reasonable time.
	\item\label{problem:large-space} The space of all possible inferences with all available clauses is very large, particularly in large knowledge bases.
\end{enumerate}
In order to make this approach more friendly to reinforcement learning, a few modifications are introduced.
An existing ATP system is employed after a fixed number of steps on the modified clause set: if the existing ATP can prove the modified problem making use of the generated clauses, a reward is applied.
This helps with issues (\ref{problem:sparse-reward}) and (\ref{problem:deep-proofs}).

To reduce the number of possible inferences (\ref{problem:large-space}), the system is allowed to select which clauses from the input it wishes to use in generating inferences.
We therefore partition the system into two disjoint clause sets:
\begin{description}
	\item\emph{selected}: initially empty, the set the system has selected
	\item\emph{available}: input clauses, and inferences from \emph{selected}
\end{description}

\section{System Description}
The system first uses the \vampire{} ATP to clausify an input problem, obtaining a set of input clauses --- those related to the conjecture are tracked and marked as such to enable the system to be goal-directed if it chooses.
Any other reasonable clausifier could be used here.

Then, the system is allowed to choose one clause at a time from \emph{available} and move it to \emph{selected}: any inferences from the new clause with others in \emph{selected} are then added to \emph{available}.
The agent's action-selection policy may be slightly randomised (e.g. \(\epsilon\)-greedy or Boltzmann selection~\cite{random-action}): in training this gives an exploration effect, in testing this allows for multiple distinct clause sets to be produced: multiple proof attempts can then be run in parallel.

After a fixed number of clause selections, the process is halted and \vampire{} runs in its default mode for a short time on all the original clauses, plus the generated clauses in \emph{selected}.
Clauses are ordered in such a way that \vampire{} prefers those in \emph{selected} before those in \emph{available}.
If clauses in \emph{selected} are used in proof search, a reward signal is propagated, and the agent may learn from its experience.
A mild punishment signal is propagated for clauses that were selected but not used.

The particular reinforcement algorithm used presently is \(Q\)-learning~\cite{q-learning}, with a graph convolutional network~\cite{gcn} applied to the entire clause set's structure as a \(Q\)-function approximator.

\section{Future Work}
As well as completing the implementation and evaluation of the system, there are several possible directions for future work:
\begin{itemize}
	\item Different (neural) function approximators. The existing neural architecture is known to perform well on similar problems, but there is much to explore here.
	\item Other learning approaches: modern approaches such as A3C~\cite{a3c} claim significantly better performance on some domains.
	\item Suitable \vampire{} modes for this environment, and the effect the underlying ATP algorithm has on learned policy.
	\item Modified reward functions: appearing in the proof is not the only way a clause may be useful within \vampire{}. For example, if a clause subsumes a large section of search space, it may still increase proof search performance despite not appearing in the eventual proof. Possible future rewards might include the number of subsumed clauses, and the time taken for the ATP to run.
	\item There are a significant number of hyper-parameters within this system, all of which may be tuned to increase system performance on a given benchmark set.
\end{itemize}
\bibliography{references}
\end{document}

### 2019-06-30 papers/FroCoS19/paper.tex
FroCoS '19 (partial) feedback

@@
 \begin{document}
-\title{A Neural, Parallel Theorem Prover*}
+\title{A Neurally-Guided, Parallel Theorem Prover*}
 \author{
@@
 
-Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
+Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode the syntactic structure of proof state and therefore lose information.
 By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
@@
 A promising direction in this area is recent research on neural methods working with graphs~\cite{graph-cnn,gcn,gcn-relational}, which have already been applied to premise selection~\cite{formula-graph}.
-Graph neural networks tend to include network layers inspired by convolution operators in image-processing networks, combining information from neighbouring nodes (pixels).
-For example, the Graph Convolutional Network~\cite{gcn} includes the following convolution layer, where \(H_i\) is the network output at layer \(i\), \(\bar{A}\) is the self-connected adjacency matrix, \(\bar{D}\) is the diagonal node degree matrix of \(\bar{A}\), \(W_i\) is a trainable weight matrix and \(\sigma\) is a non-linear activation function:
-\[
-	H_{i + 1} = \sigma\left(\bar{D}^{-\frac{1}{2}}\bar{A}\bar{D}^{-\frac{1}{2}}H_iW_i\right)
-\]
+Graph neural networks tend to include network layers inspired by convolution operators in image-processing networks, combining information from neighbouring nodes (pixels)~\cite{gcn}.
 
@@
 Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
-For such a system, we require the following from the calculus:
+For such a system, we desire the following from the calculus:
 \begin{enumerate}
-        \item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of computation and resources.
+        \item \emph{Proof state must be reasonably-sized.} Attempting to evaluate large proof states structurally requires a lot of computation and resources.
 Saturation-based provers can have very large proof states, for example.
@@
 Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
-        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space.%, without e.g.
+        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space, without e.g.
 unifiers.
@@
 where ${\sf score}$ gives the heuristic score, ${\sf visits}$ gives the total number of visits to that node so far, and \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)).
-The sub-goal with the minimal score is then selected: this prioritises subgoals considered possibly satisfiable by the heuristic, as satisfiable subgoals allows large parts of the search space to be pruned.
+The sub-goal with the minimal score is then selected: this prioritises subgoals considered possibly satisfiable by the heuristic, as satisfiable subgoals allow large parts of the search space to be pruned.
 
@@
 \paragraph{Refutation Tableaux}
-In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
+In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and then show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
 A set of inference rules of the form
@@
 \label{section:heuristic}
-A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
+A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatisfiability, based on a set of tagged formulae seen in previous proof search.
 Although the data is collected by running the system itself and might be considered \emph{reinforcement} learning, for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.
@@
 Constants, function symbols, predicate symbols, and bound variables are given their own node.
-Application of functions and predicates to arguments are represented as an ``application node'' with two children: the symbol node and an ``argument list'' node representing the list of arguments.
+Applications of functions and predicates to arguments are represented as an ``application node'' with two children: the symbol node and an ``argument list'' node representing the list of arguments.
 Propositional connectives and equality have the obvious representation, while quantifiers have two children: the variable they bind and their sub-formula.
@@
 To produce an input graph from a formula \(F\), the formula is first parsed into an abstract syntax tree.
-Common sub-trees up to \(\alpha\)-equivalence are merged, then the resulting directed acyclic graph has any named-symbol nodes replaced with an opaque, nameless label such as "predicate" or "variable" --- since distinct symbols remain as distinct nodes under this scheme, no information is lost other than the natural-language semantics of the symbol name.
+Common sub-trees up to \(\alpha\)-equivalence~\cite{barendregt} are merged, then the resulting directed acyclic graph has any named-symbol nodes replaced with an opaque, nameless label such as ``predicate'' or ``variable'' --- since distinct symbols remain as distinct nodes under this scheme, no information is lost other than the natural-language semantics of the symbol name.
 In practice, undirected graphs improved model performance so the graph is made undirected before encoding node labels as one-hot inputs to produce the final input graphs.

### 2019-07-01 papers/FroCoS19/paper.tex
final FroCoS paper

@@
 \begin{document}
-\title{A Neurally-Guided, Parallel Theorem Prover*}
+\title{A Neurally-Guided, Parallel Theorem Prover}
 \author{
-	Michael Rawson*\orcidID{0000-0001-7834-1567} \and
+	Michael Rawson\orcidID{0000-0001-7834-1567} \and
 	Giles Reger
@@
 \begin{enumerate}
-	\item Proof state in such systems may be very large, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
+	\item Proof state in such systems may be of impractical size, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
+In a saturation context, the size of the current proof state may be many times the size of the eventual proof: while neural networks are in principle capable of processing large amounts of data, throughput suffers and scalability is a concern.
 	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
@@
 Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures.
-Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.
+Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.
+Source code available at \url{https://github.com/MichaelRawson/lerna}.} takes an alternative step toward useful neural automatic theorem provers.
 
@@
 
-%A formula is either true \(\top\), false \(\bot\), an equality between terms \(t_1 = t_2\), a predicate applied to terms \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol, a negated formula \(\lnot \phi\), an impliciation \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
-%Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
-%Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).
-
 \paragraph{Automatic Theorem Provers}
@@
 CVC4~\cite{CVC4} and Z3~\cite{Z3}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
-%Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
-%Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.
 
@@
 
-Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode the syntactic structure of proof state and therefore lose information.
+Work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA,proof-mining}, which have the advantage of simplicity and relative efficiency, but do not fully encode the syntactic structure of proof state and therefore lose information.
 By contrast, a neural method which takes into account all information (as utilised in this work) should allow for greater precision in proof guidance systems.
@@
 \paragraph{The MPTP Problem Set}
-For training and evaluation purposes a set of problems exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
+For training and evaluation purposes a set of valid propositions exported from the Mizar Mathematical Library~\cite{mizar} by the MPTP~\cite{MPTP} system are used.
 Urban et al.~\cite{rlCoP} took a subset\footnote{\url{https://github.com/JUrban/deepmath/blob/master/M2k_list}} of the large \mizarlarge{} problem set (containing 32,524 problems) and called it \mizarsmall{} (containing 2004 problems).

### 2019-07-31 outline/outline.tex
added thesis outline document

\documentclass[a4paper]{article}
\usepackage{biblatex}
\bibliography{references}
\title{Thesis Outline}
\author{Michael Rawson}
\begin{document}
\maketitle
\abstract{
\noindent
Thesis outline for second-year progress report. I research the application of recently-developed machine learning techniques to automatic theorem provers in a variety of different ways.
}
\section{Introduction}
\begin{itemize}
	\item Background and history of automated reasoning and machine learning.
	\item First-order logic as a setting: TPTP and Vampire.
	\item Brief survey of the state of ATP research, including current problems and research directions without getting into literature review.
	\item Current capabilities and recent advances in relevant machine learning topics.
	\item Machine learning as applied to theorem proving: domains and success or otherwise.
\end{itemize}

\section{Literature Review}
\label{sec:lit-review}
A literature review, focussing on, but not limited to, the intersection of the two domains.
Previous PhD thesis in this area: James Bridge~\cite{bridge} --- cover advances since then.

\section{Data-Driven Theorem Proving}
\label{sec:data}
Machine learning is generally a statistical inference tool: before applying learning, what conclusions can we draw from data?
Does this data correspond to existing ATP ``folklore'' covered in Section \ref{sec:lit-review}?
\begin{itemize}
	\item Vampire Workshop 2018 talk on statistical analysis of Vampire proofs. Shows a variety of statistical properties about Vampire's proof search capabilities, including that the majority of proofs are relatively ``shallow'', for example.
	\item Unpublished work intended for LPAR 2018 (can we put on arXiv for citations perhaps?) again showing some statistical properties of TSTP proofs generated from various ATP systems. Also discovered some flaws in proof output in mainstream theorem provers, showing that the field has some way to go in the direction of verified output.
	\item CADE 2019 work (paper accepted, publication pending) on reshaping the proof search space by means of variable age/weight ratios. Also experimentally confirms folklore about optimum AWR values.
\end{itemize}
Some relevant ATP folklore/philosophy can be discussed here.
ATP systems tend to find a proof quickly, or not at all.
The task of performing well at TPTP (or other benchmarks) is therefore something of a set-cover problem: reshape the search space in several different ways (by means of \emph{strategies} in Vampire, for instance) until proof search succeeds in finding a proof quickly.

\section{Intelligent High-Level ATP Guidance}
Work published at PAAR 2018 on an approach for improving ATP performance by selecting strategies likely to succeed via a machine-learned heuristic.
This work also describes a method of strategy scheduling which avoids some performance pitfalls, playing into Section \ref{sec:efficiency}.
In a more general context, this section describes techniques for indirectly guiding an existing ATP system.

\section{Representing Logical Data for ML}
\label{sec:representation}
A central problem for integrating ML techniques into ATP systems is representation: logical data such as terms and clauses do not map well into existing ML patterns such as real vectors, sequences, or images.
Work pending publication at FroCoS 2019 describes a theorem-proving system guided by a neural network using a graph-based representation of logical data.
This representation is not novel (used previously in premise selection for higher-order logic) but this work develops it further and applies it to search guidance for first-order logic.

\section{Efficiency Concerns for Guided Provers}
\label{sec:efficiency}
One problem with the neural method described in Section \ref{sec:representation} is latency: from input to output, the neural heuristic takes significant amounts of time, which impacts prover performance sufficiently to negate the improvement brought about by guidance.
Work presented at AITP 2019 describes a prover architecture which ameliorates this issue, in exchange for a set of constraints on the prover calculus.
A na\"ive calculus is used here as a prototype.

\section{Ongoing Work: Neural Clause Inference}
A result from Section \ref{sec:representation} is that clausal forms can be represented in neural networks.
While a saturation-based theorem prover remains unfriendly to neural guidance for performance reasons, a pre-processing step in which a neural network chooses some clauses inferences to perform before feeding the resulting clause set to an existing ATP system achieves ``smart'' reshaping of the search space in order to encourage a short ATP proof, as discussed in Section \ref{sec:data}, without the disadvantages of proof guidance or the manual set-cover approach of ATP strategies.

\section{Future Work: Non-Clausal Connection Methods with Neural Guidance}
Section \ref{sec:efficiency} presents a prover prototype, with some issues introduced by the na\"ive calculus employed.
Future work proceeds down an obvious route by employing an existing-but-similar calculus with better properties, the non-clausal connection calculus.
Implementation is ongoing but theoretically this is promising.

\section{Conclusion}
Argue that ATP systems augmented with learned knowledge have a promising future, and discuss future directions for the area based on experimental results of preceding sections.

\printbibliography
\end{document}

### 2019-09-01 papers/ARW19/poster.tex
Copy ARW '18 poster for ARW '19
ARW '19 poster

--- 2019-02-12 papers/ARW18/poster.tex
@@
 \documentclass[24pt, a1paper, portrait]{tikzposter}
+\usepackage{enumitem}
+\usepackage{graphicx}
 \tikzposterlatexaffectionproofoff
-\usepackage{tikz}
-\usepackage{amsmath}
-\usepackage{enumitem}
-\usetikzlibrary{shapes, arrows}
-\usepackage[sfdefault]{roboto}
-\usepackage[T1]{fontenc}
+
+\newcommand{\vampire}{\textsc{Vampire}}
 
@@
 
-\tikzstyle{decision} = [diamond, fill=highlight, text=white, align=center, node distance=15cm]
-\tikzstyle{block} = [rectangle, fill=highlight, text=white, rounded corners, align=center, node distance=15cm, minimum height=4cm]
-\tikzstyle{cloud} = [ellipse, fill=highlight, text=white, align=center, node distance=15cm, minimum height=4cm]
-\tikzstyle{arrow} = [->, draw, draw=highlight, line width=0.3cm]
-
-\renewcommand\labelitemi{\large{\bullet}}
-%\renewcommand\labelitemi{$\vcenter{\hbox{\Large$\bullet$}}$ }
-
-\title{A Proof Calculus for Learned Heuristics}
+\title{Reinforcement-Learned Input for Saturation}
 \author{Michael Rawson, Giles Reger}
-\date{April 2018}
+\date{September 2019}
 \institute{University of Manchester, UK}
@@
 {
-	Machine learning techniques may prove capable of efficiently guiding search for proofs of human theorems.
-	Many modern automated theorem provers (E, {\sc Spass}, iProver, Vampire, \ldots) make use of proof calculi such as saturation or instance generation which, while efficient, are unsuitable for current machine-learning technology.
-	We propose the use of a type-theoretic calculus for a future theorem prover which attempts to leverage machine-learning techniques.
+	Many of today's best-performing automatic theorem provers (ATPs) for first-order logic rely on \emph{saturation} algorithms.
+	Saturation algorithms perform all possible inferences within an active set, which yields a complete proof search algorithm.
+	However, this produces a large number of clauses, which hinders attempts to guide the prover using neural networks.
+	We describe a system which infers new clauses in an incomplete fashion, guided by a neural network.
+	The system (in progress) is rewarded if generated clauses are subsequently used by the saturation-based ATP \vampire{} in a proof.
+	In this way, the system learns to selectively infer new clauses which are important in proof search, but might not otherwise be found for some time in \vampire{}'s algorithm.
+
 }
@@
 \column{0.5}
-\block{The Prover Cycle}
+\block{System Description}
 {
-	\begin{tikzfigure}[A generalisation of information flow in a modern prover. Note the impact that a strong (learning) heuristic may have.]
-		\begin{tikzpicture}
-			\node [decision] (choose) {choose next\\step};
-			\node [cloud, right of=choose] (heuristic) {heuristic};
-			\node [block, below of=choose] (perform) {execute step};
-			\node [block, right of=perform] (process) {process\\consequences};
-
-			\path [arrow] (choose) edge node [midway, right, align=center] {proof\\step} (perform);
-			\path [arrow] (perform) edge node [midway, above] {new state} (process);
-			\path [arrow] (process) edge node [sloped, above] {stable state} (choose);
-			\path [arrow] (process) edge node [midway, right, align=center] {prover\\data} (heuristic);
-			\path [arrow] (heuristic) edge node [midway, above] {guidance} (choose);
-		\end{tikzpicture}
-	\end{tikzfigure}
+	The system must use an existing ATP system, in this case \vampire{}.
+	On a given first-order problem, it acts as follows:
+	\begin{enumerate}[leftmargin=1in]
+		\item Clausify an input problem to obtain a clause set.
+		\item Loop:
+			\begin{enumerate}[leftmargin=1in]
+				\item Neurally select an inference between two clauses.
+				\item Perform the inference (using the ATP)
+				\item Add any generated clauses to the clause set.
+			\end{enumerate}
+		\item Repeat until a time or inference threshold is met.
+		\item Invoke the ATP on the new clause set and run as usual.
+	\end{enumerate}
+	\vspace{.5in}
+	This retains the advantages of advanced clausification and theorem-proving technology (such as completeness and advanced redundancy eliminations), but with the addition of a neural pre-processing step.
 }
 
-\block{Existing Calculi: Problems for ML}
+\block{Reinforcement and Reward}
 {
-	Existing calculi such as resolution do not adapt well to current machine-learning methods, for several reasons:
+	It is unclear at present what reward function should be chosen for the network, but the reward can be significantly more expressive than any manual heuristic can be.
 
-	\begin{itemize}[leftmargin=1.5cm]
-		\item Artificial pre-processing (CNF, Skolemisation) reduces human structure in proof state
-		\item Large, recursive input space: proof state
-		\item Huge number of outputs to predict: possible next steps from an unbounded set of clauses
-		\item A reliance on high performance for finding proofs
+	Examples of inferences to reward include:
+	\begin{itemize}[leftmargin=1in]
+		\item Clauses included in the eventual proof (current).
+		\item Only clauses used more than once in the proof.
+		\item Clauses which subsume other clauses.
+		\item Ground clauses (which can be used for splitting).
+		\item Any weighted combination of the above.
 	\end{itemize}
-	A better calculus would address these issues while retaining the reasonably-short proofs and high performance of existing calculi.
+	\vspace{.5in}
+	Since the clause space can be very large even on small problems, supervised learning is not very practical here.
+	Instead, reinforcement learning techniques may be leveraged, providing training data by self-play.
+	The current prototype simply generates clause sets by self-play and then trains the network to identify clauses used in the proof.
 }
@@
 \column{0.5}
-\block{Type Theory for First-Order Logic}
+\block{Graphs for Clause Processing}
 {
-	Via the Curry-Howard correspondence (``propositions-as-types''), we can encode first-order logic in a system that resembles LF.
-	For example, the theorem
-	\[
-		(\forall x, P(x)) \implies P(c) \wedge P(d)
-	\]
-	might be shown with the \(\lambda\)-term
-	\[
-		\lambda p : (\forall x, P(x)).\ \textrm{and}(p(c), p(d))
-	\]
-	Proof search is now term synthesis.
+	{
+	\centering
+	\includegraphics[width=10in]{clause-graph}
+	}
+	The complex directed graph shown is a compact, lossless representation of TPTP problem \texttt{PUZ001+1}, or ``who killed Aunt Agatha?'', with perfect term sharing.
+	Graph neural networks can process such graphs by sharing information between adjacent nodes.
+	This is a convenient end-to-end method for processing clauses in a neural context.
 }
 
-\block{Type Theory for Proof Search}
+\block{Future Work}
 {
-	By allowing ``holes'' in the system, search for an appropriate term can proceed iteratively: consider \emph{modus ponens}, translated into a type:
-	\[
-		P \to (P \to Q) \to Q
-	\]
+	Much is still left to do:
+	\begin{itemize}[leftmargin=1in]
+		\item Finish training and evaluate a prototype.
+		\item Examine the effect on proof search in \vampire{}.
+		\item Determine good strategies for the modified clause sets generated.
+		\item Experiment with other reward functions and reinforcement learning techniques, such as \(Q\)-learning.
+	\end{itemize}
+}
+\end{columns}
 
-	This can obtained by repeatedly refining holes, \(?\), such that the term at any point still satisfies the target type.
-	Note the convergence of the inferred type onto the target type as the proof proceeds, and the limited amount of choice at each step.
-
-\begin{center}
-	\begin{tabular}{l | l}
-		\\
-		Refined Term & Inferred Type \\
-		\(?\) & \(?\) \\
-		\(\lambda x:(P).?\) & \(P \to\ ?\)\\
-		\(\lambda x:(P).\lambda f:(P \to Q).?\) & \(P \to (P \to Q) \to\ ?\)\\
-		\(\lambda x:(P).\lambda f:(P \to Q). \left(?\ ?\right)\) & \(P \to (P \to Q) \to\ ?\)\\
-		\(\lambda x:(P).\lambda f:(P \to Q). \left(f\ ?\right)\) & \(P \to (P \to Q) \to\ Q\)\\
-		\(\lambda x:(P).\lambda f:(P \to Q). \left(f\ x\right)\) & \(P \to (P \to Q) \to\ Q\)\\
-		\\
-	\end{tabular}
-\end{center}
-	The input state is now more structured, and much smaller.
-	Additionally, to fill any given hole, only a subset of all available variables will be appropriate for the target type --- this reduces the output space dramatically.
-	Deductions and machine-learnt heuristics may also occur in parallel, as proof steps are not inter-dependent.
-}
-
-\end{columns}
 \end{document}

### 2019-09-11 coursework/progress-report-2/report.tex
cleanup, year 2 progress report

\documentclass[a4paper]{article}
\usepackage{biblatex}
\usepackage{graphicx}
\bibliography{references}
\title{Progress Report, Year 2}
\author{Michael Rawson}
\begin{document}
\maketitle
\section{Introduction}
I research the application of recently-developed machine learning techniques to automatic theorem provers in a variety of different ways.
This documents progress made in this area for my PhD program, and plans the content and delivery of my thesis.

\section{Progress}
Several novel pieces of research have been conducted in the time since the Year 1 Progress Report, and some is ongoing.
These include published work into representing logical data in a neural context (Section \ref{sec:representation}), prover architectures for supporting such a representation (Section \ref{sec:efficiency}), and further work on statistical analysis (Section \ref{sec:data}).
Work with neurally-guided clausal inference (Section \ref{sec:clauses}) and distance-based guidance metrics (Section \ref{sec:distance}) is ongoing and shows promise.

\section{Thesis Outline}
The proposed outline of the thesis follows.

\subsection{Introduction}
I introduce common topics and background used throughout the thesis, but which are not directly relevant to the thesis topic.
These will include:
\begin{itemize}
	\item Background and history of automated reasoning and machine learning.
	\item First-order logic as a setting: Vampire and TPTP
	\item Brief survey of the state of ATP research, including current problems and research directions without getting into literature review.
	\item Current capabilities and recent advances in relevant machine learning topics.
\end{itemize}

\subsection{Literature Review}
\label{sec:lit-review}
A literature review, focussing on, but not limited to, the intersubsection of the two domains.
In particular, cover advances since the previous PhD thesis in this area~\cite{bridge}.

\subsection{Data-Driven Theorem Proving}
\label{sec:data}
Machine learning is a statistical inference tool: before applying learning, what conclusions can we draw from data?
Does this data correspond to existing ATP ``folklore'' covered in Section \ref{sec:lit-review}?
I present work based on the following:
\begin{itemize}
	\item Vampire Workshop 2018 talk~\cite{vampire2018} on statistical analysis of Vampire proofs. Shows a variety of statistical properties about Vampire's proof search capabilities, including that the majority of proofs are relatively ``shallow'', for example.
	\item Unpublished work showing some statistical properties of TSTP proofs generated from various ATP systems. This work also discovered some flaws in proof output in mainstream theorem provers, showing that the field has some way to go in the direction of verified output.
	\item CADE 2019~\cite{cade2019} work on reshaping the proof search space by means of variable age/weight ratios. The work also experimentally confirms folklore about optimum AWR values.
\end{itemize}

\noindent
Some relevant ATP folklore/philosophy is also be discussed here.
ATP systems tend to find a proof quickly, or not at all.
Therefore, the task of performing well at TPTP (or other benchmarks) is therefore something of a set-cover problem: reshape the search space in several different ways (by means of \emph{strategies} in Vampire, for instance) until proof search succeeds in finding a proof quickly.

\subsection{Intelligent High-Level ATP Guidance}
Present work published at PAAR 2018~\cite{paar2018} on an approach for improving ATP performance by selecting strategies likely to succeed via a machine-learned heuristic at runtime.
This work also describes a method of strategy scheduling which avoids some performance pitfalls, playing into Section \ref{sec:efficiency}.
In a more general context, this subsection describes techniques for indirectly guiding an existing ATP system.

\subsection{Representing Logical Data for ML}
\label{sec:representation}
A central problem for integrating ML techniques into ATP systems is representation: logical data such as terms and clauses do not map well into existing ML patterns such as real vectors, sequences, or images.
Work published at FroCoS 2019~\cite{frocos2019} describes a theorem-proving system guided by a neural network using a graph-based representation of logical data.
This representation is not novel (used previously in premise selection for higher-order logic) but this work develops it further and applies it to search guidance for first-order logic.

\subsection{Efficiency Concerns for Guided Provers}
\label{sec:efficiency}
One problem with the neural method described in Section \ref{sec:representation} is latency: from input to output, the neural heuristic takes significant amounts of time, which impacts prover performance sufficiently to negate the improvement brought about by guidance.
Work presented at AITP 2019~\cite{aitp2019} describes a prover architecture which ameliorates this issue, in exchange for a set of constraints on the prover calculus.
A na\"ive calculus is used here as a prototype.

\subsection{Neural Clause Inference (ongoing)}
\label{sec:clauses}
A result from Section \ref{sec:representation} is that clausal forms can be represented in neural networks.
While a saturation-based theorem prover remains unfriendly to neural guidance for performance reasons, a pre-processing step in which a neural network chooses some clauses inferences to perform before feeding the resulting clause set to an existing ATP system achieves ``smart'' reshaping of the search space in order to encourage a short ATP proof, as discussed in Section \ref{sec:data}, without the disadvantages of proof guidance or the manual set-cover approach of ATP strategies.

\subsection{Connection Methods with Neural Guidance (ongoing)}
\label{sec:distance}
Section \ref{sec:efficiency} presents a prover prototype, with some issues introduced by the na\"ive calculus employed.
An obvious improvement uses an existing calculus to improve prover performance, the \emph{connection calculus}.
Another problem with Section \ref{sec:efficiency} is the lack of real-valued distance metrics, rather than binary satisfiability metrics.
The issue is addressed in this work, experimenting with regression rather than classification for guiding theorem proving.

\subsection{Conclusion}
ATP systems augmented with learned knowledge have a promising future.
I discuss future directions for the area based on theoretical, practical and experimental results of preceding subsections.

\section{Planning}
Given the relatively concise, self-contained nature of the proposed chapters, I estimate conservatively that one chapter may be written per month, starting immediately after this progress checkpoint.
This puts me on-schedule to complete and submit the thesis within the year, with some buffer time for unexpected incidents.
Figure \ref{fig:deps} shows dependencies between chapters.
\begin{figure}
	\centering
	\includegraphics[width=0.5\linewidth]{dependencies}
	\caption{\(X \rightarrow Y\) means ``Chapter X must be written before Chapter Y''.}
	\label{fig:deps}
\end{figure}

\printbibliography
\end{document}

### 2019-09-11 outline/outline.tex
cleanup, year 2 progress report

@@
 \section{Introduction}
+I introduce common topics and background used throughout the thesis, but which are not directly relevant to the thesis topic.
+These will include:
 \begin{itemize}
 	\item Background and history of automated reasoning and machine learning.
-	\item First-order logic as a setting: TPTP and Vampire.
+	\item First-order logic as a setting: Vampire and TPTP
 	\item Brief survey of the state of ATP research, including current problems and research directions without getting into literature review.
 	\item Current capabilities and recent advances in relevant machine learning topics.
-	\item Machine learning as applied to theorem proving: domains and success or otherwise.
 \end{itemize}
@@
 A literature review, focussing on, but not limited to, the intersection of the two domains.
-Previous PhD thesis in this area: James Bridge~\cite{bridge} --- cover advances since then.
+In particular, cover advances since the previous PhD thesis in this area~\cite{bridge}.
 
@@
 \label{sec:data}
-Machine learning is generally a statistical inference tool: before applying learning, what conclusions can we draw from data?
+Machine learning is a statistical inference tool: before applying learning, what conclusions can we draw from data?
 Does this data correspond to existing ATP ``folklore'' covered in Section \ref{sec:lit-review}?
+I present work based on the following:
 \begin{itemize}
-	\item Vampire Workshop 2018 talk on statistical analysis of Vampire proofs.
+	\item Vampire Workshop 2018 talk~\cite{vampire2018} on statistical analysis of Vampire proofs.
 Shows a variety of statistical properties about Vampire's proof search capabilities, including that the majority of proofs are relatively ``shallow'', for example.
-	\item Unpublished work intended for LPAR 2018 (can we put on arXiv for citations perhaps?) again showing some statistical properties of TSTP proofs generated from various ATP systems.
-Also discovered some flaws in proof output in mainstream theorem provers, showing that the field has some way to go in the direction of verified output.
-	\item CADE 2019 work (paper accepted, publication pending) on reshaping the proof search space by means of variable age/weight ratios. Also experimentally confirms folklore about optimum AWR values.
+	\item Unpublished work showing some statistical properties of TSTP proofs generated from various ATP systems.
+This work also discovered some flaws in proof output in mainstream theorem provers, showing that the field has some way to go in the direction of verified output.
+	\item CADE 2019~\cite{cade2019} work on reshaping the proof search space by means of variable age/weight ratios. The work also experimentally confirms folklore about optimum AWR values.
 \end{itemize}
-Some relevant ATP folklore/philosophy can be discussed here.
+
+\noindent
+Some relevant ATP folklore/philosophy is also be discussed here.
 ATP systems tend to find a proof quickly, or not at all.
-The task of performing well at TPTP (or other benchmarks) is therefore something of a set-cover problem: reshape the search space in several different ways (by means of \emph{strategies} in Vampire, for instance) until proof search succeeds in finding a proof quickly.
+Therefore, the task of performing well at TPTP (or other benchmarks) is therefore something of a set-cover problem: reshape the search space in several different ways (by means of \emph{strategies} in Vampire, for instance) until proof search succeeds in finding a proof quickly.
 
 \section{Intelligent High-Level ATP Guidance}
-Work published at PAAR 2018 on an approach for improving ATP performance by selecting strategies likely to succeed via a machine-learned heuristic.
+Present work published at PAAR 2018~\cite{paar2018} on an approach for improving ATP performance by selecting strategies likely to succeed via a machine-learned heuristic at runtime.
 This work also describes a method of strategy scheduling which avoids some performance pitfalls, playing into Section \ref{sec:efficiency}.
@@
 A central problem for integrating ML techniques into ATP systems is representation: logical data such as terms and clauses do not map well into existing ML patterns such as real vectors, sequences, or images.
-Work pending publication at FroCoS 2019 describes a theorem-proving system guided by a neural network using a graph-based representation of logical data.
+Work published at FroCoS 2019~\cite{frocos2019} describes a theorem-proving system guided by a neural network using a graph-based representation of logical data.
 This representation is not novel (used previously in premise selection for higher-order logic) but this work develops it further and applies it to search guidance for first-order logic.
@@
 One problem with the neural method described in Section \ref{sec:representation} is latency: from input to output, the neural heuristic takes significant amounts of time, which impacts prover performance sufficiently to negate the improvement brought about by guidance.
-Work presented at AITP 2019 describes a prover architecture which ameliorates this issue, in exchange for a set of constraints on the prover calculus.
+Work presented at AITP 2019~\cite{aitp2019} describes a prover architecture which ameliorates this issue, in exchange for a set of constraints on the prover calculus.
 A na\"ive calculus is used here as a prototype.
 
-\section{Ongoing Work: Neural Clause Inference}
+\section{Neural Clause Inference (ongoing)}
 A result from Section \ref{sec:representation} is that clausal forms can be represented in neural networks.
@@
 
-\section{Future Work: Non-Clausal Connection Methods with Neural Guidance}
+\section{Connection Methods with Neural Guidance (ongoing)}
 Section \ref{sec:efficiency} presents a prover prototype, with some issues introduced by the na\"ive calculus employed.
-Future work proceeds down an obvious route by employing an existing-but-similar calculus with better properties, the non-clausal connection calculus.
-Implementation is ongoing but theoretically this is promising.
+An obvious improvement uses an existing calculus to improve prover performance, the \emph{connection calculus}.
+Another problem with Section \ref{sec:efficiency} is the lack of real-valued distance metrics, rather than boolean satisfiability metrics.
+The issue is addressed in this work, experimenting with regression rather than classification for guiding theorem proving.
 
 \section{Conclusion}
-Argue that ATP systems augmented with learned knowledge have a promising future, and discuss future directions for the area based on experimental results of preceding sections.
+ATP systems augmented with learned knowledge have a promising future.
+I discuss future directions for the area based on theoretical, practical and experimental results of preceding sections.
 

### 2019-09-12 coursework/progress-report-2/report.tex
feedback from Giles, sent to Konstantin

@@
 \section{Introduction}
-I research the application of recently-developed machine learning techniques to automatic theorem provers in a variety of different ways.
+I research the application of recently-developed machine learning (ML) techniques to automatic theorem provers (ATPs) in a variety of different ways.
 This documents progress made in this area for my PhD program, and plans the content and delivery of my thesis.
@@
 \section{Progress}
-Several novel pieces of research have been conducted in the time since the Year 1 Progress Report, and some is ongoing.
-These include published work into representing logical data in a neural context (Section \ref{sec:representation}), prover architectures for supporting such a representation (Section \ref{sec:efficiency}), and further work on statistical analysis (Section \ref{sec:data}).
-Work with neurally-guided clausal inference (Section \ref{sec:clauses}) and distance-based guidance metrics (Section \ref{sec:distance}) is ongoing and shows promise.
+Several novel pieces of research have been conducted in the time since the preceding Progress Report, and some is ongoing.
+These include published work into representing logical data in a neural context (\S\ref{sec:representation}), prover architectures for supporting such a representation (\S\ref{sec:efficiency}), and statistical analysis of existing theorem provers (\S\ref{sec:background}).
+Work with distance-based guidance metrics (\S\ref{sec:distance}) and neurally-guided clausal inference (\S\ref{sec:clauses}) is ongoing and shows promise.
 
@@
 
-\subsection{Introduction}
-I introduce common topics and background used throughout the thesis, but which are not directly relevant to the thesis topic.
-These will include:
+\subsection{Introduction and Motivation}
+\label{sec:introduction}
+Automatic theorem provers stand to gain from the application of machine-learning techniques.
+The ideal is an ``intelligent'' ATP system: one that combines the expert intuition of the human mathematician with the raw speed and tenacity of automatic proof search.
+Such a system would be able to learn generalising experience from easier proofs found in a target domain (such as mathematics or software verification) and apply this knowledge to find proofs of more difficult conjectures, specialising the high performance of the unguided ATP to the target domain.
+In this chapter I introduce and motivate this ideal, and provide a critique of the \emph{status quo} in ATP systems.
+
+\subsection{Background and Context}
+\label{sec:background}
+Firstly, I introduce necessary historical context not directly related to the thesis topic, but assumed throughout.
+This includes:
 \begin{itemize}
-	\item Background and history of automated reasoning and machine learning.
-	\item First-order logic as a setting: Vampire and TPTP
-	\item Brief survey of the state of ATP research, including current problems and research directions without getting into literature review.
-	\item Current capabilities and recent advances in relevant machine learning topics.
+	\item A short history of automated reasoning and machine learning.
+	\item First-order logic as a setting: syntax, semantics, capabilities.
+	\item Vampire as a research vehicle.
+	\item TPTP, Mizar as benchmarks.
+	\item Brief survey of the state of ATP research, including current problems and research directions
+	\item Current capabilities and recent advances in relevant machine learning topics, in particular neural networks.
 \end{itemize}
+I then move onto a more specialised literature review focusing on, but not limited to, the intersection of machine learning and theorem proving as AI topics.
+In particular, I hope to cover advances since the previous PhD thesis in this area~\cite{bridge}.
 
-\subsection{Literature Review}
-\label{sec:lit-review}
-A literature review, focussing on, but not limited to, the intersubsection of the two domains.
-In particular, cover advances since the previous PhD thesis in this area~\cite{bridge}.
-
-\subsection{Data-Driven Theorem Proving}
-\label{sec:data}
-Machine learning is a statistical inference tool: before applying learning, what conclusions can we draw from data?
-Does this data correspond to existing ATP ``folklore'' covered in Section \ref{sec:lit-review}?
-I present work based on the following:
+One unique aspect of ATP research is ``folklore'': acquired knowledge that ATP implementors have discovered during their efforts, which is not generally empirically validated.
+I have expended some effort to validate some of these claims, and to identify some of my own --- since learning is a statistical inference tool, what statistical conclusions can we draw from data before applying a learning algorithm?
+I present the following work:
 \begin{itemize}
@@
 \end{itemize}
+Finally, I discuss a trade-off between lower-quality, indirect, computationally-inexpensive ``fast'' learned guidance and higher-quality, direct, expensive ``slow'' guidance.
+This trade-off forms the backbone of the thesis.
 
-\noindent
-Some relevant ATP folklore/philosophy is also be discussed here.
+\subsection{Indirect ATP Guidance}
 ATP systems tend to find a proof quickly, or not at all.
-Therefore, the task of performing well at TPTP (or other benchmarks) is therefore something of a set-cover problem: reshape the search space in several different ways (by means of \emph{strategies} in Vampire, for instance) until proof search succeeds in finding a proof quickly.
+Therefore, the task of performing well at any finite benchmark is therefore something of a set-cover problem: reshape the search space in several different ways (by means of \emph{strategies} in Vampire, for instance) until proof search succeeds in finding a proof quickly.
+I present work published at PAAR 2018~\cite{paar2018} on an approach for improving ATP performance by selecting strategies likely to succeed via a machine-learned heuristic at run-time.
+This work also describes a method of strategy scheduling which avoids some performance pitfalls.
+In a more general context, this subsection describes techniques for indirectly guiding an existing ATP system by monitoring several strategies: this is ``fast'' guidance.
 
-\subsection{Intelligent High-Level ATP Guidance}
-Present work published at PAAR 2018~\cite{paar2018} on an approach for improving ATP performance by selecting strategies likely to succeed via a machine-learned heuristic at runtime.
-This work also describes a method of strategy scheduling which avoids some performance pitfalls, playing into Section \ref{sec:efficiency}.
-In a more general context, this subsection describes techniques for indirectly guiding an existing ATP system.
+Limitations of this approach are discussed: can this work ever hope to do better than have an educated guess at the behaviour of the underlying algorithms?
+If a system is to actually learn from \emph{problems}, rather than system behaviour on said problems, one has to solve a problem of \emph{representation}.
 
-\subsection{Representing Logical Data for ML}
+\subsection{Representing Logical Data in Neural Networks}
 \label{sec:representation}
@@
 Work published at FroCoS 2019~\cite{frocos2019} describes a theorem-proving system guided by a neural network using a graph-based representation of logical data.
-This representation is not novel (used previously in premise selection for higher-order logic) but this work develops it further and applies it to search guidance for first-order logic.
+This representation is not novel (used previously in premise selection for higher-order logic) but this work develops it further and applies it to search guidance for first-order logic: the first ATP system purpose-built for ``slow'' guidance.
+I compare several learning methods for representing logical data, and their capacity for representation.
 
-\subsection{Efficiency Concerns for Guided Provers}
+\subsection{Efficiency Concerns for Neurally-Guided Provers}
 \label{sec:efficiency}
-One problem with the neural method described in Section \ref{sec:representation} is latency: from input to output, the neural heuristic takes significant amounts of time, which impacts prover performance sufficiently to negate the improvement brought about by guidance.
+One problem with the neural method described in \S\ref{sec:representation} is latency: from input to output, the neural heuristic takes significant amounts of time, which impacts prover performance sufficiently to negate the improvement brought about by guidance.
 Work presented at AITP 2019~\cite{aitp2019} describes a prover architecture which ameliorates this issue, in exchange for a set of constraints on the prover calculus.
-A na\"ive calculus is used here as a prototype.
+A na\"ive calculus is used here as a prototype, but this severely limits the performance of the system.
+
+\subsection{Connection Methods with Neural Guidance (ongoing)}
+\label{sec:distance}
+\S\ref{sec:efficiency} presents a prover prototype, with some issues introduced by the na\"ive calculus employed.
+An obvious improvement uses an existing calculus to improve prover performance, the \emph{connection calculus}.
+Another problem with \S\ref{sec:efficiency} is the lack of real-valued distance metrics, rather than binary satisfiability metrics.
+The issue is addressed in this work, experimenting with regression rather than classification for guiding theorem proving.
 
@@
 \label{sec:clauses}
-A result from Section \ref{sec:representation} is that clausal forms can be represented in neural networks.
-While a saturation-based theorem prover remains unfriendly to neural guidance for performance reasons, a pre-processing step in which a neural network chooses some clauses inferences to perform before feeding the resulting clause set to an existing ATP system achieves ``smart'' reshaping of the search space in order to encourage a short ATP proof, as discussed in Section \ref{sec:data}, without the disadvantages of proof guidance or the manual set-cover approach of ATP strategies.
-
-\subsection{Connection Methods with Neural Guidance (ongoing)}
-\label{sec:distance}
-Section \ref{sec:efficiency} presents a prover prototype, with some issues introduced by the na\"ive calculus employed.
-An obvious improvement uses an existing calculus to improve prover performance, the \emph{connection calculus}.
-Another problem with Section \ref{sec:efficiency} is the lack of real-valued distance metrics, rather than binary satisfiability metrics.
-The issue is addressed in this work, experimenting with regression rather than classification for guiding theorem proving.
+A result from \S\ref{sec:representation} is that clause-normal forms can be represented in neural networks.
+While a state-of-the-art saturation-based ATP remains unfriendly to neural guidance for performance reasons, a pre-processing step in which a neural network chooses some clausal inferences to perform before feeding the resulting modified clause set to an existing ATP system achieves ``smart'' reshaping of the search space in order to encourage a short ATP proof, as discussed in \S\ref{sec:background}, without the disadvantages of proof guidance or the manual set-cover approach of strategy scheduling.
+This represents a mixture of approaches: a limited amount of ``slow'', guided reasoning followed by unguided, brute-force reasoning.
 
@@
 ATP systems augmented with learned knowledge have a promising future.
-I discuss future directions for the area based on theoretical, practical and experimental results of preceding subsections.
+I discuss future directions for the area based on theoretical, practical and experimental results of preceding chapters.
 
 \section{Planning}
-Given the relatively concise, self-contained nature of the proposed chapters, I estimate conservatively that one chapter may be written per month, starting immediately after this progress checkpoint.
-This puts me on-schedule to complete and submit the thesis within the year, with some buffer time for unexpected incidents.
+Given the relatively concise, self-contained nature of the proposed chapters, I estimate that one chapter may be written per month, starting immediately after this progress checkpoint.
+This puts me on-schedule to complete and submit the thesis within the year, with some buffer time for the usual unexpected incidents.
 Figure \ref{fig:deps} shows dependencies between chapters.
@@
 	\centering
-	\includegraphics[width=0.5\linewidth]{dependencies}
+	\includegraphics[width=0.7\linewidth]{dependencies}
 	\caption{\(X \rightarrow Y\) means ``Chapter X must be written before Chapter Y''.}
